# Introduction
Next tasks:
Give a bit of time each day for next few weeks, then full tie when chapter 5 and 6 finished
Continue to go through add the next papers
Then go throuhg each section, do a lit review, and read what i need for each, add the text and references and kkep going 
Should then be easy to tie this together 
Herbst Thesis 

2 days reading each 2 days to write 
= 4 days 

= 16 days 

~3 weeks 



```{r loadLibraries, echo = FALSE}
library(ggplot2)
```

(theoretical background)
## CLL Disease biology 

### Disease Characteristcs
CD5, CD19, CD23+, lowlevelsofCD20 

Reading: 

From Kipps 2017: 
Incidence (get for 2021)
Siegel R, et al. Cancer treatment and survivorship statistics, 2012. CA Cancer J. Clin. 2012; 62:220–241. [PubMed: 22700443]


\@ref(fig:CLLmorphology) Taken from Kipps et al 2017: "Wright–Giemsa-stained blood smears showing the typical chronic lymphocytic leukaemia (CLL) B lymphocyte (part a). _Figure from @Kipps2017._"

```{r CLLmorphology, fig.cap='(ref:CLLmorphology)', echo = FALSE}

knitr::include_graphics("figures/CLL_morphology.eps")

```
_Figure taken from Kipps et al 2017_ 

\@ref(fig:proliferationCentres) Taken from Kipps et al 2017: "a | Tissue sections of a lymph node stained with haemotoxylin and eosin showing numerous pale-staining pseudofollicles, which are circled (original magnification ×20)." _Figure from @Kipps2017_

```{r proliferationCentres, fig.cap='(ref:proliferationCentres)', echo = FALSE}

knitr::include_graphics("figures/proliferation_centres.eps")

```
_Figure taken from Kipps et al 2017_ 

[@Kipps2017] 10.1038/nrdp.2016.96.Chronic
"Chronic lymphocytic leukaemia (CLL) is a malignancy of CD5+ B cells that is characterized by the accumulation of small, mature-appearing lymphocytes in the blood, marrow and lymphoid tissues. Signalling via surface immunoglobulin, which constitutes the major part of the B cell receptor, and several genetic alterations play a part in CLL pathogenesis, in addition to interactions between CLL cells and other cell types, such as stromal cells, T cells and nurse-like cells in the lymph nodes. The clinical progression of CLL is heterogeneous and ranges from patients whorequire treatment soon after diagnosis to others who do not require therapy for many years, if at all. Several factors, including the immunoglobulin heavy-chain variable region gene (IGHV) mutational status, genomic changes, patient age and the presence of comorbidities, should be considered when defining the optimal management strategies, which include chemotherapy, chemoimmunotherapy and/or drugs targeting B cell receptor signalling or inhibitors of apoptosis, such as BCL-2. Research on the biology of CLL has profoundly enhanced our ability to identify patients who are at higher risk for disease progression and our capacity to treat patients with drugs that selectively target distinctive phenotypic or physiological features of CLL. "

Get for 2021:
CLL is estimated to account for ~19,000 of all newly detected cancers in the United States in 2016 (REF. 11). The average incidence of CLL varies between individuals in different geographical regions and ranges from <0.01% of individuals in eastern Asia to ~0.06% of individuals in Europe and the United States. The risk of developing CLL is about two-times higher for men than for women and increases with age; the median age at diagnosis ranges from 70 to 72 years11–14.
The

### IGHV status and the role of BCR signalling
Reading: 
From Kipps 2017: 
IGHV status:
Hamblin TJ, Davis Z, Gardiner A, Oscier DG, Stevenson FK. Unmutated Ig V(H) genes are associated with a more aggressive form of chronic lymphocytic leukemia. Blood. 1999; 94:1848– 1854. [PubMed: 10477713]

Damle RN, et al. Ig V gene mutation status and CD38 expression as novel prognostic indicators in chronic lymphocytic leukemia. Blood. 1999; 94:1840–1847. [PubMed: 10477712] References 1 and 2 are landmark papers that describe two main subsets of patients with different disease progression tendencies based on IGHV mutation status of the immunoglobulins that are expressed by CLL cells.

Restricted repetoire: 
Kipps TJ, et al. Developmentally restricted immunoglobulin heavy chain variable region gene expressed at high frequency in chronic lymphocytic leukemia. Proc. Natl Acad. Sci. USA. 1989; 86:5913–5917. [PubMed: 2503826] This paper describes the discovery that the immunoglobulin repertoire of CLL cells may be highly restricted, suggesting that the antibodies expressed by CLL cells are most likely selected based on their capacity to bind to some common self-antigens.

Fais F, et al. Chronic lymphocytic leukemia B cells express restricted sets of mutated and unmutated antigen receptors. J. Clin. Invest. 1998; 102:1515–1525. [PubMed: 9788964]

Widhopf G, F 2nd, et al. Chronic lymphocytic leukemia B cells of more than 1% of patients express virtually identical immunoglobulins. Blood. 2004; 104:2499–2504. [PubMed: 15217828]


\@ref(fig:BCRsignalling) Adapted from: Kipps et al 2017: "B cell receptor (BCR) signalling is initiated by SRC-family kinase-dependent phosphorylation (mainly LYN) of CD79A and CD79B that creates a docking site for the binding and activation of spleen tyrosine kinase (SYK). SYK then triggers the formation of a multi-component ‘signalosome’, comprising Bruton tyrosine kinase (BTK), AKT, phosphoinositide 3-kinase (PI3K), phospholipase Cγ2 (PLCγ2) and B cell-linker protein (BLNK), among others. CD19 is a co-receptor for BCR and is important for PI3K activation, which recruits and activates PLCγ2, BTK and AKT. PLCγ2 generates diacylglycerol (DAG) and inositol-1,4,5-trisphosphate (Ins(1,4,5)P3), which triggers Ca2+ release from the endoplasmic reticulum, leading to the activation of the MEK–extracellular signal-regulated kinase (ERK) and nuclear factor-κB (NF-κB) signalling pathways. Other effects of BCR signalling include activation of mechanistic target of rapamycin complex 1 (mTORC1) and of Rho-family GTPases, RAC1 and RHOA, which can affect the cytoskeleton. Inhibitors of SYK, PI3K and BTK are shown. Note that this figure describes the main molecules and interactions that are involved in positive BCR signalling, but is not an exhaustive description of all signalling pathways or molecules activated. IKK, IκB kinase; PKC, protein kinase C."

```{r BCRsignalling, fig.cap='(ref:BCRsignalling)', out.width = "60%",  echo = FALSE}

knitr::include_graphics("figures/BCR_signalling.eps")

```

\@ref(fig:IGHVstatus) Taken from Kipps et al 2017: "Normal naive B cells that have undergone successful V(D)J recombination and express functional B cell receptors that are capable of binding to antigen interact with CD4+ T cells and accessory cells, which aggregate to form follicles that become germinal centres. Germinal cells each have a dark zone, comprising rapidly dividing B cells, and a light zone, comprising B cells mixed with follicular dendritic cells (FDCs), macrophages and helper T cells (TH cells). The B cells enter the dark zone of the germinal centre where they experience rapid proliferation and somatic hypermutation (SHM) in the genes encoding the immunoglobulin variable regions of the heavy chain (IGHV) and the light chain (IGVL). As they pass through to the light zone, the B cells that express the fittest B cell receptors for binding antigen are selected and may undergo immunoglobulin class-switch recombination. Chronic lymphocytic leukaemia (CLL) cells that use unmutated IGHV apparently originate from CD5+ B cells prior to experiencing SHM, whereas CLL cells that use mutated IGHV most likely originate from CD5+ B cells that have passed through and differentiated in the germinal centre. Some CLL cells might be derived from B cells that also have undergone immunoglobulin class-switch recombination and express immunoglobulin isotypes other than IgM and IgD, for example, IgG or IgA. Another subset is one with CLL cells that express immunoglobulin with only modest somatic mutations, such as CLL cells that use IGHV3-21 with ~97% homology to the inherited IGHV3-21 gene and an immunoglobulin light chain encoded by an unmutated IGLV3-21; these cells might derive from a B cell that has had constrained SHM, possibly owing to a limited need for immunoglobulin somatic diveresification and selection. Dashed arrows indicate speculated pathways." _Figure from @Kipps2017_. 

```{r IGHVstatus, fig.cap='(ref:IGHVstatus)', out.width="60%", echo = FALSE}

knitr::include_graphics("figures/IGHV_status.eps")

```

[@Kipps2017]
The BCR is composed of a ligand-binding trans-membrane immunoglobulin molecule (either IgA, IgD, IgE, IgG or IgM) and the signalling Igα (also known as CD79A)–Igβ (also known as CD79B) heterodimer. CLL cells typically co-express IgD and IgM, although at low levels compared with normal B cells; less than a few per cent of CLL cases express class-switched isotypes, most commonly IgG. The CD79A and CD79B molecules contain immunoreceptor tyrosine-based activation motifs, which can be phosphorylated following the crosslinking of surface immunoglobulin, thereby triggering BCR signalling. A functional BCR is required for the survival of mature B cells76 and is maintained in most mature B cell malignancies, including CLL. In CLL, evidence suggests that the surface immunoglobulin of CLL B cells is engaged by autoantigen, which leads to constitutive BCR signalling in vivo77–79. The importance of this interaction is underscored by the clinical success of kinase inhibitors that block BCR signalling (see Management), although effects on other receptors might also have a role80.
Like most cancers, CLL is heterogeneous and the outcome of BCR signalling ranges from enhanced B cell activation to B cell anergy81,82. The main pathways that lead to cell survival and proliferation downstream of the BCR are shown in FIG. 3, along with drugs targeted against key signalling intermediates. BCR signalling that leads to anergy is less well defined, but seems to involve biased activation of inhibitory molecules with only partial activation of the pathways that are typically associated with B cell activation81. One important molecule that may be involved is the inositol lipid phosphatase SHIP1. SHIP1 is activated by the tyrosine-protein kinase LYN and may limit B cell activation by counteracting phosphoinositide 3-kinase (PI3K) activity at both chronically engaged receptors and distant non-ligated BCRs, rendering them insensitive to stimulation82,83.
Enhanced B cell activation is more commonly observed in CLL that expresses unmutated IGHV, whereas anergy predominates in most cases of CLL that express mutated IGHV84. Anergy is a state of cellular lethargy induced by chronic engagement of the surface antigen receptors in the absence of adequate T cell help. Although capable of reversing their phenotype, anergic cells are less likely to proliferate in response to BCR signalling than more activated cells, which might, in part, account for the observation that patients with CLL cells that express mutated IGHV generally have more indolent disease than patients with CLL cells with unmutated IGHV85. The fate of the cell (activation versus anergy) might be influenced by the CLL cell of origin (FIG. 1), as the cell types that can form CLL differ in their patterns of DNA methylation73, and are likely to respond differently to autoantigens. An unresolved question is whether anergy can be reversed in vivo, mirroring what occurs in vitro78.
The BCR also coordinates the activity of other cell surface receptors, including integrins, such as α4β1 integrin. BCR stimulation can result in increased adhesion of CLL cells to α4β1 integrin substrates, for example, fibronectin and vascular cell adhesion protein 1 (REF. 86). By contrast, CXC-chemokine receptor 4 (CXCR4) is downmodulated by BCR engagement and both can trigger ‘inside-out’ signalling, resulting in the activation of α4β1 integrin87,88. Thus, recognized antigen encountered in lymphoid tissue is likely to affect adhesion and migration of CLL cells. Modulation of these pathways, coupled with the role of BTK and PI3K in chemokine receptor signalling89, contribute to the increased lymphocytosis observed in patients upon initiation of treatment with inhibitors of BTK or PI3K (see Management).


[@Kipps2017] 10.1038/nrdp.2016.96.Chronic
"CLL can be divided into two main subsets, which differ in their clinical behaviour. These subsets are distinguished by whether CLL cells express an unmutated or mutated immunoglobulin heavy-chain variable region gene (IGHV), reflecting the stage of normal B cell differentiation from which they originate1,2. CLL cells that express an unmutated IGHV originate from a B cell that has not undergone differentiation in germinal centres, which are the sites in the lymph nodes where B cells experience somatic hypermutation in their immunoglobulin variable region genes and selection during an immune response. Patients with CLL cells that express an unmutated IGHV typically have more-aggressive disease than patients with CLL cells that express a mutated IGHV. CLL cells with mutated IGHV arise from a post-germinal centre B cell that expresses immunoglobulin that has undergone somatic hypermutation and, in some cases, also immunoglobulin isotype switching (FIG. 1), similar to what occurs in normal B cells during an immune response to antigen. It should be emphasized that the high level of somatic mutations that arise in IGHV in the germinal centre are a natural part of affinity maturation of antibodies and, unlike mutations in other genes, are not pathological. The tumours are simply reflecting the stage of maturation of the parental B cell. In addition, some CLL cells have been described that are similar to unmutated IGHV CLL, but originate from B cells with limited somatic mutation, such as CLL with immunoglobulin heavy chains encoded by mutated IGHV3–21 and immunoglobulin light chains encoded by unmutated IGLV3–21 (REFS 3,4).
The repertoire of immunoglobin molecules produced by the CLL cells of all patients is considerably more limited than the repertoire of immunoglobulin molecules that can be made by the B cells of any one person5,6, reflecting the biased use in CLL of certain IGHV genes that have restricted somatic mutation and limited junctional and heavy-light chain combinatorial diversity. In as many as one-third of patients, the CLL cells express immunoglobulin ‘stereotypes’, which are stretches of primary structure in the variable region that can also be identified in the immunoglobulins produced by the CLL cells of other patients7. The restricted immunoglobulin repertoire in CLL is underscored by the finding that ~1 in 75 patients have CLL cells that express immunoglobulin molecules that are virtually identical8. The limited immunoglobulin diversity provides compelling evidence that CLL B cells are selected based on the binding activity of their expressed surface immunoglobulin, suggesting that B cell receptor (BCR) signalling plays a crucial part in CLL pathogenesis"


### Recurrent genetic features in CLL
Genetic alterations in CLL can include chromosomal alterations, mutations, alterations in the expression of mi RNAs and epigenetic modifications.

(for purpose of TLR sotry)
These included del(11q) and *ATM* ,
del(17p) and *TP53*, 
*SF3B1*. 


REading: 
From Kipps et al: 
38 Dohner H, et al. Genomic aberrations and survival in chronic lymphocytic leukemia. N. Engl. J. Med. 2000; 343:1910–1916. [PubMed: 11136261]

Del(13q)mechanism: 39. Klein U, et al. The DLEU2/miR-15a/16-1 cluster controls B cell proliferation and its deletion leads to chronic lymphocytic leukemia. Cancer Cell. 2010; 17:28–40. [PubMed: 20060366]

Del(17p) mechanism 40. Zenz T, Mertens D, Kuppers R, Dohner H, Stilgenbauer S. From pathogenesis to treatment of chronic lymphocytic leukaemia. Nat. Rev. Cancer. 2010; 10:37–50. [PubMed: 19956173]

del(11q)

Mutations and outcome: 
38: Dohner H, et al. Genomic aberrations and survival in chronic lymphocytic leukemia. N. Engl. J. Med. 2000; 343:1910–1916. [PubMed: 11136261]


JIC paper, explaon PACE 
Improved in recent years: 41: Van Dyke DL, et al. The Dohner fluorescence in situ hybridization prognostic classification of chronic lymphocytic leukaemia (CLL): the CLL Research Consortium experience. Br. J. Haematol. 2016; 173:105–113. [PubMed: 26848054]

CLl ahrbours a high degree of genetic diversity: 
Fabbri G, et al. Analysis of the chronic lymphocytic leukemia coding genome: role of NOTCH1 mutational activation. J. Exp. Med. 2011; 208:1389–1401. [PubMed: 21670202]
43. Pleasance ED, et al. A comprehensive catalogue of somatic mutations from a human cancer genome. Nature. 2010; 463:191–196. [PubMed: 20016485]
44. Puente XS, et al. Whole-genome sequencing identifies recurrent mutations in chronic lymphocytic leukaemia. Nature. 2011; 475:101–105. [PubMed: 21642962]
45. Wang L, et al. SF3B1 and other novel cancer genes in chronic lymphocytic leukemia. N. Engl. J. Med. 2011; 365:2497–2506. [PubMed: 22150006]

Functional role of mutations: 
WNT: 47 Wang L, et al. Somatic mutation as a mechanism of Wnt/beta-catenin pathway activation in CLL. Blood. 2014; 124:1089–1098. [PubMed: 24778153]

POT1: reference thereview

SF3B1: 
45 Wang L, et al. SF3B1 and other novel cancer genes in chronic lymphocytic leukemia. N. Engl. J. Med. 2011; 365:2497–2506. [PubMed: 22150006]
 48 Ferreira PG, et al. Transcriptome characterization by RNA sequencing identifies a major molecular and clinical subdivision in chronic lymphocytic leukemia. Genome Res. 2014; 24:212–226. [PubMed: 24265505]
 
 49. Quesada V, et al. Exome sequencing identifies recurrent mutations of the splicing factor SF3B1 gene in chronic lymphocytic leukemia. Nat. Genet. 2012; 44:47–52.
50. Te Raa GD, et al. The impact of SF3B1 mutations in CLL on the DNA-damage response. Leukemia. 2015; 29:1133–1142. [PubMed: 25371178]

Seqencing studies: 
two seminal studies: 
51. Landau DA, et al. Mutations driving CLL and their evolution in progression and relapse. Nature. 2015; 526:525–530. [PubMed: 26466571]
52. Puente XS, et al. Non-coding recurrent mutations in chronic lymphocytic leukaemia. Nature. 2015; 526:519–524. [PubMed: 26200345] References 51 and 52 describe landmark studies on whole-exome sequencing of CLL cells obtained from a large cohort of patients, reporting new driver mutations in CLL.


Epigenetic changes: 
Global hypomethylation and local hypermethylation
66. Cahill N, et al. 450K–array analysis of chronic lymphocytic leukemia cells reveals global DNA
methylation to be relatively stable over time and similar in resting and proliferative compartments. Leukemia. 2013; 27:150–158. [PubMed: 22922567]
67. Wahlfors J, et al. Genomic hypomethylation in human chronic lymphocytic leukemia. Blood. 1992; 80:2074–2080. [PubMed: 1382719]
68. Ziller MJ, et al. Charting a dynamic DNA methylation landscape of the human genome. Nature. 2013; 500:477–481. [PubMed: 23925113]

Methylation for classification: 
69. Kulis M, et al. Epigenomic analysis detects widespread gene-body DNA hypomethylation in chronic lymphocytic leukemia. Nat. Genet. 2012; 44:1236–1242. [PubMed: 23064414] This paper describes an analysis of DNA methylation in CLL cases, showing that the two molecular subtypes of CLL have differing DNA methylomes that seem to represent epigenetic imprints from distinct normal B cell subpopulations.

74. Bhoi S, et al. Prognostic impact of epigenetic classification in chronic lymphocytic leukemia: the
case of subset #2. Epigenetics. 2016; 11:449–455. [PubMed: 27128508]


\@ref(fig:CLLmutations) Taken from Kipps et al 2017: "Genes that are mutated in chronic lymphocytic leukaemia (CLL) are involved in several cellular pathways (blue boxes). As such, mutations in these genes can lead to a range of cellular consequences, such as aberrant DNA repair and B cell receptor (BCR) signalling, among others51,213. The minus sign from GBN1 to the MAPK–ERK pathway indicates negative regulation. *For more detail of the BCR and its associated signalling, see FIG. 3. ASXL1, additional sex combs-like protein 1; ATM, ataxia telangiectasia mutated; BAZ2A, bromodomain adjacent to zinc-finger domain protein 2A; BCOR, BCL-6 co-repressor; BIRC3, baculoviral IAP repeat-containing protein 3; BRCC3, BRCA1/BRCA2-containing complex subunit 3; C-NOTCH, carboxy-terminal domain of NOTCH; CARD11, caspase recruitment domain-containing protein 11; CHD2, chromodomainhelicase-DNA-binding protein 2; CHK2, checkpoint kinase 2; Co-A, co-activator; CSL, CBF1–Suppressor of Hairless–LAG1 (also known as RBPJ); DDX3X, ATP-dependent RNA helicase DDX3X; DYRK1A, dual-specificity tyrosine-phosphorylation-regulated kinase 1A; EGR2, early growth response 2; ELF4, ETS-related transcription factor Elf-4; ERK, extracellular signal-regulated kinase; EWSR1, Ewing sarcoma breakpoint region 1 protein; FBXW7, F-box/WD repeat-containing protein 7; FUBP1, far upstream element-binding protein 1; GNB1, guanine nucleotide-binding protein G(I)/G(S)/G(T) subunit β1; H3K4, histone H3 lysine 4; IC, intracellular domain; IKZF3, Ikaros family zinc-finger protein 3; IL-1R, IL-1 receptor; IRF4, interferon regulatory factor 4; ITPKB, inositol-trisphosphate 3-kinase B; LRP, low-density lipoprotein receptor-related protein; MAP2K1, dual-specificity mitogen-activated protein kinase kinase 1; MAPK, mitogen-activated protein kinase; MED12, Mediator of RNA polymerase II transcription subunit 12; MGA, MAX gene-associated protein; MYD88, myeloid differentiation primary response protein MyD88; NF-κB, nuclear factor-κB; NXF1, nuclear RNA export factor 1; P, phosphate; POT1, protection of telomeres protein 1; PTPN11, tyrosine-protein phosphatase non-receptor type 11; RIPK1, receptor-interacting serine/threonine-protein kinase 1; RPS15, 40S ribosomal protein S15; SAMHD1, SAM domain and HD domain-containing protein 1; SF3B1, splicing factor 3B subunit 1; SHP1, Src homology region 2 domain-containing phosphatase 1 (also known as PTPN6); SYK, spleen tyrosine kinase; TCF/LEF, T cell factor/lymphoid enhancer factor; TLR8, Toll-like receptor 8; TNFR1, tumour necrosis factor receptor 1 (also known as TNFRSF1A); TRAF, TNFR-associated factor; XPO, exportin; ZMYM3, zinc-finger MYM-type protein 3. Adapted with permission from REF. 51, Macmillan Publishers Limited." _Figure from @Kipps2017_

```{r CLLmutations, fig.cap='(ref:CLLmutations)', out.width="60%", echo = FALSE}

knitr::include_graphics("figures/genetic_mutations.eps")

```


*mutations*
[@Kipps2017] 
Several large genetic studies have revealed numerous genetic alterations in CLL, including single- nucleotide polymorphisms (SNPs), chromosomal alterations and alterations in non- coding RNA, such as microRNA (miRNA), some of which can be used to determine prognosis and to guide management strategies. Interactions between CLL cells and their microenvironment, including interactions with other cell types, such as T cells, nurse-like cells and stromal cells, can induce B cell proliferation and contribute to disease.

[@Kipps2017]
The altered expression of genes that are located in or near CLL-associated SNPs might contribute to disease development. For example, a SNP in IRF4 is associated with low expression of interferon regulatory factor 4; mice that are deficient in this protein can develop CLL23, partly owing to hyperactivation of Notch signalling24. 

23: Shukla V, Ma S, Hardy RR, Joshi SS, Lu R. A role for IRF4 in the development of CLL. Blood. 2013; 122:2848–2855. [PubMed: 23926303]
24. Shukla V, Shukla A, Joshi SS, Lu R. Interferon regulatory factor 4 attenuates notch signaling to suppress the development of chronic lymphocytic leukemia. Oncotarget. 2016; 7:41081–41094. [PubMed: 27232759]


[@Kipps2017]
Somatic mutations—The advent of massively parallel sequencing and the application of whole-exome sequencing to CLL have transformed our understanding of the genetic heterogeneity of CLL and have established that CLL harbours a high degree of genetic variability42–45 (FIG. 2). From these studies, recurrent somatic mutations have been consistently observed in genes that have a role in DNA damage (for example, TP53 and ATM), mRNA processing (for example, SF3B1 and XPO1), chromatin modification (for example, HIST1H1E, CHD2 and ZMYM3), WNT signalling, Notch signalling (for example, NOTCH1) and inflammatory pathways (for example, MYD88). Other mutations, such as those found in EGR2 or BRAF, can affect B cell-related signalling and transcription46 (FIG. 2).
The functional role of several putative driver mutations has been confirmed; for example, silencing mutated WNT pathway genes in primary CLL cells resulted in decreased cell viability47. Mutations in POT1, which has a role in the protection of telomeres, prevented the binding of protection of telomeres protein 1 to telomeric DNA, resulting in numerous chromosomal abnormalities, in addition to the development of abnormal telomeres. Mutations in SF3B1 have been found to be associated with aberrant RNA splicing45,48,49 and an altered DNA-damage response50. SAMHD1 encodes a protein involved in the regulation of intracellular deoxy-nucleotide pools, which are recruited to the site of DNA damage and are probably involved in the response to DNA double-strand breaks50.
The detection of somatic mutations and their relative frequencies is variable, which possibly reflects differences in the composition of the cohorts studied worldwide. Two seminal studies have provided the largest sequenced collections to date51,52, in which >500 CLL samples were characterized by whole-exome sequencing or whole- genome sequencing. The clinical and/or biological features of the patients examined in each study were notably distinct; one study analysed matched pretreatment samples from patients who required initial treatment and noted mutations in SF3B1 (21% of patients), ATM (15% of patients), TP53 (7% of patients), NOTCH1 (6% of patients) or BIRC3 (4% of patients). The other study assessed patients with earlystage CLL and patients with monoclonal B cell lymphocytosis and identified NOTCH1 (12.6% of patients), A T M (11% of patients), BIRC3 (8.8% of patients) and SF3B1 (8.6% of patients) as the most frequently mutated genes.
These large sample cohorts have provided the sensitivity to discover novel candidate cancer genes that are altered in CLL. Both studies also identified somatic mutations in MGA and PTPN11, which encode modulators of MYC, IKZF3, which encodes a key transcription factor, and RPS15, which encodes 40S ribosomal protein S15 and is recurrently mutated in ~20% of patients who relapse after combination chemotherapy53. Other recurrent somatic mutations include those in the 3′ untranslated region of NOTCH1 and a PAX5 enhancer, which increases the expression of these B cell-associated transcription factors54,55. Patients with mutations in the 3′ untranslated region of NOTCH1 have a shorter time from diagnosis to treatment and poorer overall survival, similar to that of patients with non- synonymous mutations, which alter the amino acid sequence of NOTCH1.
Next-generation sequencing has revealed intra-tumoural heterogeneity in CLL. Some somatic mutations, such as those in MYD88, or chromosomal abnormalities, such trisomy 12 or del(13q), are most often found in all the CLL cells of any one patient, indicating that these genetic alterations occurred early in the evolution of the leukaemia. Other mutations, such as those found in SF3B1 or NOTCH1, or chromosomal alterations, such as del(17p), are typically found in only a fraction of the leukaemia cells and thus represent subclonal events, which occur after the development of CLL. Across studies, subclonal driver mutations are associated with more-aggressive disease, particularly when two or more are found concurrent in a leukaemia cell population51,56,57. In addition, studies have demonstrated that large clonal shifts can occur following chemotherapy, owing to increases in the proportions of CLL cells that have a TP53 mutation or del(17p), indicating that such genetic changes provide a strong fitness advantage in the setting of therapy51. By contrast, one study of CLL cells from patients treated with ibrutinib (an inhibitor of Bruton tyrosine kinase (BTK)), revealed mutations associated with drug resistance that were distinct from those observed in CLL cells of patients treated with standard chemotherapy58. CLL was the first human disease that was found to be associated with alterations in miRNA.


(can compare size of large studies with ours)



*copy number variants*
[@Kipps2017]
Approximately 80% of patients with CLL carry at least one of four common chromosomal alterations: a deletion in chromosome 13q14.3 (del(13q)), del(11q), del(17p) and trisomy 12 (REF. 38). Del(13q) is the most common chromosomal alteration, evident in >50% of patients, and is associated with favourable prognosis. Within this deleted region is the DLEU2-mir-15–16 cluster, which regulates the expression of proteins that can inhibit apoptosis or that are involved in cell cycle progression39 (see Supplementary information S2 (table)). Del(17p) is found in 7% of patients and is associated with loss of the tumour suppressor gene TP53 (REF. 40), whereas del(11q) is found in 18% of patients and is often associated with alterations in ATM (which encodes a protein involved in DNA repair); each of these chromosomal alterations is associated with adverse clinical outcome38, although this has improved in recent years41. Trisomy 12 is found in 16% of patients with CLL and is associated with an intermediate prognosis. Unlike the neoplastic B cells in mantle cell lymphoma, CLL cells do not have the translocation t(11;14) (q13;q32) or other genetic alterations that enhance the expression of CCND1, which encodes the cell cycle regulator cyclin D1

*Epigenetic changes*
The CLL epigenome shows global hypomethylation combined with local hypermethylation, as has been observed in other cancers66–68. Indeed, comprehensive methylation profiling has demonstrated substantial intra-tumoural methylation heterogeneity52,69–73. Increasing methylation heterogeneity has consistently been associated with increased genetic complexity owing to the acquisition of subclonal mutations, thus linking genomic and methylomic evolution in CLL52,70,71. Indeed, locally disordered methylation in CLL might enhance the evolutionary adaptive capacity of CLL cells by increasing the background ‘noise’ of the genome, thereby providing increased opportunities for somatic mutations within the leukaemia clone. In support of this notion is the observed association between methylation evolution and adverse clinical outcome52,70,71.

[@Kipps2017]
Methylation signatures can classify distinct clinical CLL subgroups69,74. As these methylation patterns are a heritable trait, they have been used to ‘trace’ back to the type of normal B cell from which the CLL cells were derived75. These studies revealed that the CLL cells of different patients derive from a continuum of B cell maturation states, which are not restricted to discrete maturation stages. Nevertheless, CLL cells that use unmutated IGHV versus mutated IGHV generally have distinctive methylation patterns, which respectively approximate to those of pre-germinal centre versus post-germinal centre memory B cells, as depicted in FIG. 1. The diversity in the likely cells of origin of CLL cells highlights the biological and phenotypic heterogeneity of this disease. These findings also suggest that epigenetic programming that is dependent of transcription factors has a potentially important role in the development of CLL.


### The incompletely understood role of trisomy12 {#intro-trisomy12}
Bosch, F., Dalla-Favera, R. Chronic lymphocytic leukaemia: from genetics to treatment. Nat Rev Clin Oncol 16, 684–701 (2019). https://doi.org/10.1038/s41571-019-0239-8

Trisomy 12 chronic lymphocytic leukemia expresses a unique set of activated and targetable pathways
DOI: https://doi.org/10.3324/haematol.2018.190132
Issue: Vol. 103 No. 12 (2018): December, 2018

The Protein Landscape of Chronic Lymphocytic Leukemia (CLL)
Fabienne Meier-Abt https://doi.org/10.1182/blood.2020009741

what's understood about the overespssopn of genes and proliferative drive; 
DOI: 10.1200/JCO.2005.02.568 Journal of Clinical Oncology 23, no. 16 (June 01, 2005) 3780-3792. 

### Stratification of CLL patients
Reading: 
(from Kipps)

12. Nabhan C, et al. The impact of race, ethnicity, age and sex on clinical outcome in chronic lymphocytic leukemia: a comprehensive Surveillance, Epidemiology, and End Results analysis in the modern era. Leuk. Lymphoma. 2014; 55:2778–2784. [PubMed: 24588735]
13. Li Y, Wang Y, Wang Z, Yi D, Ma S. Racial differences in three major NHL subtypes: descriptive epidemiology. Cancer Epidemiol. 2015; 39:8–13. [PubMed: 25560974]
14. Pulte D, Redaniel MT, Bird J, Jeffreys M. Survival for patients with chronic leukemias in the US and Britain: age-related disparities and changes in the early 21st century. Eur. J. Haematol. 2015; 94:540–545. [PubMed: 25315799]

\@ref(fig:CLLstratification) Taken from Kipps et al 2017: "Indications for therapy of patients with chronic lymphocytic leukaemia (CLL) include late-stage disease, evidence for rapid disease progression or disease-related symptoms. Patients with del(17p) or mutated TP53 should be treated with therapy that does not require functional TP53, such as ibrutinib (a Bruton tyrosine kinase (BTK) inhibitor), given the relatively poor outcome for such patients with chemotherapy. For patients without del(17p) or known mutations in TP53, immunoglobulin heavy-chain variable region (IGHV) mutational status can help to define the treatment strategy; patients with unmutated IGHV could be considered for therapy with a BTK inhibitor (such as ibrutinib) and patients with mutated IGHV might be good candidates for chemoimmunotherapy (CIT), if amenable. Indeed, patients with mutated IGHV can have excellent outcomes with CIT regimens, such as fludarabine, cyclophosphamide and rituximab, with >50% of patients having a median progression-free survival of >10 years, including the potential for cure. If the patient is amenable to CIT, age, medical comorbidities and myeloid reserve should be taken into consideration. Patients >65 years of age commonly have medical comorbidities and are less able to tolerate myelosuppressive regimens, such as fludarabine, cyclophosphamide and rituximab. Thus, considerations should be given to using reduced dose or less myelosuppressive chemotherapy regimens, such as chlorambucil or reduced-dose bendamustine and an anti-CD20 monoclonal antibody for patients with limited myeloid reserve. Patients who either do not respond, have a poor tolerance to CIT or relapse following CIT, should be re-evaluated for del(17p) or TP53 mutations. Patients who develop de novo del(17p) or TP53 mutations, or have known del(17p) and/or TP53 mutations, or who develop resistance or intolerance to ibrutinib, could be considered for therapy with idelalisib and rituximab or the BCL-2 inhibitor venetoclax. Patients treated with CIT who do not have del(17p) or TP53 mutations could be considered for repeat CIT if their progression-free survival after CIT is >2 years and the patient has sufficient myeloid reserve. Such patients also might be treated with a BTK inhibitor or a phosphoinositide 3-kinase (PI3K) inhibitor, which also could be considered for patients who develop intolerance or resistance to therapy with ibrutinib. Patients who develop resistance or intolerance to inhibitors of BTK, PI3K and/or BCL-2 should be considered for clinical trials or alternative agents. LDT, lymphocyte doubling time." _Figure from @Kipps2017 _.

```{r CLLstratification, fig.cap='(ref:CLLstratification)', out.width = '60%', echo = FALSE}

knitr::include_graphics("figures/disease_stratification.eps")

```


Disease staging
[@Kipps2017] Two clinical staging systems are widely used to divide patients with CLL into three broad prognostic groups114,115. The Rai staging system (TABLE 1) is more commonly used in the United States, whereas the Binet classification (TABLE 2) is more commonly used in Europe. The staging systems each recognize the importance of marrow function and define late-stage, or high-risk, disease by the presence of pronounced anaemia or thrombocytopenia.

[@Kipps2017]
Prognostic factors that can help to identify patients who require therapy relatively soon after diagnosis include certain clinical features and genetic, molecular and biochemical characteristics of the CLL cell. Multivariable models, prognostic indexes116–118 and nomograms119 have been developed to consolidate such prognostic factors so that they can more robustly predict clinical outcome. Commonly used parameters that are associated with poorer outcome are male sex,
65 years of age, poor performance status due to medical
comorbidities, certain CLL cell characteristics, such as the expression of unmutated IGHV1,2, ZAP70 (REFS 120,121), CD49d (also known as integrin α4)122 or CD38 (REF.2), the presence of del(17p)38 or del(11q)123, high serum levels of β2-microglobulin (>3.5 mg per l)124, complex karyotype (that is, the presence of three or more chromosomal aberrations observed on a karyotype test)125,126, or a high absolute lymphocyte count (>50,000 cells per µl) and/or late-stage disease at initial presentation. Del(17p) is often associated with inactivating mutations in TP53 and is a predictor of poor outcome to treatment with regimens that involve conventional chemotherapy127.

[@Kipps2017]
Generally, indications to initiate therapy include pronounced disease-related anaemia or thrombocytopenia (patients with Rai stage III or stage IV disease, or Binet stage C disease), symptomatic lymphadenopathy and/or symptoms that are associated with active disease, such as night sweats, fatigue, unintentional weight loss or fever without evidence of infection107. However, when basing a treatment decision on constitutional symptoms alone, the physician should consider other medical conditions, such as hypothyroidism, hyperthyroidism, hypoglycaemia, chronic inflammation, uncommon opportunistic infections or sleep disorders, including sleep apnoea.

For patients who need treatment, the presence of del(17p) or mutated TP53 are the most important features that are currently directing the choice of therapy (FIG. 8). Next, advanced age of >65 years, the presence of medical comorbidities and the objectives of treatment have substantial bearing on the choice of therapy. Increasingly, IGHV mutational status is considered as a parameter when determining the type of therapy; for example, chemotherapy-based regimens are reserved for patients with CLL and mutated IGHV. Conversely, the specific Rai or Binet stage of the patient who requires treatment does not necessarily influence the choice of therapy.
Systemic

### Prognosis and where the problems lie
*harsh treatments, especially when most are elderly*
*not a proper cure - combo therapies may cover all bases* 
*heterogenous - combo therapies may cover all bases* 
The clinical course of newly diagnosed CLL is extremely variable; some patients remain free of symptoms and are fully active for decades, whereas others rapidly become symptomatic or develop high-risk disease, which requires treatment soon after diagnosis and might result in death due to therapy-related and/or disease-related complications. However, most patients have a clinical course that is in between these two extremes.

[@Kipps2017]
Adverse effects of ibrutinib include fatigue, diarrhoea, bleeding, ecchymoses, rash, arthralgia, myalgia, increased blood pressure and atrial fibrillation. Clinical trials are currently evaluating second-generation BTK inhibitors (for example, acalabrutinib148, ONO/ GS-4059 (REF. 149) or BGB-3111) to determine whether any one of these drugs has a superior therapeutic index than that of ibrutinib150.


[@Kipps2017] Toxicities of venetoclax include gastrointestinal disturbances, neutropenia and tumour lysis syndrome159, which is characterized by hyperkalaemia, hyper-uricaemia and/or azotaemia. Tumour lysis syndrome results from the rapid destruction of cancer cells and the release of their cellular contents into the blood. Tumour lysis syndrome typically occurs when initiating venetoclax therapy or when dosing is increased. Thus, patients start venetoclax with a low daily dose, which is escalated each week over 5 weeks to mitigate the risk of developing tumour lysis syndrome. Even with this strategy, patients who are at high risk for tumour lysis syndrome because of bulky lymphadenopathy and/or lymphocytosis of >25,000 cells per µl must be hydrated and closely monitored during therapy initiation and during dose escalation.

[@Kipps2017]
The outlook for patients with CLL has improved substantially over the past several years. Through research on the immune biology and genetics of CLL, patients can be stratified into subgroups with distinctive clinical features, which has improved our capacity to assess prognosis or govern therapy. However, an understanding of the mechanisms that contribute to immune dysfunction or how it contributes to autoimmune disease, such as autoimmune haemolytic anaemia, therapy resistance or therapy-related complications is unknown. Whether tyrosine kinase inhibitors can affect clonal evolution, induce and/or select for drug resistance, or can achieve complete responses if used earlier in the course of the disease is also unknown.

New therapies:   
[@Kipps2017]
Several therapies are currently under preclinical and clinical investigation for the treatment of patients with CLL, including new drugs and treatment modalities that can modulate the immune system, and cell transplantation.

*Car-T Therapies*
[@Kipps2017]
CD19-targeted CAR T cells have yielded long-term PFS and relapse-free survival durations in patients with CLL; in 14 patients with relapsed or refractory CLL, four patients achieved a complete response and four patients achieved a partial response197. None of the complete responders had MRD and none relapsed (a median follow-up of 19 months). However, the efficacy of CAR T cell therapy in patients with CLL has been modest compared with that in patients with acute lymphoblastic leukaemia; this might be owing to qualitative defects in the T cells of patients with CLL, who are generally older than patients with acute lymphoblastic leukaemia and already have immune dysfunction that reflects disease-associated anergy (see BCR and B cell signalling). Ibrutinib might partially correct some of these defects198. Larger

*Allogenic stem cell transplantation*
[@Kipps2017]
Allogeneic stem cell transplantation is a potentially curative strategy for patients with relapsed or refractory CLL, including patients with high-risk features such as del(17p).Research efforts are ongoing to develop better-tolerated cell-based therapy with a similar curative potential that can be used without the immunosuppression and associated long-term morbidity and mortality of allogeneic stem cell transplantation.
Donor availability, advanced patient age, associated toxicities of myelosuppression, graft- versus-host disease and impaired resistance to infections limit the application of allogeneic stem cell transplantation in patients with CLL. In addition, the advent of BCR signalling inhibitors and BCL-2 inhibitors provide multiple treatment options that afford well- tolerated, long-term disease control, making allogeneic stem cell transplantation the least desirable option for most patients.

*Immune-modulatory drugs*
[@Kipps2017]
Immune-modulatory drugs, such as thalidomide and lenalidomide, are approved for the treatment of patients with multiple myeloma, mantle cell lymphoma or myelodysplastic disease. Although these drugs have clinical activity in patients with CLL, they have had limited application unless used in combination with an anti-CD20 monoclonal antibody166,186. In CLL, lenalidomide can induce the expression of p21WAF1/CIP1, which inhibits cyclin-dependent kinase and CLL cell proliferation187, and can improve immune synapse formation, potentially enhancing immune function.

*Immune Checkpoint inhibitors*
[@Kipps2017]
The immune checkpoint receptor PD-1, and its ligands PD-L1 and PD-L2, is the most important cognate receptor involved in the suppression of cellular immune activation. CLL cells express high levels of PD-L1 and PD-L2 and can suppress the responses of PD-1- expressing effector T cells106, leading to an exhausted (that is, no longer functional) T cell phenotype. Preclinical studies in mouse models have demonstrated that checkpoint inhibitors can reactivate immune effector cells to have anti-leukaemia activity202. However,

*Combination targeted therapies*
[@Kipps2017]
We can now target the distinctive phenotypic or physiological features of CLL with targeted therapeutic agents, which have a higher therapeutic index than standard chemotherapy. Through the use of combination therapy, which targets different B cell survival signalling pathways and/or achieves better eradication of CLL cells, we might be able to define curative treatments for most patients with this disease.
Research on leukaemia cell survival signalling pathways, such as those governed by interactions between leukaemia cells and cells or secreted factors within the microenvironment (FIG. 4), might identify pathways that are not affected by BCR inhibitors. For example, BCR inhibitors, such as ibrutinib, cannot block ROR1-dependent WNT5A signalling, which enhances CLL cell proliferation, migration and survival93; as such, antibodies that block ROR1-dependent signalling could potentially have synergistic activity when used in combination with BCR inhibitors203. Furthermore, the interaction between CLL cells and accessory cells in the microenvironment might enhance CLL cell expression of anti-apoptotic proteins other than BCL-2, such as MCL1, thereby contributing to therapy resistance. As such, the therapeutic use of a selective BCL-2 antagonist, such as venetoclax, might be more effective when used in combination with BCL-2 inhibitors161,162, which also interfere with the homing of CLL cells to the microenvironment. Conceivably, combination target therapy with agents that have synergistic activity will provide highly effective and potentially curative treatment of patients with CLL.

## The tumour microenvironment in CLL 
### The concept of the protective niche 
_the behaoir of CLL cells in realtion to the ME_
[@Kipps2017]
CLL cells depend on survival signals that they receive in lymphoid tissues from neighbouring non-neoplastic cells within the so-called cancer microenvironment. CLL cells follow chemokine gradients into lymph nodes, where they form ‘proliferation centres’ (REF. 77), as opposed to normal germinal centres. In these proliferation centres, the CLL cells contact non- malignant stromal cells, nurselike cells (also known as lymphoma-associated macrophages), T cells and mesenchymal-derived stromal cells (FIG. 4). Engagement with autoantigen may occur during this transit, thereby stimulating CLL cell activation and proliferation if sufficient T cell help is available. Only a few per cent of the CLL cells undergo proliferation at any one time; the remainder of the cells are either unstimulated or driven into anergy84. However, within such proliferation centres, all CLL cells are exposed to chemokines, integrins, cytokines and survival factors.

CLL cells circulating
in the blood are mostly found in a resting state, up to 3 % of cancer cells start to proliferate
in the lymph nodes (Herndon et al., 2017).

When they leave the lymph node: 
[@Kipps2017]
As CLL cells leave the tissue site, engagement with antigen will be transient and its effects are likely to reverse in the blood, leading to variable increases in the expression of surface IgM and CXCR4 (REF. 98). CLL cell expression of CXCR4 is downmodulated upon exposure to CXC-chemokine ligand 12 (CXCL12)99, which is produced by nurse-like cells in the microenvironment100. Consequently, CLL cells in the blood that have just exited lymphoid tissue express low levels of CXCR4 (known as CXCR4dim cells) and higher levels of CD5 (known as CD5bright cells) relative to the CLL cells that are poised to re-enter lymphoid compartments101. For unexplained reasons, a high level of expression of CXCR4 by circulating CLL cells is associated with poorer prognosis in patients with CLL that use mutated IGHV102, possibly by influencing tissue re-entry. In terms of treatment effects, kinase inhibitors, such as ibrutinib, inhibit BCR-associated pathways, which remain important for cancer cells that are retained in lymphoid tissues, but can also directly inhibit integrin-mediated and chemokine- mediated pathways, thereby contributing to the increased lymphocytosis that occurs following the initiation of kinase inhibitor therapy103.


_lymph nodes, bone marrow, blood_
the lymph node and the bone marrow Burger, J. A. and O’Brien, S. (2018). Evolution of CLL treatment — from chemoimmunotherapy to targeted and individualized therapy. Nature Reviews Clinical Oncology 15, 510–527 <- this isnt really the most direct 






_cell-cell contacts_
gneaux, L., Delforge, A., Bron, D., De Bruyn, C. and Stryckmans, P. (1998). Chronic lym-
phocytic leukemic B cells but not normal B cells are rescued from apoptosis by contact with
normal bone marrow stromal cells. Blood 91, 2387–96

anayiotidis, P., Jones, D., Ganeshaguru, K., Foroni, L. and Hoffbrand, A. V. (1996). Human
bone marrow stromal cells prevent apoptosis and support the survival of chronic lymphocytic
leukaemia cells in vitro. British Journal of Haematology 92, 97–103

Herbst, S. A., Kim, V., Schitter, E., Roider, T., Bruch, P., Kolb, C., Knoll, M., Lu, J., H ̈ullein,
J., von Kalle, C., Lutz, C., Dreger, P., M ̈uller-Tidow, C., Zenz, T., Huber, W. and Dietrich,
S. (2020a). The Influence of the Bone Marrow Niche on Drug Response Phenotypes of Blood
Cancers. Manuscript in preparation.

By providing these signals from the microenvironment to CLL cells cultured ex-vivo spontaneous
apoptosis can be prevented. Thus, it has been shown that BMSCs, one of the most important
interaction partners of B-cells in the bone marrow niche (Murphy, 2012), can rescue CLL cells
from spontaneous apoptosis (Herbst et al., 2020a; Lagneaux et al., 1998; Panayiotidis et al.,1996



_soluble factors_ 

within proliferation centres, all CLL cells are exposed to chemokines, integrins, cytokines and survival factors (such as tumour necrosis factor (TNF) ligand super-family member 13B (also known as BAFF) or TNF ligand superfamily member 13 (also known as APRIL)), which activate canonical nuclear factor-κB (NF-κB)90, before they exit to the blood. Activation of NF-κB can induce the expression of mir-155, which enhances BCR signalling and activation by reducing the expression of INPP5D, which encodes SHIP1 (REF. 65). Cytokines that are secreted by T cells, such as IL-4, can upregulate surface IgM, which potentially facilitates the interaction of the CLL cell with autoantigen91. In addition, the elaboration of various WNT proteins by cells in the microenvironment can activate canonical and non-canonical WNT signalling pathways92,93. Activation of the tyrosine-kinase-like transmembrane receptor ROR1 by WNT5A can induce the activation of RAC1 and RHOA, and thereby enhance CLL cell proliferation and promote migration in response to chemokines93; in part, for this reason, high-level CLL cell expression of ROR1 is associated with accelerated disease progression94. Finally, Notch signalling in response to Jagged, or Hedgehog signalling in response to Sonic Hedgehog or Desert Hedgehog, can provide pro-survival stimulation for at least a subset of patients with CLL, particularly those with trisomy 12 (REFS 95–97).

[90] BADD and APRIL activate nfkb:  Endo T, et al. BAFF and APRIL support chronic lymphocytic leukemia B-cell survival through activation of the canonical NF-kappaB pathway. Blood. 2007;109:703–710.

[65.] ACtivation of nkb enhances bcr signalling: Cui B, et al. MicroRNA-155 influences B-cell receptor signaling and associates with aggressive disease in chronic lymphocytic leukemia. Blood. 2014;124:546–554. 

[91] Aguilar-Hernandez MM, et al. IL-4 enhances expression and function of surface IgM in CLL cells. Blood. 2016;127:3015–3025

Finally, Notch signalling in response to Jagged, or Hedgehog signalling in response to Sonic Hedgehog or Desert Hedgehog, can provide pro-survival stimulation for at least a subset of patients with CLL, particularly those with trisomy 12: 
95. Seke Etet PF, Vecchio L, Nwabo Kamdje AH. Interactions between bone marrow stromal microenvironment and B-chronic lymphocytic leukemia cells: any role for Notch, Wnt and Hh signaling pathways? Cell Signal. 2012;24:1433–1443. 

96. Decker S, et al. Trisomy 12 and elevated GLI1 and PTCH1 transcript levels are biomarkers for Hedgehog-inhibitor responsiveness in CLL. Blood. 2012;119:997–1007. [PubMed] [Google Scholar]
97. De Falco F, et al. Notch signaling sustains the expression of Mcl-1 and the activity of eIF4E to promote cell survival in CLL. Oncotarget. 2015;6:16559–16572. [PMC free article] [PubMed] [Google Scholar]

[77] The lymph node microenvironment promotes B-cell receptor signaling, NF-κB activation, and tumor proliferation in chronic lymphocytic leukemia 2011 10.1182/blood-2010-05-284984

_cutting edge of whats knwon about how it operates in vivo_ 

_whats the cutting edge of wahts knwon about drug resistance_
As well as providing protective signals that support tumour growth, the microenvironment is also believed to reduce treatment efficacy within the protective niche [@Ahn2018] 

Bakker, E., Qattan, M., Mutti, L., Demonacos, C. and Krstic-Demonacos, M. (2016). The role
of microenvironment and immunity in drug response in leukemia. Biochimica et Biophysica
Acta 1863, 414–426

During therapy, especially with new targeted agents, clearance of leukemia cells from the blood
can often be achieved. However, in most cases minimal residual disease, particularly in the bone
marrow, persists (Burger and O’Brien, 2018). Due to this, it is assumed that the niche of the
bone marrow protects CLL cells from drug-induced apoptosis and, thereby, contributes to the
emergence of resistant clones (Ten Hacken and Burger, 2016).

in vitro: 
Few studies exist which have investigated the influence of the bone marrow on the response to selected drugs to further understand the mechanisms of drug resistance.  

Examples: The chemotherapeutics fludarabine, oxaliplatin, chlo-rambucil, cyclophosphamide and doxorubicine have been described to have diminished effects in co-cultures with stroma cells (Kay et al., 2007; Kurtova et al., 2009; Mraz et al., 2011; Zhang et al., 2012):

Kay, N. E., Shanafelt, T. D., Strege, A. K., Lee, Y. K., Bone, N. D. and Raza, A. (2007). Bone
biopsy derived marrow stromal elements rescue chronic lymphocytic leukemia B-cells from
spontaneous and drug induced cell death and facilitates an ”angiogenic switch”. Leukemia
Research 31, 899–906.

Kurtova, A. V., Balakrishnan, K., Chen, R., Ding, W., Schnabl, S., Quiroga, M. P., Sivina, M.,
Wierda, W. G., Estrov, Z., Keating, M. J., Shehata, M., J ̈ager, U., Gandhi, V., Kay, N. E.,
Plunkett, W. and Burger, J. A. (2009). Diverse marrow stromal cells protect CLL cells from
spontaneous and drug-induced apoptosis: Development of a reliable and reproducible system
to assess stromal cell adhesion-mediated drug resistance. Blood 114, 4441–4450

Mraz, M., Zent, C. S., Church, A. K., Jelinek, D. F., Wu, X., Pospisilova, S., Ansell, S. M.,
Novak, A. J., Kay, N. E., Witzig, T. E. and Nowakowski, G. S. (2011). Bone marrow stromal
cells protect lymphoma B-cells from rituximab-induced apoptosis and targeting integrin α-4-
β-1 (VLA-4) with natalizumab can overcome this resistance. British Journal of Haematology
155, 53–64

Zhang, W., Trachootham, D., Liu, J., Chen, G., Pelicano, H., Garcia-Prieto, C., Lu, W., Burger,
J. A., Croce, C. M., Plunkett, W., Keating, M. J. and Huang, P. (2012). Stromal control of
cystine metabolism promotes cancer cell survival in chronic lymphocytic leukaemia. Nature
Cell Biology 14, 276–286.

reduced efficiency for the corticosteroid medicationshydrocortisone and dexamethasone (Kurtova et al., 2009; Panayiotidis et al., 1996). 

Kurtova, A. V., Balakrishnan, K., Chen, R., Ding, W., Schnabl, S., Quiroga, M. P., Sivina, M.,
Wierda, W. G., Estrov, Z., Keating, M. J., Shehata, M., J ̈ager, U., Gandhi, V., Kay, N. E.,
Plunkett, W. and Burger, J. A. (2009). Diverse marrow stromal cells protect CLL cells from
spontaneous and drug-induced apoptosis: Development of a reliable and reproducible system
to assess stromal cell adhesion-mediated drug resistance. Blood 114, 4441–4450

protection of CLL cells from carilzomib-induced apoptosis has been mentioned (Gupta
et al., 2013). 

Gupta, S. V., Hertlein, E., Lu, Y., Sass, E. J., Lapalombella, R., Chen, T. L., Davis, M. E.,
Woyach, J. A., Lehman, A., Jarjoura, D., Byrd, J. C. and Lucas, D. M. (2013). The Pro-
teasome Inhibitor Carfilzomib Functions Independently of p53 to Induce Cytotoxicity and an
Atypical NF- B Response in Chronic Lymphocytic Leukemia Cells. Clinical Cancer Research
19, 2406–2419.

IL4 induced resistance to ibrutinib - aguilar - hernandes: 


Tlr induced restiance to venetoclax: 

 stroma cell mediated resistances to ibrutinib
treatment (Cheng et al., 2014; Guo et al., 2017), which might explain why complete eradication
of leukemia cells from the bone marrow is rare during therapy with ibrutinib (Byrd et al., 2013).
also enalrged lymph nodes with ibrutinib treatment: 

Cheng, S., Ma, J., Guo, A., Lu, P., Leonard, J. P., Coleman, M., Liu, M., Buggy, J. J., Furman,
R. R. and Wang, Y. L. (2014). BTK inhibition targets in vivo CLL proliferation through its
effects on B-cell receptor signaling activity. Leukemia 28, 649–57

Guo, A., Lu, P., Coffey, G., Conley, P., Pandey, A. and Wang, Y. L. (2017). Dual SYK/JAK
inhibition overcomes ibrutinib resistance in chronic lymphocytic leukemia: Cerdulatinib, but
not ibrutinib, induces apoptosis of tumor cells protected by the microenvironment. Oncotarget
8, 12953–12967.

Byrd, J. C., Furman, R. R., Coutre, S. E., Flinn, I. W., Burger, J. A., Blum, K. A., Grant,
B., Sharman, J. P., Coleman, M., Wierda, W. G., Jones, J. A., Zhao, W., Heerema, N. A.,
Johnson, A. J., Sukbuntherng, J., Chang, B. Y., Clow, F., Hedrick, E., Buggy, J. J., James,
D. F. and O’Brien, S. (2013). Targeting BTK with ibrutinib in relapsed chronic lymphocytic
leukemia. The New England Journal of Medicine 369, 32–42.

Strategies to overcome: 
One suggestion for overcoming these resistances is the simultaneous targeting of SYK and JAK
with the drug cerdulatinib (Guo et al., 2017). 


Guo, A., Lu, P., Coffey, G., Conley, P., Pandey, A. and Wang, Y. L. (2017). Dual SYK/JAK
inhibition overcomes ibrutinib resistance in chronic lymphocytic leukemia: Cerdulatinib, but
not ibrutinib, induces apoptosis of tumor cells protected by the microenvironment. Oncotarget
8, 12953–12967.

the BCL-2 inhibitor AT-101 was ble to overcome stroma mediated protection, which might represent an interesting option for avoiding resistance (Balakrishnan et al., 2009). 

Balakrishnan, K., Burger, J. A., Wierda, W. G. and Gandhi, V. (2009). AT-101 induces apopto-
sis in CLL B cells and overcomes stromal cell-mediated Mcl-1 induction and drug resistance.
Blood 113, 149–53.


These investigations provide important insights
into the effect of stroma cells on drug action. However, to date a systematic investigation of the
effect of stroma cells on drug response is lacking. Only a comprehensive evaluation will enable
a better understanding of these interactions and allow the development of efficient strategies to
overcome the emergence of drug resistances in the bone marro

_whats the studies that have been done? what otehr approachs have poelpe used_
Studies on each pathway: 
Kyle Crassini, Yandong Shen, Stephen Mulligan & O. Giles Best (2016): Modeling the chronic lymphocytic leukemia microenvironment in vitro, Leukemia & Lymphoma, DOI: 10.1080/10428194.2016.1204654 To link to this article: http://dx.doi.org/10.1080/10428194.2016.1204654

1. individual pathways 
One strategy is to supplement CLL cell mono-cultures with spe-
cific signalling molecules known to be present in the respective niches to analyse their impact
on cancer cells (Crassini et al., 2017). However, the addition of selected stimuli only probes the
influence of specific pathways and cannot recapitulate the complexity of the microenvironment.

2. entire soluble factor repetoire: 
tudy the influence of soluble components on CLL cells.
For this, conditioned medium is produced from cells from the microenvironment and transferred
to CLL mono-cultures. This model focuses on understanding the influence of the entire soluble
microenvironment, rather than on dissecting individual pathways in an isolated setting.

3. Cell-cell contacts
Models using conditioned media take soluble factors into account, yet, they neglect the direct
interactions through cell-cell contact, which are essential for leukemia cells (Bakker et al., 2016).
Therefore, models using co-cultures of CLL and microenvironmental cells are indispensable to
gain a more complete picture on the impact of non-malignant cells on leukemia cells. 

Use different cell tyepes: 
a) T cells One major interaction partner of CLL cells are T-cells. Such studies have
revealed that especially stimulation with CD40L and IL-21 contributes to T-cell induced prolif-
eration of CLL cells (Pascutti et al., 2013). Instead of the direct usage of T-cells, CD40L can
be expressed on the surface of fibroblasts to induce CD40-mediated signalling in CLL (Crassini
et al., 2017). Even though the CD40/CD40L system is commonly used to simulate the microen-
vironment, it again only relies on a single pathway to activate CLL cells.

b) Nurse-lie cells 
Nurse-like cells have been first described by Burger et al. (2000) and protect CLL cells
from spontaneous apoptosis. The cells spontaneously outgrow from cultures of peripheral blood
mononuclear cells (PBMCs) and are of monocytic origin (Ten Hacken and Burger, 2016). Their
protective effect on CLL cells has especially been attributed to the activation of BCR and NFκB
within CLL cells upon co-culturing with nurse-like cells (Burger et al., 2009)
cite: 
Burger, J. A., Quiroga, M. P., Hartmann, E., B ̈urkle, A., Wierda, W. G., Keating, M. J. and
Rosenwald, A. (2009). High-level expression of the T-cell chemokines CCL3 and CCL4 by
chronic lymphocytic leukemia B cells in nurselike cell cocultures and after BCR stimulation.
Blood 113, 3050–8.

d) mimic the bone marrow wtih BMSCs: 
To recapitulate the interactions of CLL with the bone marrow microenvironment, especially
the influence of BMSCs needs to be taken into account. The first approaches into this direction
have been taken by co-culturing primary mesenchymal stem cells (MSCs), isolated from bone
marrow aspirates of donors, with CLL cells. These experiments have shown that MSCs generally
protect CLL cells from spontaneous apoptosis ex-vivo through soluble factors and direct cell-cell
contact (Lagneaux et al., 1998; Panayiotidis et al., 1996; Plander et al., 2011). As primary
material is not readily available for all research groups and primary cells are in general more
challenging to handle, multiple bone marrow derived stroma cell lines have been established from
murine or human donors (Kurtova et al., 2009). Two of the most commonly used human bone
marrow stroma cell lines are HS-5 and NKTert (Kawano et al., 2003; Roecklein and Torok-Storb,
1995). Both cell lines have been shown to provide protection to CLL cells from spontaneous
apoptosis, however, the effect of NKTert cells was especially pronounced (Balakrishnan et al.,
2015; Cheng et al., 2014; Fiorcari et al., 2013; Guo et al., 2017; Kurtova et al., 2009; Vangapandu
et al., 2018; Zhang et al., 2012). The cell lines have been widely used in many research studies
and helped to gain valuable insights into how CLL and the bone marrow microenvironment
interact. However, investigators should always be aware that these cell lines only represent
artificial systems and that the obtained results should always be validated in other models

cell - cell contact is essential for leukemia cells: 
Bakker, E., Qattan, M., Mutti, L., Demonacos, C. and Krstic-Demonacos, M. (2016). The role
of microenvironment and immunity in drug response in leukemia. Biochimica et Biophysica
Acta 1863, 414–426

\@ref(fig:microenvironmentTissues) _Adapted from Davids et al 2012_

```{r microenvironmentTissues, fig.cap='(ref:microenvironmentTissues)',out.width = "60%",  echo = FALSE}

knitr::include_graphics("figures/microenvironmentTissues.eps")

```



### Microenvironmental pathways and cross-talk {#intro-ME-pathways}
Whats knwonc about each pathway: Modeling the chronic lymphocytic leukemia microenvironment in vitro
Kyle Crassini, Yandong Shen, Stephen Mulligan & O. Giles Best (2016): Modeling the chronic lymphocytic leukemia microenvironment in vitro, Leukemia & Lymphoma, DOI: 10.1080/10428194.2016.1204654 To link to this article: http://dx.doi.org/10.1080/10428194.2016.1204654

pathways esp IL4 and TLR    
The microenvironment in CLL constitutes cross talk via soluble factors and cell-cell contacts, in the blood, bone marrow and lymph nodes.
_BCr signalling is central_ 
_other pathways_
_the impact that they have_
The important role of these niches in disease
pathogenesis is demonstrated by the fact that CLL cells undergo spontaneous apoptosis ex-
vivo (Collins et al., 1989).

\@ref(fig:microenvironmentOverview) _Adapted from: Weistner et al 2015_

```{r microenvironmentCrosstalk, fig.cap='(ref:microenvironmentOverview)', out.width = "60%", echo = FALSE}

knitr::include_graphics("figures/microenvironmentCrosstalk.eps")

```



\@ref(fig:microenvironmentSignalling) Taken from Kipps et al 2017; "Migration of chronic lymphocytic leukaemia (CLL) cells into the lyphoid tissu is primarily mediated through CXC-chemokine receptor 4 (CXCR4) in response to CXC-chemokine ligand 12 (CXCL12), which is secreted mainly by nurse-like cells (NLCs) and mesenchymal-derived stromal cells. Migration of CLL cells into lymph nodes also occurs via CC-chemokine receptor 7 (CCR7) in response to CC-chemokine ligand 19 (CCL19) and CCL21, which are produced by the endothelial cells of high endothelial venules (HEVs). HEV endothelial cells also express hyaluronan, which can interact with CD44, to facilitate B cell signalling and might enhance the production of active matrix metalloproteinase 9 (MMP9). Once in tissues, several chemokines promote B cell survival, including CXCL12, B cell-activating factor (BAFF; also known as TNFSF13B) and a proliferation-inducing ligand (APRIL; also known as TNFSF13). In addition, CLL cell survival can be promoted through cognate interactions between CD31 and CD38, and the production by stromal cells of WNT factors, which can interact with ROR1, ROR2 and/or various Frizzled receptors. CLL cell contact with mesenchymal stromal cells can also be established through vascular cell adhesion protein 1 (VCAM1)– α4β 1 integrin interactions that contribute to CLL cell survival. In turn, CLL cells can secrete chemokines, such as CCL3 and CCL4, which can recruit T cells and NLC-precursor cells (monocytes) to the CLL microenvironment. Activated T cells can provide CLL cells with proliferative signals through CD40 ligand (CD40L)-CD40 interactions and the secretion of several cytokines, such as IL-2, IL-4 and IL-10. In turn, activated CLL cells secrete CCL12 and CCL22, which attract more T cells into the CLL microenvironment. In tissues, CLL cells can be exposed to environmental and/or self-antigens that might trigger B cell activation through interactions with the surface immunoglobulin; this could amplify the responsiveness of CLL cells to the signals and factors that are provided by the CLL microenvironment. BAFFR, BAFF receptor (also known as TNFRSF13C); BCMA, B cell maturation protein (also known as TNFRSF17); BCR, B cell receptor; TACI, transmembrane activator and CAML interactor (also known as TNFRSF13B). _Figure from @Kipps2017 ._ "

```{r microenvironmentSignalling, fig.cap='(ref:microenvironmentSignalling)', out.width = "60%", echo = FALSE}

knitr::include_graphics("figures/microenivironmental_signalling.eps")

```



## Therapies in CLL 
### Clinical regimes 
[@Kipps2017]
The treatment of patients with CLL can include chemotherapy, a combination of chemotherapy and immunotherapy, or drugs that target the signalling pathways that promote the growth and/or survival of CLL cells (for example, BCR signalling and BCL-2)128,129.

Chemotherapy—Chemotherapy has been the mainstay of therapy for the past 50 years. Purine analogues (most commonly fludarabine, but also pentostatin or cladribine) and alkylating agents (including chlorambucil, cyclophosphamide or bendamustine) are used in the treatment of CLL130–132. Chemotherapy-based regimens can cause myelosuppression, an increased risk of infections and, in a small subset of patients, post-therapy myelodysplasia or secondary cancers, such as acute myeloid leukaemia (see Secondary cancers).

Chemoimmunotherapy—Phase III clinical trials have validated the benefit of anti-CD20 monoclonal antibodies, such as rituximab, obinutuzumab or ofatumumab, in combination with chemotherapy for the treatment of patients with CLL. In one trial (the CLL8 trial of the German CLL Study Group), patients who received fludarabine and cyclophosphamide with rituximab had higher response rates and a longer median progression-free survival (PFS) than patients who were treated with fludarabine and cyclophosphamide133. In a separate study (the CLL11 trial), patients >65 years of age with medical comorbidities who were treated with chlorambucil and either obinutuzumab or rituximab had improved response rates and longer median PFS than patients who were treated with chlorambucil alone134. However, the median PFS was significantly longer for patients who received obinutuzumab (26.7 months) than in those who received rituximab (11.1 months). In a third phase III trial, median PFS significantly improved from 13.1 months for patients treated with chlorambucil to only 22.4 months for patients treated with chlorambucil and ofatumumab135. As a consequence of these three trials, the US FDA approved the use of rituximab, obinutuzumab or ofatumumab in combination with chemotherapy for the first-line treatment of patients with CLL. The FDA also approved the use of ofatumumab as a single agent for the treatment of patients with relapsed or refractory disease based on data from a phase II study136.


### New Targeted therapies
in particular Ibrutinib and idelalisib 

[@Kipps2017]
Three main classes of drugs that each can inhibit BCR signalling have been evaluated in patients with CLL: BTK inhibitors, PI3K inhibitors and spleen tyrosine kinase (SYK) inhibitors86,143. CLL cells with unmutated IGHV seem to be more sensitive to inhibitors of BCR signalling than CLL cells with mutated IGHV144, but whether inhibitors, such as ibrutinib, are more effective in patients with CLL and unmutated IGHV, remains to be validated in clinical trials.
Ibrutinib has been approved in the United States and Europe for use as initial therapy, as well as in patients with relapsed disease, which followed results from a randomized trial that showed a significantly higher response rate to therapy with ibrutinib than with ofatumumab145. In addition, with continuous therapy, patients treated with ibrutinib had a significantly longer median PFS and overall survival than patients treated for 8 months with ofatumumab. Approval of ibrutinib as initial therapy was based on the results of a randomized trial that showed a significant improvement in median PFS and overall survival in patients 65 years of age without del(17p) who were treated indefinitely with ibrutinib than in patients treated for up to 48 weeks with chlorambucil146.
Upon initiation of treatment with ibrutinib, lymphadenopathy is rapidly reduced, which is associated with a concomitant increase in absolute lymphocyte count147. The rise in absolute lymphocyte count is related to the inhibition of chemokine receptor signalling, which inhibits the migration of CLL cells from the blood into the lymphoid tissues. This resulting lymphocytosis should not be considered a sign of progression; over time, the lymphocytosis subsides as the overall tumour burden decreases with continued therapy.

[@Kipps2017]
PI3K inhibitors include idelalisib, duvelisib (also known as IPI-145), TGR-1022 and ACP-319 (also known as AMG-319)151; the latter three drugs are being evaluated in clinical trials, whereas idelalisib was approved in the United States and Europe for the treatment of patients with relapsed CLL

[@Kipps2017]
Phase I/II clinical trials of fostamatinib, an oral SYK inhibitor, caused reduction in lymphadenopathy with concomitant lymphocytosis, an improvement in disease-related cytopenias and relief of disease-related symptoms in most of the treated patients with CLL156. However, dose-limiting toxicities of fostamatinib treatment include neutropenia,thrombocytopenia and diarrhoea. Other inhibitors of SYK, such as entospletinib, are being evaluated in preclinical and clinical studies.

[@Kipps2017]
BCL-2 inhibitors—Venetoclax is a small molecule that functions as a BH3 mimetic to inhibit BCL-2 (REF. 157). This drug is highly potent in inducing apoptosis in CLL cells, possibly by diminishing the capacity of BCL-2 to sequester the pro-apoptotic molecule BCL-2-interacting mediator of cell death (BIM; also known as BCL2L11)158. As such, venetoclax is effective in patients with relapsed and/or refractory disease159 or in patients with relapsed disease and del(17p)160. Indeed, the overall response rate for patients with relapsed disease and del(17p) was 79%, with 8% achieving a complete response. In addition, the estimated 12-month PFS was 72% and overall survival was 87%. On the basis of these results, the FDA approved the use of venetoclax for patients with relapsed disease and del(17p). Ongoing studies have shown that venetoclax can be safely combined with rituximab or obinutuzumab. Moreover, studies are examining the use of venetoclax with or without an anti-CD20 monoclonal antibody, and with or without ibrutinib161,162, which might provide higher response rates to therapy than that with venetoclax alone.

Idelalisib: 
Brown, J. R., Byrd, J. C., Coutre, S. E., Benson, D. M., Flinn, I. W., Wagner-Johnston, N. D.,
Spurgeon, S. E., Kahl, B. S., Bello, C., Webb, H. K., Johnson, D. M., Peterman, S., Li, D.,
Jahn, T. M., Lannutti, B. J., Ulrich, R. G., Yu, A. S., Miller, L. L. and Furman, R. R. (2014).
Idelalisib, an inhibitor of phosphatidylinositol 3-kinase p110δ, for relapsed/refractory chronic
lymphocytic leukemia. Blood 123, 3390–7

Ibrutinib: Byrd, J. C., Furman, R. R., Coutre, S. E., Flinn, I. W., Burger, J. A., Blum, K. A., Grant,
B., Sharman, J. P., Coleman, M., Wierda, W. G., Jones, J. A., Zhao, W., Heerema, N. A.,
Johnson, A. J., Sukbuntherng, J., Chang, B. Y., Clow, F., Hedrick, E., Buggy, J. J., James,
D. F. and O’Brien, S. (2013). Targeting BTK with ibrutinib in relapsed chronic lymphocytic
leukemia. The New England Journal of Medicine 369, 32–42.


### Drug resistances in CLL
incl ME -mediated and ibrutinib - IL4 

### BET inhibitors, and their potential for the treatment of leukemia

## Background to the approache used in this thesis
### Ex-vivo drug sensitivity screens

### Studies of the microenivornment, and attempts to replicate the impact of the microenvironment 
cancer biology is integratively influenced by cell-intrinsic and cell-extrinsic features<- evidence, and jsutify why we need to knwo more about this 
### ATAC sequncing as a proxy for TF activity justification {#intro-ATACseq}

### Measurement of outcomes? 
LDT, TTT, TFT, OS 

[@Kipps2017]
Currently, the most reliable prognostic models are those developed for treatment-free survival, as evolving treatments have yet to change the indications for therapy. Predictive models to define overall survival with a given type of therapy are challenged by the chronicity of CLL and the fact that patients often receive serial treatments, each of which can affect outcome; moreover, death might be due to an indirect or unrelated cause. Furthermore, treatment options are changing, with newly identified, highly effective agents that are clearly prolonging survival and have activity among patients who would have been considered high risk when the only option was conventional chemotherapy.

[@Kipps2017]
Historically, a favourable response to therapy has been defined as a partial remission or complete remission. Partial remission requires a 50% reduction in tumour bulk (for example, lymphadenopathy and splenomegaly), a 50% reduction in lymphocytosis, and platelet counts of >100,000 cells per µl (or 50% improvement over baseline) or a haemoglobin level of >11 g per dl (or 50% improvement over baseline) without requiring transfusions or exogenous growth factors107. Complete remission requires the normalization of blood counts, resolution in lymphadenopathy and splenomegaly, and normal marrow function. The use of CT to assess response in CLL is becoming more common, particularly in clinical trials. However, the benefit of using repeated CT scans to monitor disease is uncertain, and seems unlikely to change patient outcome. Because of the distinct pattern of response observed with BCR inhibitors, a new response category, namely, partial response with lymphocytosis, has been described. Partial response with lymphocytosis is defined as a >50% reduction in lymphadenopathy and splenomegaly, with persistent lymphocytosis; often the blood lymphocyte counts are equal to or greater than those observed prior to therapy.

In clinical trials, it is becoming more common to evaluate for MRD with 0.01% of CLL cells among the total population of mononuclear cells in the blood or marrow. MRD can be measured by flow cytometry or PCR with next-generation sequencing of the clonal immunoglobulin variable region gene rearrangements163. In most clinical trials for patients with CLL, particularly those conducted in Europe, evaluation of MRD has been performed by flow cytometry of mononuclear cells from the marrow aspirate (the preferred method) or from the peripheral blood. In the 6 months following anti-CD20 monoclonal antibody treatment, the assessment of MRD is more sensitive on the mononuclear cells of the marrow aspirate than on cells that are isolated from the blood, which will often lack detectable CLL even when they are readily found in the marrow. Beyond a complete response, the best predictor of long-term PFS and overall survival is the achievement of a complete remission without evidence for MRD.

### Multiomics datasets to study CLL 
The power of multi-omics approaches especially the use of ATAC sequencing and combining ATAC and RNA
 
### Mathematical modelling {#intro-multivariate-modelling}
Mathematical background in particular, lasso-regularised regression, and the caveats associated with feature selection
wrt cluster assignment 

Also use of bliss and other interactin methods? 

use method, explain how it works, use line plots: https://stattrek.com/multiple-regression/interaction.aspx
we need mor ltieracy about these things 
discuovery of mulitple layers of biology rather thatn jsut using t tests

how this can be used to build classfiers


caviats - screenign approach, wont find everything, rigid (as in drug - cyt -gene interactions), if two things are coerrelated, one could be srhuk to ze or both share weighting 
### CLL as a model for studying tumour biology

## Aims of this thesis

## Overview of the approach
Including diagrams 

## Note on contributions
Many of the figures, adn their cpations are taken directly from the amunscript. Where this si the case, it is indicated in the figure caption. 
