# Introduction

```{r loadLibraries, echo = FALSE}
library(ggplot2)
```

Brief: 
Referencing: 
* What do we already know about the subject (literature review)? Use citations:
  @JCIpaper shows that... Alternative Forms of the Wald test are
  considered [@JCIpaper].

* Provide an overview of your results.

* Outline of the paper:\
  _The thesis is organized as follows. The next section describes the model
  under investigation. Section "[Data]" describes the data set
  and Section "[Results]" presents the results. Finally, Section
  "[Conclusion]" concludes._
  
* The introduction should  be around  8 pages.

The microenvironment in CLL constitutes cross talk via soluble factors and cell-cell contacts, in the blood, bone marrow and lymph nodes. \@ref(fig:microenvironmentOverview)
```{r microenvironmentCrosstalk, fig.cap= "Summary of microenvironmental in CLL", echo = FALSE}

#knitr::include_graphics("figures/microenvironmentCrosstalk.eps")

```

```{r microenvironmentTissues, fig.cap= "The tissues that constitute the microenvironment in CLL", echo = FALSE}

#knitr::include_graphics("figures/microenvironmentTissues.eps")

```


