# Introduction
Next tasks:
Give a bit of time each day for next few weeks, then full tie when chapter 5 and 6 finished
Continue to go through add the next papers
Then go throuhg each section, do a lit review, and read what i need for each, add the text and references and kkep going 
Should then be easy to tie this together 
Herbst Thesis 

2 days reading each 2 days to write 
= 4 days 

= 16 days 

~3 weeks 



```{r loadLibraries, echo = FALSE}
library(ggplot2)
```

(theoretical background)
## CLL Disease biology 

### Disease Characteristcs
CD5, CD19, CD23+, lowlevelsofCD20 

Reading: 

From Kipps 2017: 
Incidence (get for 2021)
Siegel R, et al. Cancer treatment and survivorship statistics, 2012. CA Cancer J. Clin. 2012; 62:220–241. [PubMed: 22700443]


\@ref(fig:CLLmorphology) Taken from Kipps et al 2017: "Wright–Giemsa-stained blood smears showing the typical chronic lymphocytic leukaemia (CLL) B lymphocyte (part a). _Figure from @Kipps2017._"

```{r CLLmorphology, fig.cap='(ref:CLLmorphology)', echo = FALSE}

knitr::include_graphics("figures/CLL_morphology.eps")

```
_Figure taken from Kipps et al 2017_ 

\@ref(fig:proliferationCentres) Taken from Kipps et al 2017: "a | Tissue sections of a lymph node stained with haemotoxylin and eosin showing numerous pale-staining pseudofollicles, which are circled (original magnification ×20)." _Figure from @Kipps2017_

```{r proliferationCentres, fig.cap='(ref:proliferationCentres)', echo = FALSE}

knitr::include_graphics("figures/proliferation_centres.eps")

```
_Figure taken from Kipps et al 2017_ 

[@Kipps2017] 10.1038/nrdp.2016.96.Chronic
"Chronic lymphocytic leukaemia (CLL) is a malignancy of CD5+ B cells that is characterized by the accumulation of small, mature-appearing lymphocytes in the blood, marrow and lymphoid tissues. Signalling via surface immunoglobulin, which constitutes the major part of the B cell receptor, and several genetic alterations play a part in CLL pathogenesis, in addition to interactions between CLL cells and other cell types, such as stromal cells, T cells and nurse-like cells in the lymph nodes. The clinical progression of CLL is heterogeneous and ranges from patients whorequire treatment soon after diagnosis to others who do not require therapy for many years, if at all. Several factors, including the immunoglobulin heavy-chain variable region gene (IGHV) mutational status, genomic changes, patient age and the presence of comorbidities, should be considered when defining the optimal management strategies, which include chemotherapy, chemoimmunotherapy and/or drugs targeting B cell receptor signalling or inhibitors of apoptosis, such as BCL-2. Research on the biology of CLL has profoundly enhanced our ability to identify patients who are at higher risk for disease progression and our capacity to treat patients with drugs that selectively target distinctive phenotypic or physiological features of CLL. "

Get for 2021:
CLL is estimated to account for ~19,000 of all newly detected cancers in the United States in 2016 (REF. 11). The average incidence of CLL varies between individuals in different geographical regions and ranges from <0.01% of individuals in eastern Asia to ~0.06% of individuals in Europe and the United States. The risk of developing CLL is about two-times higher for men than for women and increases with age; the median age at diagnosis ranges from 70 to 72 years11–14.
The

### IGHV status and the role of BCR signalling


Reading: 
REview: 
B cell receptor signaling in chronic lymphocytic leukemia
Author links open overlay panelJan A.Burger1
NicholasChiorazzi2
https://doi.org/10.1016/j.it.2013.07.002

There is increasing evidence that BCR signaling plays a relevant role in CLL pathogenesis [114, 115]. :

114. Burger JA, Chiorazzi N. B cell receptor signaling in chronic lymphocytic leukemia. Trends Immunol. 2013;34:592–601. [PMC free article] [PubMed] [Google Scholar]
115. Iacovelli S, Hug E, Bennardo S, Duehren-von Minden M, Gobessi S, Rinaldi A, Suljagic M, Bilbao D, Bolasco G, Eckl-Dorna J, Niederberger V, Autore F, Sica S, Laurenti L, Wang H, Cornall RJ, Clarke SH, Croce CM, Bertoni F, Jumaa H, Efremov DG. Two types of BCR interactions are positively selected during leukemia development in the Emu-TCL1 transgenic mouse model of CLL. Blood. 2015;125:1578–1588. 

BCR pathway: 
[@tenHacken20167]
The B-cell receptor is a multimeric complex composed by the antigen-specific surface immunoglobulin (sIg) and the Ig-α/Ig-β hetero-dimers (CD79A, CD79B) (Figure 2). Antigen binding to the sIg induces activation of upstream kinases, including SYK and the Src kinase LYN, which phosphorylate immunoreceptor tyrosine-based activation motifs (ITAMs) in the cytoplasmatic tails of CD79A and CD79B. This, in turn, activates the cytoskeleton, including hematopoietic cell-specific LYN substrate-1 (HS1) protein [112, 113] and the related F-actin polymerization, as well as other upstream kinases, including BTK and PI3Ks and downstream pathways, including phospholipase C gamma 2 (PLC-γ2), calcium signaling, PKC, NF-κB signaling, mitogen-activated protein kinases (MAPKs) and nuclear transcription. 

From Kipps 2017: 
IGHV status:
Hamblin TJ, Davis Z, Gardiner A, Oscier DG, Stevenson FK. Unmutated Ig V(H) genes are associated with a more aggressive form of chronic lymphocytic leukemia. Blood. 1999; 94:1848– 1854. [PubMed: 10477713]

Damle RN, et al. Ig V gene mutation status and CD38 expression as novel prognostic indicators in chronic lymphocytic leukemia. Blood. 1999; 94:1840–1847. [PubMed: 10477712] References 1 and 2 are landmark papers that describe two main subsets of patients with different disease progression tendencies based on IGHV mutation status of the immunoglobulins that are expressed by CLL cells.

Restricted repetoire: 
Kipps TJ, et al. Developmentally restricted immunoglobulin heavy chain variable region gene expressed at high frequency in chronic lymphocytic leukemia. Proc. Natl Acad. Sci. USA. 1989; 86:5913–5917. [PubMed: 2503826] This paper describes the discovery that the immunoglobulin repertoire of CLL cells may be highly restricted, suggesting that the antibodies expressed by CLL cells are most likely selected based on their capacity to bind to some common self-antigens.

Fais F, et al. Chronic lymphocytic leukemia B cells express restricted sets of mutated and unmutated antigen receptors. J. Clin. Invest. 1998; 102:1515–1525. [PubMed: 9788964]

Widhopf G, F 2nd, et al. Chronic lymphocytic leukemia B cells of more than 1% of patients express virtually identical immunoglobulins. Blood. 2004; 104:2499–2504. [PubMed: 15217828]

Targeting BCR : 
Targeting B cell receptor signalling in cancer: preclinical and clinical advances
Jan A. Burger & Adrian Wiestner 
Nature Reviews Cancer volume 18, pages 148–167 (2018) https://www.nature.com/articles/nrc.2017.121 


\@ref(fig:BCRsignalling) Adapted from: Kipps et al 2017: "B cell receptor (BCR) signalling is initiated by SRC-family kinase-dependent phosphorylation (mainly LYN) of CD79A and CD79B that creates a docking site for the binding and activation of spleen tyrosine kinase (SYK). SYK then triggers the formation of a multi-component ‘signalosome’, comprising Bruton tyrosine kinase (BTK), AKT, phosphoinositide 3-kinase (PI3K), phospholipase Cγ2 (PLCγ2) and B cell-linker protein (BLNK), among others. CD19 is a co-receptor for BCR and is important for PI3K activation, which recruits and activates PLCγ2, BTK and AKT. PLCγ2 generates diacylglycerol (DAG) and inositol-1,4,5-trisphosphate (Ins(1,4,5)P3), which triggers Ca2+ release from the endoplasmic reticulum, leading to the activation of the MEK–extracellular signal-regulated kinase (ERK) and nuclear factor-κB (NF-κB) signalling pathways. Other effects of BCR signalling include activation of mechanistic target of rapamycin complex 1 (mTORC1) and of Rho-family GTPases, RAC1 and RHOA, which can affect the cytoskeleton. Inhibitors of SYK, PI3K and BTK are shown. Note that this figure describes the main molecules and interactions that are involved in positive BCR signalling, but is not an exhaustive description of all signalling pathways or molecules activated. IKK, IκB kinase; PKC, protein kinase C."

```{r BCRsignalling, fig.cap='(ref:BCRsignalling)', out.width = "60%",  echo = FALSE}

knitr::include_graphics("figures/BCR_signalling.eps")

```

\@ref(fig:IGHVstatus) Taken from Kipps et al 2017: "Normal naive B cells that have undergone successful V(D)J recombination and express functional B cell receptors that are capable of binding to antigen interact with CD4+ T cells and accessory cells, which aggregate to form follicles that become germinal centres. Germinal cells each have a dark zone, comprising rapidly dividing B cells, and a light zone, comprising B cells mixed with follicular dendritic cells (FDCs), macrophages and helper T cells (TH cells). The B cells enter the dark zone of the germinal centre where they experience rapid proliferation and somatic hypermutation (SHM) in the genes encoding the immunoglobulin variable regions of the heavy chain (IGHV) and the light chain (IGVL). As they pass through to the light zone, the B cells that express the fittest B cell receptors for binding antigen are selected and may undergo immunoglobulin class-switch recombination. Chronic lymphocytic leukaemia (CLL) cells that use unmutated IGHV apparently originate from CD5+ B cells prior to experiencing SHM, whereas CLL cells that use mutated IGHV most likely originate from CD5+ B cells that have passed through and differentiated in the germinal centre. Some CLL cells might be derived from B cells that also have undergone immunoglobulin class-switch recombination and express immunoglobulin isotypes other than IgM and IgD, for example, IgG or IgA. Another subset is one with CLL cells that express immunoglobulin with only modest somatic mutations, such as CLL cells that use IGHV3-21 with ~97% homology to the inherited IGHV3-21 gene and an immunoglobulin light chain encoded by an unmutated IGLV3-21; these cells might derive from a B cell that has had constrained SHM, possibly owing to a limited need for immunoglobulin somatic diveresification and selection. Dashed arrows indicate speculated pathways." _Figure from @Kipps2017_. 

```{r IGHVstatus, fig.cap='(ref:IGHVstatus)', out.width="60%", echo = FALSE}

knitr::include_graphics("figures/IGHV_status.eps")

```

[@Kipps2017]
The BCR is composed of a ligand-binding trans-membrane immunoglobulin molecule (either IgA, IgD, IgE, IgG or IgM) and the signalling Igα (also known as CD79A)–Igβ (also known as CD79B) heterodimer. CLL cells typically co-express IgD and IgM, although at low levels compared with normal B cells; less than a few per cent of CLL cases express class-switched isotypes, most commonly IgG. The CD79A and CD79B molecules contain immunoreceptor tyrosine-based activation motifs, which can be phosphorylated following the crosslinking of surface immunoglobulin, thereby triggering BCR signalling. A functional BCR is required for the survival of mature B cells76 and is maintained in most mature B cell malignancies, including CLL. In CLL, evidence suggests that the surface immunoglobulin of CLL B cells is engaged by autoantigen, which leads to constitutive BCR signalling in vivo77–79. The importance of this interaction is underscored by the clinical success of kinase inhibitors that block BCR signalling (see Management), although effects on other receptors might also have a role80.
Like most cancers, CLL is heterogeneous and the outcome of BCR signalling ranges from enhanced B cell activation to B cell anergy81,82. The main pathways that lead to cell survival and proliferation downstream of the BCR are shown in FIG. 3, along with drugs targeted against key signalling intermediates. BCR signalling that leads to anergy is less well defined, but seems to involve biased activation of inhibitory molecules with only partial activation of the pathways that are typically associated with B cell activation81. One important molecule that may be involved is the inositol lipid phosphatase SHIP1. SHIP1 is activated by the tyrosine-protein kinase LYN and may limit B cell activation by counteracting phosphoinositide 3-kinase (PI3K) activity at both chronically engaged receptors and distant non-ligated BCRs, rendering them insensitive to stimulation82,83.
Enhanced B cell activation is more commonly observed in CLL that expresses unmutated IGHV, whereas anergy predominates in most cases of CLL that express mutated IGHV84. Anergy is a state of cellular lethargy induced by chronic engagement of the surface antigen receptors in the absence of adequate T cell help. Although capable of reversing their phenotype, anergic cells are less likely to proliferate in response to BCR signalling than more activated cells, which might, in part, account for the observation that patients with CLL cells that express mutated IGHV generally have more indolent disease than patients with CLL cells with unmutated IGHV85. The fate of the cell (activation versus anergy) might be influenced by the CLL cell of origin (FIG. 1), as the cell types that can form CLL differ in their patterns of DNA methylation73, and are likely to respond differently to autoantigens. An unresolved question is whether anergy can be reversed in vivo, mirroring what occurs in vitro78.
The BCR also coordinates the activity of other cell surface receptors, including integrins, such as α4β1 integrin. BCR stimulation can result in increased adhesion of CLL cells to α4β1 integrin substrates, for example, fibronectin and vascular cell adhesion protein 1 (REF. 86). By contrast, CXC-chemokine receptor 4 (CXCR4) is downmodulated by BCR engagement and both can trigger ‘inside-out’ signalling, resulting in the activation of α4β1 integrin87,88. Thus, recognized antigen encountered in lymphoid tissue is likely to affect adhesion and migration of CLL cells. Modulation of these pathways, coupled with the role of BTK and PI3K in chemokine receptor signalling89, contribute to the increased lymphocytosis observed in patients upon initiation of treatment with inhibitors of BTK or PI3K (see Management).


[@Kipps2017] 10.1038/nrdp.2016.96.Chronic
"CLL can be divided into two main subsets, which differ in their clinical behaviour. These subsets are distinguished by whether CLL cells express an unmutated or mutated immunoglobulin heavy-chain variable region gene (IGHV), reflecting the stage of normal B cell differentiation from which they originate1,2. CLL cells that express an unmutated IGHV originate from a B cell that has not undergone differentiation in germinal centres, which are the sites in the lymph nodes where B cells experience somatic hypermutation in their immunoglobulin variable region genes and selection during an immune response. Patients with CLL cells that express an unmutated IGHV typically have more-aggressive disease than patients with CLL cells that express a mutated IGHV. CLL cells with mutated IGHV arise from a post-germinal centre B cell that expresses immunoglobulin that has undergone somatic hypermutation and, in some cases, also immunoglobulin isotype switching (FIG. 1), similar to what occurs in normal B cells during an immune response to antigen. It should be emphasized that the high level of somatic mutations that arise in IGHV in the germinal centre are a natural part of affinity maturation of antibodies and, unlike mutations in other genes, are not pathological. The tumours are simply reflecting the stage of maturation of the parental B cell. In addition, some CLL cells have been described that are similar to unmutated IGHV CLL, but originate from B cells with limited somatic mutation, such as CLL with immunoglobulin heavy chains encoded by mutated IGHV3–21 and immunoglobulin light chains encoded by unmutated IGLV3–21 (REFS 3,4).
The repertoire of immunoglobin molecules produced by the CLL cells of all patients is considerably more limited than the repertoire of immunoglobulin molecules that can be made by the B cells of any one person5,6, reflecting the biased use in CLL of certain IGHV genes that have restricted somatic mutation and limited junctional and heavy-light chain combinatorial diversity. In as many as one-third of patients, the CLL cells express immunoglobulin ‘stereotypes’, which are stretches of primary structure in the variable region that can also be identified in the immunoglobulins produced by the CLL cells of other patients7. The restricted immunoglobulin repertoire in CLL is underscored by the finding that ~1 in 75 patients have CLL cells that express immunoglobulin molecules that are virtually identical8. The limited immunoglobulin diversity provides compelling evidence that CLL B cells are selected based on the binding activity of their expressed surface immunoglobulin, suggesting that B cell receptor (BCR) signalling plays a crucial part in CLL pathogenesis"


### Recurrent genetic features in CLL
Genetic alterations in CLL can include chromosomal alterations, mutations, alterations in the expression of mi RNAs and epigenetic modifications.

(for purpose of TLR sotry)
These included del(11q) and *ATM* ,
del(17p) and *TP53*, 
*SF3B1*. 


REading: 
From Kipps et al: 
38 Dohner H, et al. Genomic aberrations and survival in chronic lymphocytic leukemia. N. Engl. J. Med. 2000; 343:1910–1916. [PubMed: 11136261]

Del(13q)mechanism: 39. Klein U, et al. The DLEU2/miR-15a/16-1 cluster controls B cell proliferation and its deletion leads to chronic lymphocytic leukemia. Cancer Cell. 2010; 17:28–40. [PubMed: 20060366]

Del(17p) mechanism 40. Zenz T, Mertens D, Kuppers R, Dohner H, Stilgenbauer S. From pathogenesis to treatment of chronic lymphocytic leukaemia. Nat. Rev. Cancer. 2010; 10:37–50. [PubMed: 19956173]

del(11q)

Mutations and outcome: 
38: Dohner H, et al. Genomic aberrations and survival in chronic lymphocytic leukemia. N. Engl. J. Med. 2000; 343:1910–1916. [PubMed: 11136261]


JIC paper, explaon PACE 
Improved in recent years: 41: Van Dyke DL, et al. The Dohner fluorescence in situ hybridization prognostic classification of chronic lymphocytic leukaemia (CLL): the CLL Research Consortium experience. Br. J. Haematol. 2016; 173:105–113. [PubMed: 26848054]

CLl ahrbours a high degree of genetic diversity: 
Fabbri G, et al. Analysis of the chronic lymphocytic leukemia coding genome: role of NOTCH1 mutational activation. J. Exp. Med. 2011; 208:1389–1401. [PubMed: 21670202]
43. Pleasance ED, et al. A comprehensive catalogue of somatic mutations from a human cancer genome. Nature. 2010; 463:191–196. [PubMed: 20016485]
44. Puente XS, et al. Whole-genome sequencing identifies recurrent mutations in chronic lymphocytic leukaemia. Nature. 2011; 475:101–105. [PubMed: 21642962]
45. Wang L, et al. SF3B1 and other novel cancer genes in chronic lymphocytic leukemia. N. Engl. J. Med. 2011; 365:2497–2506. [PubMed: 22150006]

Functional role of mutations: 
WNT: 47 Wang L, et al. Somatic mutation as a mechanism of Wnt/beta-catenin pathway activation in CLL. Blood. 2014; 124:1089–1098. [PubMed: 24778153]

POT1: reference thereview

SF3B1: 
45 Wang L, et al. SF3B1 and other novel cancer genes in chronic lymphocytic leukemia. N. Engl. J. Med. 2011; 365:2497–2506. [PubMed: 22150006]
 48 Ferreira PG, et al. Transcriptome characterization by RNA sequencing identifies a major molecular and clinical subdivision in chronic lymphocytic leukemia. Genome Res. 2014; 24:212–226. [PubMed: 24265505]
 
 49. Quesada V, et al. Exome sequencing identifies recurrent mutations of the splicing factor SF3B1 gene in chronic lymphocytic leukemia. Nat. Genet. 2012; 44:47–52.
50. Te Raa GD, et al. The impact of SF3B1 mutations in CLL on the DNA-damage response. Leukemia. 2015; 29:1133–1142. [PubMed: 25371178]

Seqencing studies: 
two seminal studies: 
51. Landau DA, et al. Mutations driving CLL and their evolution in progression and relapse. Nature. 2015; 526:525–530. [PubMed: 26466571]
52. Puente XS, et al. Non-coding recurrent mutations in chronic lymphocytic leukaemia. Nature. 2015; 526:519–524. [PubMed: 26200345] References 51 and 52 describe landmark studies on whole-exome sequencing of CLL cells obtained from a large cohort of patients, reporting new driver mutations in CLL.


Epigenetic changes: 
Global hypomethylation and local hypermethylation
66. Cahill N, et al. 450K–array analysis of chronic lymphocytic leukemia cells reveals global DNA
methylation to be relatively stable over time and similar in resting and proliferative compartments. Leukemia. 2013; 27:150–158. [PubMed: 22922567]
67. Wahlfors J, et al. Genomic hypomethylation in human chronic lymphocytic leukemia. Blood. 1992; 80:2074–2080. [PubMed: 1382719]
68. Ziller MJ, et al. Charting a dynamic DNA methylation landscape of the human genome. Nature. 2013; 500:477–481. [PubMed: 23925113]

Methylation for classification: 
69. Kulis M, et al. Epigenomic analysis detects widespread gene-body DNA hypomethylation in chronic lymphocytic leukemia. Nat. Genet. 2012; 44:1236–1242. [PubMed: 23064414] This paper describes an analysis of DNA methylation in CLL cases, showing that the two molecular subtypes of CLL have differing DNA methylomes that seem to represent epigenetic imprints from distinct normal B cell subpopulations.

74. Bhoi S, et al. Prognostic impact of epigenetic classification in chronic lymphocytic leukemia: the
case of subset #2. Epigenetics. 2016; 11:449–455. [PubMed: 27128508]


\@ref(fig:CLLmutations) Taken from Kipps et al 2017: "Genes that are mutated in chronic lymphocytic leukaemia (CLL) are involved in several cellular pathways (blue boxes). As such, mutations in these genes can lead to a range of cellular consequences, such as aberrant DNA repair and B cell receptor (BCR) signalling, among others51,213. The minus sign from GBN1 to the MAPK–ERK pathway indicates negative regulation. *For more detail of the BCR and its associated signalling, see FIG. 3. ASXL1, additional sex combs-like protein 1; ATM, ataxia telangiectasia mutated; BAZ2A, bromodomain adjacent to zinc-finger domain protein 2A; BCOR, BCL-6 co-repressor; BIRC3, baculoviral IAP repeat-containing protein 3; BRCC3, BRCA1/BRCA2-containing complex subunit 3; C-NOTCH, carboxy-terminal domain of NOTCH; CARD11, caspase recruitment domain-containing protein 11; CHD2, chromodomainhelicase-DNA-binding protein 2; CHK2, checkpoint kinase 2; Co-A, co-activator; CSL, CBF1–Suppressor of Hairless–LAG1 (also known as RBPJ); DDX3X, ATP-dependent RNA helicase DDX3X; DYRK1A, dual-specificity tyrosine-phosphorylation-regulated kinase 1A; EGR2, early growth response 2; ELF4, ETS-related transcription factor Elf-4; ERK, extracellular signal-regulated kinase; EWSR1, Ewing sarcoma breakpoint region 1 protein; FBXW7, F-box/WD repeat-containing protein 7; FUBP1, far upstream element-binding protein 1; GNB1, guanine nucleotide-binding protein G(I)/G(S)/G(T) subunit β1; H3K4, histone H3 lysine 4; IC, intracellular domain; IKZF3, Ikaros family zinc-finger protein 3; IL-1R, IL-1 receptor; IRF4, interferon regulatory factor 4; ITPKB, inositol-trisphosphate 3-kinase B; LRP, low-density lipoprotein receptor-related protein; MAP2K1, dual-specificity mitogen-activated protein kinase kinase 1; MAPK, mitogen-activated protein kinase; MED12, Mediator of RNA polymerase II transcription subunit 12; MGA, MAX gene-associated protein; MYD88, myeloid differentiation primary response protein MyD88; NF-κB, nuclear factor-κB; NXF1, nuclear RNA export factor 1; P, phosphate; POT1, protection of telomeres protein 1; PTPN11, tyrosine-protein phosphatase non-receptor type 11; RIPK1, receptor-interacting serine/threonine-protein kinase 1; RPS15, 40S ribosomal protein S15; SAMHD1, SAM domain and HD domain-containing protein 1; SF3B1, splicing factor 3B subunit 1; SHP1, Src homology region 2 domain-containing phosphatase 1 (also known as PTPN6); SYK, spleen tyrosine kinase; TCF/LEF, T cell factor/lymphoid enhancer factor; TLR8, Toll-like receptor 8; TNFR1, tumour necrosis factor receptor 1 (also known as TNFRSF1A); TRAF, TNFR-associated factor; XPO, exportin; ZMYM3, zinc-finger MYM-type protein 3. Adapted with permission from REF. 51, Macmillan Publishers Limited." _Figure from @Kipps2017_

```{r CLLmutations, fig.cap='(ref:CLLmutations)', out.width="60%", echo = FALSE}

knitr::include_graphics("figures/genetic_mutations.eps")

```


*mutations*
[@Kipps2017] 
Several large genetic studies have revealed numerous genetic alterations in CLL, including single- nucleotide polymorphisms (SNPs), chromosomal alterations and alterations in non- coding RNA, such as microRNA (miRNA), some of which can be used to determine prognosis and to guide management strategies. Interactions between CLL cells and their microenvironment, including interactions with other cell types, such as T cells, nurse-like cells and stromal cells, can induce B cell proliferation and contribute to disease.

[@Kipps2017]
The altered expression of genes that are located in or near CLL-associated SNPs might contribute to disease development. For example, a SNP in IRF4 is associated with low expression of interferon regulatory factor 4; mice that are deficient in this protein can develop CLL23, partly owing to hyperactivation of Notch signalling24. 

23: Shukla V, Ma S, Hardy RR, Joshi SS, Lu R. A role for IRF4 in the development of CLL. Blood. 2013; 122:2848–2855. [PubMed: 23926303]
24. Shukla V, Shukla A, Joshi SS, Lu R. Interferon regulatory factor 4 attenuates notch signaling to suppress the development of chronic lymphocytic leukemia. Oncotarget. 2016; 7:41081–41094. [PubMed: 27232759]


[@Kipps2017]
Somatic mutations—The advent of massively parallel sequencing and the application of whole-exome sequencing to CLL have transformed our understanding of the genetic heterogeneity of CLL and have established that CLL harbours a high degree of genetic variability42–45 (FIG. 2). From these studies, recurrent somatic mutations have been consistently observed in genes that have a role in DNA damage (for example, TP53 and ATM), mRNA processing (for example, SF3B1 and XPO1), chromatin modification (for example, HIST1H1E, CHD2 and ZMYM3), WNT signalling, Notch signalling (for example, NOTCH1) and inflammatory pathways (for example, MYD88). Other mutations, such as those found in EGR2 or BRAF, can affect B cell-related signalling and transcription46 (FIG. 2).
The functional role of several putative driver mutations has been confirmed; for example, silencing mutated WNT pathway genes in primary CLL cells resulted in decreased cell viability47. Mutations in POT1, which has a role in the protection of telomeres, prevented the binding of protection of telomeres protein 1 to telomeric DNA, resulting in numerous chromosomal abnormalities, in addition to the development of abnormal telomeres. Mutations in SF3B1 have been found to be associated with aberrant RNA splicing45,48,49 and an altered DNA-damage response50. SAMHD1 encodes a protein involved in the regulation of intracellular deoxy-nucleotide pools, which are recruited to the site of DNA damage and are probably involved in the response to DNA double-strand breaks50.
The detection of somatic mutations and their relative frequencies is variable, which possibly reflects differences in the composition of the cohorts studied worldwide. Two seminal studies have provided the largest sequenced collections to date51,52, in which >500 CLL samples were characterized by whole-exome sequencing or whole- genome sequencing. The clinical and/or biological features of the patients examined in each study were notably distinct; one study analysed matched pretreatment samples from patients who required initial treatment and noted mutations in SF3B1 (21% of patients), ATM (15% of patients), TP53 (7% of patients), NOTCH1 (6% of patients) or BIRC3 (4% of patients). The other study assessed patients with earlystage CLL and patients with monoclonal B cell lymphocytosis and identified NOTCH1 (12.6% of patients), A T M (11% of patients), BIRC3 (8.8% of patients) and SF3B1 (8.6% of patients) as the most frequently mutated genes.
These large sample cohorts have provided the sensitivity to discover novel candidate cancer genes that are altered in CLL. Both studies also identified somatic mutations in MGA and PTPN11, which encode modulators of MYC, IKZF3, which encodes a key transcription factor, and RPS15, which encodes 40S ribosomal protein S15 and is recurrently mutated in ~20% of patients who relapse after combination chemotherapy53. Other recurrent somatic mutations include those in the 3′ untranslated region of NOTCH1 and a PAX5 enhancer, which increases the expression of these B cell-associated transcription factors54,55. Patients with mutations in the 3′ untranslated region of NOTCH1 have a shorter time from diagnosis to treatment and poorer overall survival, similar to that of patients with non- synonymous mutations, which alter the amino acid sequence of NOTCH1.
Next-generation sequencing has revealed intra-tumoural heterogeneity in CLL. Some somatic mutations, such as those in MYD88, or chromosomal abnormalities, such trisomy 12 or del(13q), are most often found in all the CLL cells of any one patient, indicating that these genetic alterations occurred early in the evolution of the leukaemia. Other mutations, such as those found in SF3B1 or NOTCH1, or chromosomal alterations, such as del(17p), are typically found in only a fraction of the leukaemia cells and thus represent subclonal events, which occur after the development of CLL. Across studies, subclonal driver mutations are associated with more-aggressive disease, particularly when two or more are found concurrent in a leukaemia cell population51,56,57. In addition, studies have demonstrated that large clonal shifts can occur following chemotherapy, owing to increases in the proportions of CLL cells that have a TP53 mutation or del(17p), indicating that such genetic changes provide a strong fitness advantage in the setting of therapy51. By contrast, one study of CLL cells from patients treated with ibrutinib (an inhibitor of Bruton tyrosine kinase (BTK)), revealed mutations associated with drug resistance that were distinct from those observed in CLL cells of patients treated with standard chemotherapy58. CLL was the first human disease that was found to be associated with alterations in miRNA.


(can compare size of large studies with ours)



*copy number variants*
[@Kipps2017]
Approximately 80% of patients with CLL carry at least one of four common chromosomal alterations: a deletion in chromosome 13q14.3 (del(13q)), del(11q), del(17p) and trisomy 12 (REF. 38). Del(13q) is the most common chromosomal alteration, evident in >50% of patients, and is associated with favourable prognosis. Within this deleted region is the DLEU2-mir-15–16 cluster, which regulates the expression of proteins that can inhibit apoptosis or that are involved in cell cycle progression39 (see Supplementary information S2 (table)). Del(17p) is found in 7% of patients and is associated with loss of the tumour suppressor gene TP53 (REF. 40), whereas del(11q) is found in 18% of patients and is often associated with alterations in ATM (which encodes a protein involved in DNA repair); each of these chromosomal alterations is associated with adverse clinical outcome38, although this has improved in recent years41. Trisomy 12 is found in 16% of patients with CLL and is associated with an intermediate prognosis. Unlike the neoplastic B cells in mantle cell lymphoma, CLL cells do not have the translocation t(11;14) (q13;q32) or other genetic alterations that enhance the expression of CCND1, which encodes the cell cycle regulator cyclin D1

*Epigenetic changes*
The CLL epigenome shows global hypomethylation combined with local hypermethylation, as has been observed in other cancers66–68. Indeed, comprehensive methylation profiling has demonstrated substantial intra-tumoural methylation heterogeneity52,69–73. Increasing methylation heterogeneity has consistently been associated with increased genetic complexity owing to the acquisition of subclonal mutations, thus linking genomic and methylomic evolution in CLL52,70,71. Indeed, locally disordered methylation in CLL might enhance the evolutionary adaptive capacity of CLL cells by increasing the background ‘noise’ of the genome, thereby providing increased opportunities for somatic mutations within the leukaemia clone. In support of this notion is the observed association between methylation evolution and adverse clinical outcome52,70,71.

[@Kipps2017]
Methylation signatures can classify distinct clinical CLL subgroups69,74. As these methylation patterns are a heritable trait, they have been used to ‘trace’ back to the type of normal B cell from which the CLL cells were derived75. These studies revealed that the CLL cells of different patients derive from a continuum of B cell maturation states, which are not restricted to discrete maturation stages. Nevertheless, CLL cells that use unmutated IGHV versus mutated IGHV generally have distinctive methylation patterns, which respectively approximate to those of pre-germinal centre versus post-germinal centre memory B cells, as depicted in FIG. 1. The diversity in the likely cells of origin of CLL cells highlights the biological and phenotypic heterogeneity of this disease. These findings also suggest that epigenetic programming that is dependent of transcription factors has a potentially important role in the development of CLL.


### The incompletely understood role of trisomy12 {#intro-trisomy12}
Bosch, F., Dalla-Favera, R. Chronic lymphocytic leukaemia: from genetics to treatment. Nat Rev Clin Oncol 16, 684–701 (2019). https://doi.org/10.1038/s41571-019-0239-8

Trisomy 12 chronic lymphocytic leukemia expresses a unique set of activated and targetable pathways
DOI: https://doi.org/10.3324/haematol.2018.190132
Issue: Vol. 103 No. 12 (2018): December, 2018

The Protein Landscape of Chronic Lymphocytic Leukemia (CLL)
Fabienne Meier-Abt https://doi.org/10.1182/blood.2020009741

what's understood about the overespssopn of genes and proliferative drive; 
DOI: 10.1200/JCO.2005.02.568 Journal of Clinical Oncology 23, no. 16 (June 01, 2005) 3780-3792. 

### Stratification of CLL patients
Reading: 
(from Kipps)

12. Nabhan C, et al. The impact of race, ethnicity, age and sex on clinical outcome in chronic lymphocytic leukemia: a comprehensive Surveillance, Epidemiology, and End Results analysis in the modern era. Leuk. Lymphoma. 2014; 55:2778–2784. [PubMed: 24588735]
13. Li Y, Wang Y, Wang Z, Yi D, Ma S. Racial differences in three major NHL subtypes: descriptive epidemiology. Cancer Epidemiol. 2015; 39:8–13. [PubMed: 25560974]
14. Pulte D, Redaniel MT, Bird J, Jeffreys M. Survival for patients with chronic leukemias in the US and Britain: age-related disparities and changes in the early 21st century. Eur. J. Haematol. 2015; 94:540–545. [PubMed: 25315799]

\@ref(fig:CLLstratification) Taken from Kipps et al 2017: "Indications for therapy of patients with chronic lymphocytic leukaemia (CLL) include late-stage disease, evidence for rapid disease progression or disease-related symptoms. Patients with del(17p) or mutated TP53 should be treated with therapy that does not require functional TP53, such as ibrutinib (a Bruton tyrosine kinase (BTK) inhibitor), given the relatively poor outcome for such patients with chemotherapy. For patients without del(17p) or known mutations in TP53, immunoglobulin heavy-chain variable region (IGHV) mutational status can help to define the treatment strategy; patients with unmutated IGHV could be considered for therapy with a BTK inhibitor (such as ibrutinib) and patients with mutated IGHV might be good candidates for chemoimmunotherapy (CIT), if amenable. Indeed, patients with mutated IGHV can have excellent outcomes with CIT regimens, such as fludarabine, cyclophosphamide and rituximab, with >50% of patients having a median progression-free survival of >10 years, including the potential for cure. If the patient is amenable to CIT, age, medical comorbidities and myeloid reserve should be taken into consideration. Patients >65 years of age commonly have medical comorbidities and are less able to tolerate myelosuppressive regimens, such as fludarabine, cyclophosphamide and rituximab. Thus, considerations should be given to using reduced dose or less myelosuppressive chemotherapy regimens, such as chlorambucil or reduced-dose bendamustine and an anti-CD20 monoclonal antibody for patients with limited myeloid reserve. Patients who either do not respond, have a poor tolerance to CIT or relapse following CIT, should be re-evaluated for del(17p) or TP53 mutations. Patients who develop de novo del(17p) or TP53 mutations, or have known del(17p) and/or TP53 mutations, or who develop resistance or intolerance to ibrutinib, could be considered for therapy with idelalisib and rituximab or the BCL-2 inhibitor venetoclax. Patients treated with CIT who do not have del(17p) or TP53 mutations could be considered for repeat CIT if their progression-free survival after CIT is >2 years and the patient has sufficient myeloid reserve. Such patients also might be treated with a BTK inhibitor or a phosphoinositide 3-kinase (PI3K) inhibitor, which also could be considered for patients who develop intolerance or resistance to therapy with ibrutinib. Patients who develop resistance or intolerance to inhibitors of BTK, PI3K and/or BCL-2 should be considered for clinical trials or alternative agents. LDT, lymphocyte doubling time." _Figure from @Kipps2017 _.

```{r CLLstratification, fig.cap='(ref:CLLstratification)', out.width = '60%', echo = FALSE}

knitr::include_graphics("figures/disease_stratification.eps")

```


Disease staging
[@Kipps2017] Two clinical staging systems are widely used to divide patients with CLL into three broad prognostic groups114,115. The Rai staging system (TABLE 1) is more commonly used in the United States, whereas the Binet classification (TABLE 2) is more commonly used in Europe. The staging systems each recognize the importance of marrow function and define late-stage, or high-risk, disease by the presence of pronounced anaemia or thrombocytopenia.

[@Kipps2017]
Prognostic factors that can help to identify patients who require therapy relatively soon after diagnosis include certain clinical features and genetic, molecular and biochemical characteristics of the CLL cell. Multivariable models, prognostic indexes116–118 and nomograms119 have been developed to consolidate such prognostic factors so that they can more robustly predict clinical outcome. Commonly used parameters that are associated with poorer outcome are male sex,
65 years of age, poor performance status due to medical
comorbidities, certain CLL cell characteristics, such as the expression of unmutated IGHV1,2, ZAP70 (REFS 120,121), CD49d (also known as integrin α4)122 or CD38 (REF.2), the presence of del(17p)38 or del(11q)123, high serum levels of β2-microglobulin (>3.5 mg per l)124, complex karyotype (that is, the presence of three or more chromosomal aberrations observed on a karyotype test)125,126, or a high absolute lymphocyte count (>50,000 cells per µl) and/or late-stage disease at initial presentation. Del(17p) is often associated with inactivating mutations in TP53 and is a predictor of poor outcome to treatment with regimens that involve conventional chemotherapy127.

[@Kipps2017]
Generally, indications to initiate therapy include pronounced disease-related anaemia or thrombocytopenia (patients with Rai stage III or stage IV disease, or Binet stage C disease), symptomatic lymphadenopathy and/or symptoms that are associated with active disease, such as night sweats, fatigue, unintentional weight loss or fever without evidence of infection107. However, when basing a treatment decision on constitutional symptoms alone, the physician should consider other medical conditions, such as hypothyroidism, hyperthyroidism, hypoglycaemia, chronic inflammation, uncommon opportunistic infections or sleep disorders, including sleep apnoea.

For patients who need treatment, the presence of del(17p) or mutated TP53 are the most important features that are currently directing the choice of therapy (FIG. 8). Next, advanced age of >65 years, the presence of medical comorbidities and the objectives of treatment have substantial bearing on the choice of therapy. Increasingly, IGHV mutational status is considered as a parameter when determining the type of therapy; for example, chemotherapy-based regimens are reserved for patients with CLL and mutated IGHV. Conversely, the specific Rai or Binet stage of the patient who requires treatment does not necessarily influence the choice of therapy.
Systemic

### Prognosis and where the problems lie
*harsh treatments, especially when most are elderly*
*not a proper cure - combo therapies may cover all bases* 
*heterogenous - combo therapies may cover all bases* 
The clinical course of newly diagnosed CLL is extremely variable; some patients remain free of symptoms and are fully active for decades, whereas others rapidly become symptomatic or develop high-risk disease, which requires treatment soon after diagnosis and might result in death due to therapy-related and/or disease-related complications. However, most patients have a clinical course that is in between these two extremes.

[@Kipps2017]
Adverse effects of ibrutinib include fatigue, diarrhoea, bleeding, ecchymoses, rash, arthralgia, myalgia, increased blood pressure and atrial fibrillation. Clinical trials are currently evaluating second-generation BTK inhibitors (for example, acalabrutinib148, ONO/ GS-4059 (REF. 149) or BGB-3111) to determine whether any one of these drugs has a superior therapeutic index than that of ibrutinib150.


[@Kipps2017] Toxicities of venetoclax include gastrointestinal disturbances, neutropenia and tumour lysis syndrome159, which is characterized by hyperkalaemia, hyper-uricaemia and/or azotaemia. Tumour lysis syndrome results from the rapid destruction of cancer cells and the release of their cellular contents into the blood. Tumour lysis syndrome typically occurs when initiating venetoclax therapy or when dosing is increased. Thus, patients start venetoclax with a low daily dose, which is escalated each week over 5 weeks to mitigate the risk of developing tumour lysis syndrome. Even with this strategy, patients who are at high risk for tumour lysis syndrome because of bulky lymphadenopathy and/or lymphocytosis of >25,000 cells per µl must be hydrated and closely monitored during therapy initiation and during dose escalation.

[@Kipps2017]
The outlook for patients with CLL has improved substantially over the past several years. Through research on the immune biology and genetics of CLL, patients can be stratified into subgroups with distinctive clinical features, which has improved our capacity to assess prognosis or govern therapy. However, an understanding of the mechanisms that contribute to immune dysfunction or how it contributes to autoimmune disease, such as autoimmune haemolytic anaemia, therapy resistance or therapy-related complications is unknown. Whether tyrosine kinase inhibitors can affect clonal evolution, induce and/or select for drug resistance, or can achieve complete responses if used earlier in the course of the disease is also unknown.

New therapies:   
[@Kipps2017]
Several therapies are currently under preclinical and clinical investigation for the treatment of patients with CLL, including new drugs and treatment modalities that can modulate the immune system, and cell transplantation.

*Car-T Therapies*
[@Kipps2017]
CD19-targeted CAR T cells have yielded long-term PFS and relapse-free survival durations in patients with CLL; in 14 patients with relapsed or refractory CLL, four patients achieved a complete response and four patients achieved a partial response197. None of the complete responders had MRD and none relapsed (a median follow-up of 19 months). However, the efficacy of CAR T cell therapy in patients with CLL has been modest compared with that in patients with acute lymphoblastic leukaemia; this might be owing to qualitative defects in the T cells of patients with CLL, who are generally older than patients with acute lymphoblastic leukaemia and already have immune dysfunction that reflects disease-associated anergy (see BCR and B cell signalling). Ibrutinib might partially correct some of these defects198. Larger

*Allogenic stem cell transplantation*
[@Kipps2017]
Allogeneic stem cell transplantation is a potentially curative strategy for patients with relapsed or refractory CLL, including patients with high-risk features such as del(17p).Research efforts are ongoing to develop better-tolerated cell-based therapy with a similar curative potential that can be used without the immunosuppression and associated long-term morbidity and mortality of allogeneic stem cell transplantation.
Donor availability, advanced patient age, associated toxicities of myelosuppression, graft- versus-host disease and impaired resistance to infections limit the application of allogeneic stem cell transplantation in patients with CLL. In addition, the advent of BCR signalling inhibitors and BCL-2 inhibitors provide multiple treatment options that afford well- tolerated, long-term disease control, making allogeneic stem cell transplantation the least desirable option for most patients.

*Immune-modulatory drugs*
[@Kipps2017]
Immune-modulatory drugs, such as thalidomide and lenalidomide, are approved for the treatment of patients with multiple myeloma, mantle cell lymphoma or myelodysplastic disease. Although these drugs have clinical activity in patients with CLL, they have had limited application unless used in combination with an anti-CD20 monoclonal antibody166,186. In CLL, lenalidomide can induce the expression of p21WAF1/CIP1, which inhibits cyclin-dependent kinase and CLL cell proliferation187, and can improve immune synapse formation, potentially enhancing immune function.

*Immune Checkpoint inhibitors*
[@Kipps2017]
The immune checkpoint receptor PD-1, and its ligands PD-L1 and PD-L2, is the most important cognate receptor involved in the suppression of cellular immune activation. CLL cells express high levels of PD-L1 and PD-L2 and can suppress the responses of PD-1- expressing effector T cells106, leading to an exhausted (that is, no longer functional) T cell phenotype. Preclinical studies in mouse models have demonstrated that checkpoint inhibitors can reactivate immune effector cells to have anti-leukaemia activity202. However,

*Combination targeted therapies*
[@Kipps2017]
We can now target the distinctive phenotypic or physiological features of CLL with targeted therapeutic agents, which have a higher therapeutic index than standard chemotherapy. Through the use of combination therapy, which targets different B cell survival signalling pathways and/or achieves better eradication of CLL cells, we might be able to define curative treatments for most patients with this disease.
Research on leukaemia cell survival signalling pathways, such as those governed by interactions between leukaemia cells and cells or secreted factors within the microenvironment (FIG. 4), might identify pathways that are not affected by BCR inhibitors. For example, BCR inhibitors, such as ibrutinib, cannot block ROR1-dependent WNT5A signalling, which enhances CLL cell proliferation, migration and survival93; as such, antibodies that block ROR1-dependent signalling could potentially have synergistic activity when used in combination with BCR inhibitors203. Furthermore, the interaction between CLL cells and accessory cells in the microenvironment might enhance CLL cell expression of anti-apoptotic proteins other than BCL-2, such as MCL1, thereby contributing to therapy resistance. As such, the therapeutic use of a selective BCL-2 antagonist, such as venetoclax, might be more effective when used in combination with BCL-2 inhibitors161,162, which also interfere with the homing of CLL cells to the microenvironment. Conceivably, combination target therapy with agents that have synergistic activity will provide highly effective and potentially curative treatment of patients with CLL.

## The tumour microenvironment in CLL 
Plan: read the burger review [x] 
Read the thesis [x] 
Copy in the important text and papers [x] 
Do a final scan on google scholar = comprehensive 
Read the important bits, copy the text
Copy what's important into dicsussion / other parts ot the text 


^today 
Write

Repeat for each other section 

Also:
include background on each pathway? or jsut a table? 

Reviews: 
Newest revieww: 
https://jhoonline.biomedcentral.com/articles/10.1186/s13045-021-01134-x 

Best review: Microenvironment interactions and B-cell receptor signaling in Chronic Lymphocytic Leukemia: Implications for disease pathogenesis and treatment☆
lElisa ten HackenJan A.Burge 2016

https://www.nature.com/articles/nrc3774
The tumour microenvironment in B cell lymphomas
David W. Scott & Randy D. Gascoyne

Other reviews: 
The microenvironment in mature B-cell malignancies: a target for new treatment strategies  
Jan A. Burger,Blood (2009) 114 (16): 3367–3375.
https://doi.org/10.1182/blood-2009-06-225326

The microenvironment in chronic lymphocytic leukemia (CLL) and other B cell malignancies: Insight into disease biology and new targeted therapies 2014
Jan A.Burger
John G.Gribben
https://doi.org/10.1016/j.semcancer.2013.08.011

Nurture versus Nature: The Microenvironment in Chronic Lymphocytic Leukemia
Jan A. Burger 2011


Structure:
The ME is this
It is important because
It consists of these cell types

these interact by the following:
cell-cell contacts
souble factors

these can be modelled

these are importance in drug response

this has also been modelling 


### The concept of the protective niche 
_definition of the ME_
get from Ten Hacken 

_evidence of the importance of the ME_

Laura Llaó Cid  PhD Thesis: 
Normal  B-cells  are  programmed  to  interact  with  their  microenvironment  during  the 
processes of differentiation, antigen presentation, maturation, and antibody secretion, in order 
to  execute  a  well-coordinated  and  tightly  regulated  immune  response  against  infection 
(Herishanu et al. 2013)
Herishanu Y, Katz BZ, Lipsky A, Wiestner A. 2013. Biology of chronic lymphocytic leukemia in 
different microenvironments: clinical and therapeutic implications. Hematol Oncol Clin 
North Am 27: 173-206. 
Herishanu Y, Perez-Galan P, Liu D, Biancotto A, Pittaluga S, Vire B, Gibellini F, Njuguna N, 
Lee  E,  Stennett  L  et  al.  2011.  The  lymph  node  microenvironment  promotes  B-cell 
receptor  signaling,  NF-kappaB  activation,  and  tumor  proliferation  in  chronic 
lymphocytic leukemia. Blood 117: 563-574. 


It is thus unsurprising that CLL cells retain the ability to communicate 
with their microenvironment, even though the normal compartmentalization and regulation of 
the immune response is altered. In fact, CLL cells are highly reliant on the interactions with the 
microenvironment for their survival and proliferation, especially in BM and secondary lymphoid 
structures.  Such  strong  dependency  is  evidenced  by  the  fact  that  in vitro  cultured  CLL  cells 
rapidly  undergo  apoptosis,  whereas  their  survival  is  significantly  extended  when  co-cultured 
with non-malignant cells such as  monocyte-derived nurse-like cells (NLCs) or mesenchymal 
bone marrow stromal cells (BMSCs) (Burger et al. 2000; Deaglio and Malavasi 2009). proliferation of CLL cells in patients is restricted to the pseudofollicles of secondary lymphoid 
organs, where tight interactions with the surrounding cells that shape the transcriptomic profile 
of CLL cells occur (Messmer et al. 2005; Herishanu et al. 2011). More specifically, CLL cells 
residing in the LNs overexpress genes related to BCR signaling as well as genes downstream 
of the NF-kB and NFAT pathways compared to PB and BM CLL cells (Herishanu et al. 2011)

Burger  JA,  Tsukada  N,  Burger  M,  Zvaifler  NJ,  Dell'Aquila  M,  Kipps  TJ.  2000.  Blood-derived 
nurse-like  cells  protect  chronic  lymphocytic  leukemia  B  cells  from  spontaneous 
apoptosis through stromal cell-derived factor-1. Blood 96: 2655-2663

Deaglio  S,  Malavasi  F.  2009.  Chronic  lymphocytic  leukemia  microenvironment:  shifting  the 
balance from apoptosis to proliferation. Haematologica 94: 752-

Herishanu Y, Perez-Galan P, Liu D, Biancotto A, Pittaluga S, Vire B, Gibellini F, Njuguna N, 
Lee  E,  Stennett  L  et  al.  2011.  The  lymph  node  microenvironment  promotes  B-cell 
receptor  signaling,  NF-kappaB  activation,  and  tumor  proliferation  in  chronic 
lymphocytic leukemia. Blood 117: 563-574. 

Messmer  BT,  Messmer  D,  Allen  SL,  Kolitz  JE,  Kudalkar  P,  Cesar  D,  Murphy  EJ,  Koduru  P, 
Ferrarini M, Zupo S et al. 2005. In vivo measurements document the dynamic cellular 
kinetics of chronic lymphocytic leukemia B cells. J Clin Invest 115: 755-764.

from Lemm phd thesis: 

The microenvironment has long been recognised to play an important role in the 
pathogenesis of CLL. This is first indicated by the fact that CLL cells readily accumulate in 
vivo however they rapidly undergo spontaneous apoptosis when cultured in vitro, implying 
that their apoptosis resistance in vivo relies on external survival factors rather than being 
an intrinsic factor of the CLL cell itself (Burger et al., 2000, Kurtova et al., 2009, Collins et 
al., 1989). These observations indicate that signals from the microenvironment are 
important for the prolonged survival of CLL cells in vivo. Secondly, many cells of the CLL 
microenvironment have been shown to confer a protective effect against 
chemotherapeutic agents. Thirdly, despite major improvements in the treatment, CLL still 
remains  an  incurable  disease  with  many  patients  relapsing  from  MRD  and  there  is 
growing  evidence  that  the  microenvironment  harbours  malignant  cells  acting  as  a  shield 
from the cytotoxic effects of chemotherapy, thus serving as a reservoir from which relapse 
may occur. Finally, the role of the microenvironment is further exemplified by the 
lymphocytosis  which  is  seen  in  patients  following  the  BCR  kinase  inhibitors  ibrutinib  and 
idelalisib  (Woyach  et  al.,  2014,  Brown  et  al.,  2014b).  This  striking  lymphocytosis  and 
effects of ibrutinib and idelalisib on migration and adhesion in vitro highlights the 
importance  of  the  microenvironment  in  CLL  cell  recirculation  through  the  PB  and  tissue 
microenvironment.  

Woyach  et  al.,  2014,  Brown  et  al.,  2014b


_description of the ME macro-scale_
[@tenHacken2016]
CLL cell interactions with the supportive tissue microenvironment play a critical role in 
disease pathogenesis [30]. CLL cells recirculate between peripheral blood and secondary 
lymphoid organs, where they proliferate in distinct tissue areas, termed “pseudofollicles”, at 
a daily birth rate of approximately 1–2% of the entire clone, as determined by deuterated 
water labeling. 

[30] Caligaris-Cappio F, Bertilaccio MT, Scielzo C. How the microenvironment wires the natural history of chronic lymphocytic leukemia. Seminars in cancer biology. 2014;24:43–48.

[31]Messmer BT, Messmer D, Allen SL, Kolitz JE, Kudalkar P, Cesar D, Murphy EJ, Koduru P, Ferrarini M, Zupo S, Cutrona G, Damle RN, Wasil T, Rai KR, Hellerstein MK, Chiorazzi N. In vivo measurements document the dynamic cellular kinetics of chronic lymphocytic leukemia B cells. The Journal of clinical investigation. 2005;115:755–764.

[@Kipps2017]
CLL cells depend on survival signals that they receive in lymphoid tissues from neighbouring non-neoplastic cells within the so-called cancer microenvironment. CLL cells follow chemokine gradients into lymph nodes, where they form ‘proliferation centres’ (REF. 77), as opposed to normal germinal centres. In these proliferation centres, the CLL cells contact non- malignant stromal cells, nurselike cells (also known as lymphoma-associated macrophages), T cells and mesenchymal-derived stromal cells (FIG. 4). Engagement with autoantigen may occur during this transit, thereby stimulating CLL cell activation and proliferation if sufficient T cell help is available. Only a few per cent of the CLL cells undergo proliferation at any one time; the remainder of the cells are either unstimulated or driven into anergy84. However, within such proliferation centres, all CLL cells are exposed to chemokines, integrins, cytokines and survival factors.

CLL cells circulating
in the blood are mostly found in a resting state, up to 3 % of cancer cells start to proliferate
in the lymph nodes (Herndon et al., 2017).

When they leave the lymph node: 
[@Kipps2017]
As CLL cells leave the tissue site, engagement with antigen will be transient and its effects are likely to reverse in the blood, leading to variable increases in the expression of surface IgM and CXCR4 (REF. 98). CLL cell expression of CXCR4 is downmodulated upon exposure to CXC-chemokine ligand 12 (CXCL12)99, which is produced by nurse-like cells in the microenvironment100. Consequently, CLL cells in the blood that have just exited lymphoid tissue express low levels of CXCR4 (known as CXCR4dim cells) and higher levels of CD5 (known as CD5bright cells) relative to the CLL cells that are poised to re-enter lymphoid compartments101. For unexplained reasons, a high level of expression of CXCR4 by circulating CLL cells is associated with poorer prognosis in patients with CLL that use mutated IGHV102, possibly by influencing tissue re-entry. In terms of treatment effects, kinase inhibitors, such as ibrutinib, inhibit BCR-associated pathways, which remain important for cancer cells that are retained in lymphoid tissues, but can also directly inhibit integrin-mediated and chemokine- mediated pathways, thereby contributing to the increased lymphocytosis that occurs following the initiation of kinase inhibitor therapy103.


_lymph nodes_
 The lymph node microenvironment promotes B-cell receptor signaling, NF-κB activation, and tumor proliferation in chronic lymphocytic leukemia
Yair Herishanu,
https://doi.org/10.1182/blood-2010-05-284984

Chronic Lymphocytic Leukemia Cells in a Lymph Node Microenvironment Depict Molecular Signature Associated with an Aggressive Disease
Amit K. Mittal, Nagendra K. Chaturvedi, Karan J. Rai, Christine E. Gilling-Cutucache, Tara M. Nordgren, Margaret Moragues, Runqing Lu, Rene Opavsky, Greg R. Bociek, Dennis D. Weisenburger, Javeed Iqbal & Shantaram S. Joshi 
Molecular Medicine volume 20, pages 290–301 

Chronic lymphocytic leukemia: the pathologist's view of lymph node microenvironment
Ponzoni MD
https://doi.org/10.1053/j.semdp.2011.02.014

Treatment of chronic lymphocytic leukemia requires targeting of the protective lymph node environment with novel therapeutic approaches
Rachel E. Hayden
2012
https://doi.org/10.3109/10428194.2011.610014 


CLL microenvironment: macro important
Jonathan W. Friedberg
Crossmark: Check for Updates
Blood (2011) 117 (2): 377–378.
https://doi.org/10.1182/blood-2010-11-314468

Direct in vivo evidence for increased proliferation of CLL cells in lymph nodes compared to bone marrow and peripheral blood
T M Herndon


_bone marrow_
the lymph node and the bone marrow Burger, J. A. and O’Brien, S. (2018). Evolution of CLL treatment — from chemoimmunotherapy to targeted and individualized therapy. Nature Reviews Clinical Oncology 15, 510–527 <- this isnt really the most direct 


The bone marrow microenvironment and leukemia: biology and therapeutic targeting
Edward Allan R Sison

_description on a micro scale (ie cell types)_
[@tenHacken2016]
Critical components of the microenvironment are monocyte-derived nurselike cells (NLCs), mesenchymal stromal cells, T cells and NK cells, which communicate with CLL cells through a complex network of adhesion molecules, chemokine receptors, tumor necrosis factor (TNF) family members, and soluble factors. (Auto-) antigens and/or autonomous mechanisms activate the B-cell receptor (BCR) and its downstream signaling cascade in secondary lymphatic tissues, playing a central pathogenetic role in CLL.

Co-culture of primary CLL cells with bone marrow mesenchymal cells, CD40 ligand and CpG ODN promotes proliferation of chemoresistant CLL cells phenotypically comparable to those proliferating in vivo
Noelia Purroy,

[@tenHacken2016]
Homing to tissues is dependent on a tightly regulated interaction 
between chemokines that are secreted by stromal cells within the tissues, which attract and 
retain CLL cells to tissues sites via corresponding chemokine receptors, in cooperation with 
adhesion molecules on the leukemia cells and respective tissue ligands. Over the years, 
several cellular components of the CLL microenvironment have been described, along with 
the signaling pathways involved in CLL homing, survival and proliferation, which now 
provides a rationale for targeting the CLL microenvironment.

each type: 
[@tenHacken2016]
1. Nurselike cells and mesenchymal stromal cells
NLCs represent a critical component of the CLL microenvironment (Figure 1 and Table 1). 
NLCs are cells of monocytic origin, which spontaneously differentiate in vitro from 
monocytes in high-density cultures of CLL peripheral blood mononuclear cells [32] and 
which can be found in situ in lymphoid organs from CLL patients [33, 34]. Gene expression 
profile analyses of CLL cells after CLL-NLC co-culture showed that NLCs activate the 
BCR and nuclear factor kappa B (NF-κB) signaling pathways in CLL cells [35]; similar 
gene signatures were identified in CLL cells isolated from lymph nodes of patients [36], 
demonstrating that NLCs are a valid model for studying the CLL microenvironment. NLCs 
induce chemotaxis and promote survival of CLL cells through secretion of chemokines C-X-
C motif ligand 12 (CXCL12) [32] and CXCL13 [34], and expression of TNF family 
members B-cell activating factor (BAFF) and a proliferation-inducing ligand (APRIL) [37], 
and they promote CLL disease progression in vivo in mouse models of CLL [38, 39]

[32.] Burger JA, Tsukada N, Burger M, Zvaifler NJ, Dell’Aquila M, Kipps TJ. Blood-derived nurse-like cells protect chronic lymphocytic leukemia B cells from spontaneous apoptosis through stromal cell-derived factor-1. Blood. 2000;96:2655–2663. [PubMed] [Google Scholar]

33. Tsukada N, Burger JA, Zvaifler NJ, Kipps TJ. Distinctive features of “nurselike” cells that differentiate in the context of chronic lymphocytic leukemia. Blood. 2002;99:1030–1037. [PubMed] [Google Scholar]

34. Burkle A, Niedermeier M, Schmitt-Graff A, Wierda WG, Keating MJ, Burger JA. Overexpression of the CXCR5 chemokine receptor, and its ligand, CXCL13 in B-cell chronic lymphocytic leukemia. Blood. 2007;110:3316–3325. [PubMed] [Google 
Scholar]
35. Burger JA, Quiroga MP, Hartmann E, Burkle A, Wierda WG, Keating MJ, Rosenwald A. High-level expression of the T-cell chemokines CCL3 and CCL4 by chronic lymphocytic leukemia B cells in nurselike cell cocultures and after BCR stimulation. Blood. 2009;113:3050–3058. [PMC free article] [PubMed] [Google Scholar]
36. Herishanu Y, Perez-Galan P, Liu D, Biancotto A, Pittaluga S, Vire B, Gibellini F, Njuguna N, Lee E, Stennett L, Raghavachari N, Liu P, McCoy JP, Raffeld M, Stetler-Stevenson M, Yuan C, Sherry R, Arthur DC, Maric I, White T, Marti GE, Munson P, Wilson WH, Wiestner A. The lymph node microenvironment promotes B-cell receptor signaling, NF-kappaB activation, and tumor proliferation in chronic lymphocytic leukemia. Blood. 2011;117:563–574.


From LLD phd thesis: 
Burger et al. first described a critical involvement of monocytes in supporting CLL with the 
observation  that  PB  monocytes  co-cultured  with  CLL  cells  differentiate  into  ‘nurse-like  cells’ 
that prolong leukemic cell survival (Burger et al. 2000). This in vitro model system has been 
extensively used to examine the mechanisms by which monocytes and macrophages support 
leukemia  cells.  For  example,  NLCs  attract  and  enhance  CLL  cell  survival  via  secretion  of 
chemokines CXCL12 and CXCL13, as well as through BAFF and APRIL expression (Burger 
et al. 2000; Nishio et al. 2005; Burkle et al. 2007)

Nurse like cells: chronic lymphocytic leukemia associated macrophages
Frédéric Boissard https://doi.org/10.3109/10428194.2014.991731 

2. Mesenchymal stromal cells
Mesenchymal stromal cells, such as bone marrow stromal cells (BMSCs), are “feeder” 
layers for normal hematopoietic progenitor cells and contribute to normal bone marrow 
architecture. Mesenchymal stromal cells are also commonly found in secondary lymphatic 
tissues of CLL patients [47], where they provide survival and migration signals to CLL cells 
(Figure 1 and Table 1). CLL cells are protected from spontaneous and drug-induced 
apoptosis by direct contact with BMSCs [48, 49], and they are able to co-opt and disrupt 
normal bone marrow architecture [50]. Stromal cells constitutively secrete chemokines, 
which organize CLL-cell trafficking and tissue homing [51], and provide additional signals 
that support CLL survival and promote drug resistance. BMSCs induce up-regulation of 
aggressive disease markers in CLL cells, including ZAP70 and CD38, as well as down-
regulation of C-X-C motif receptor 4 (CXCR4) [52]. BMSCs have also been recently shown 
to down-modulate CD20 expression from the surface of CLL cells [53], with implications 
for resistance to anti-CD20 antibody treatment. In addition, stromal cells promote 
glutathione synthesis in CLL cells [54], and induce glycolysis through NOTCH-mediated c-
MYC activation, thus promoting cell survival and drug resistance [55]

47. Ruan J, Hyjek E, Kermani P, Christos PJ, Hooper AT, Coleman M, Hempstead B, Leonard JP, Chadburn A, Rafii S. Magnitude of stromal hemangiogenesis correlates with histologic subtype of non-Hodgkin’s lymphoma. Clinical cancer research: an official journal of the American Association for Cancer Research. 2006;12:5622–5631. [PubMed] [Google Scholar]
48. Lagneaux L, Delforge A, Bron D, De Bruyn C, Stryckmans P. Chronic lymphocytic leukemic B cells but not normal B cells are rescued from apoptosis by contact with normal bone marrow stromal cells. Blood. 1998;91:2387–2396. [PubMed] [Google Scholar]
49. Kurtova AV, Balakrishnan K, Chen R, Ding W, Schnabl S, Quiroga MP, Sivina M, Wierda WG, Estrov Z, Keating MJ, Shehata M, Jager U, Gandhi V, Kay NE, Plunkett W, Burger JA. Diverse marrow stromal cells protect CLL cells from spontaneous and drug-induced apoptosis: development of a reliable and reproducible system to assess stromal cell adhesion-mediated drug resistance. Blood. 2009;114:4441–4450. [PMC free article] [PubMed] [Google Scholar]
50. Janel A, Dubois-Galopin F, Bourgne C, Berger J, Tarte K, Boiret-Dupre N, Boisgard S, Verrelle P, Dechelotte P, Tournilhac O, Berger MG. The chronic lymphocytic leukemia clone disrupts the bone marrow microenvironment. Stem cells and development. 2014;23:2972–2982. [PubMed] [Google Scholar]
51. Burger JA, Ghia P, Rosenwald A, Caligaris-Cappio F. The microenvironment in mature B-cell malignancies: a target for new treatment strategies. Blood. 2009;114:3367–3375. [PMC free article] [PubMed] [Google Scholar]
52. Purroy N, Abrisqueta P, Carabia J, Carpio C, Palacio C, Bosch F, Crespo M. Co-culture of primary CLL cells with bone marrow mesenchymal cells, CD40 ligand and CpG ODN promotes proliferation of chemoresistant CLL cells phenotypically comparable to those proliferating in vivo. Oncotarget. 2015;6:7632–7643. [PMC free article] [PubMed] [Google Scholar]
53. Marquez ME, Hernandez-Uzcategui O, Cornejo A, Vargas P, Da Costa O. Bone marrow stromal mesenchymal cells induce down regulation of CD20 expression on B-CLL: implications for rituximab resistance in CLL. British journal of haematology. 2015;169:211–218. [PubMed] [Google Scholar]
54. Zhang W, Trachootham D, Liu J, Chen G, Pelicano H, Garcia-Prieto C, Lu W, Burger JA, Croce CM, Plunkett W, Keating MJ, Huang P. Stromal control of cystine metabolism promotes cancer cell survival in chronic lymphocytic leukaemia. Nature cell biology. 2012;14:276–286. [PMC free article] [PubMed] [Google Scholar]
55. Jitschin R, Braun M, Qorraj M, Saul D, Le Blanc K, Zenz T, Mougiakakos D. Stromal cell-mediated glycolytic switch in CLL-cells involves Notch-c-Myc signaling. Blood. 2015 Epub ahead of print. [PubMed] [Google Scholar]

LLD Phd THesis: 
The majority of studies investigating the role of stromal cells in CLL have been performed 
using BMSCs. These cells provide attachment and growth factors in in vitro co-cultures with 
CLL  cells  via  the  CXCR4-CXCL12  axis,  and  protect  CLL  cells  from  spontaneous  and  drug-
induced apoptosis (Panayiotidis et al. 1996; Lagneaux et al. 1998; Burger et al. 1999; Kurtova 
et  al.  2009).  Other  cytokines,  chemokines,  proangiogenic  factors,  and  extracellular  matrix 
components secreted by mesenchymal cells promote CLL cell migration and survival. These 
include integrins b1 and b2, vascular cell adhesion molecule 1 (VCAM1), VEGF, and MMP9 
(Lagneaux et al. 1998; de la Fuente et al. 2002; Ringshausen et al. 2004; Redondo-Munoz et 
al. 2006; Gehrke et al. 2011). Importantly, the interaction between CLL cells and stromal cells 
results  in  the  activation  of  both  cell  types  and  induces  the  release  of  C-C  motif  chemokine 
ligand (CCL) 3 (CCL3) and CCL4 by CLL cells for the recruitment of other leukocytes like T-
cells (Burger et al. 2009; Ding et al. 2009).

Panayiotidis P, Jones D, Ganeshaguru K, Foroni L, Hoffbrand AV. 1996. Human bone marrow 
stromal  cells  prevent  apoptosis  and  support  the  survival  of  chronic  lymphocytic 
leukaemia cells in vitro. Br J Haematol 92: 97-103.

Burger JA, Burger M, Kipps TJ. 1999. Chronic lymphocytic leukemia B cells express functional 
CXCR4  chemokine  receptors  that  mediate  spontaneous  migration  beneath  bone 
marrow stromal cells. Blood 94: 3658-3667. 

Kurtova AV, Balakrishnan K, Chen R, Ding W, Schnabl S, Quiroga MP, Sivina M, Wierda WG, 
Estrov Z, Keating MJ et al. 2009. Diverse marrow stromal cells protect CLL cells from 
spontaneous and drug-induced apoptosis: development of a reliable and reproducible 
system  to  assess  stromal  cell  adhesion-mediated  drug  resistance.  Blood  114:  4441-
4450

de la Fuente MT, Casanova B, Moyano JV, Garcia-Gila M, Sanz L, Garcia-Marco J, Silva A, 
Garcia-Pardo A. 2002. Engagement of alpha4beta1 integrin by fibronectin induces in 
vitro resistance of B chronic lymphocytic leukemia cells to fludarabine. J Leukoc Biol 
71: 495-502.

Ringshausen  I,  Dechow  T,  Schneller  F,  Weick  K,  Oelsner  M,  Peschel  C,  Decker  T.  2004. 
Constitutive  activation  of  the  MAPkinase  p38  is  critical  for  MMP-9  production  and 
survival of B-CLL cells on bone marrow stromal cells. Leukemia 18: 1964-1970.

Redondo-Munoz J, Escobar-Diaz E, Samaniego R, Terol MJ, Garcia-Marco JA, Garcia-Pardo 
A. 2006. MMP-9 in B-cell chronic lymphocytic leukemia is up-regulated by alpha4beta1 
integrin  or  CXCR4  engagement  via  distinct  signaling  pathways,  localizes  to 
podosomes, and is involved in cell invasion and migration. Blood 108: 3143-3151.

Gehrke  I,  Gandhirajan  RK,  Poll-Wolbeck  SJ,  Hallek  M,  Kreuzer  KA.  2011.  Bone  marrow 
stromal  cell-derived  vascular  endothelial  growth  factor  (VEGF)  rather  than  chronic 
lymphocytic leukemia (CLL) cell-derived VEGF is essential for the apoptotic resistance 
of cultured CLL cells. Mol Med 17: 619-627.

c) Endothelial cells and follicular dendritic cells
Additional cellular elements in the CLL microenvironment include endothelial cells and 
follicular dendritic cells (FDCs) (Figure 1 and Table 1), which are essential for tissue 
homing and CLL retention to tissues. Adhesion to microvascular endothelial cells promotes 
CLL survival, activation and drug resistance [59–63]. CLL cells bind to β1 and β2 integrins 
[62] and to BAFF and APRIL on the surface of microvascular endothelial cells [60]. In 
addition, endothelin 1 (ET-1) engagement on CLL cells by endothelin subtype A receptor 
(ETAR) on endothelial cells promotes CLL survival and drug resistance, which can be 
blocked by ETAR inhibition [63]. In vitro culture with FDCs rescues CLL cells from 
spontaneous apoptosis by direct cell contact, dependent on ligation of CD44 on CLL cells 
and subsequent up-regulation of myeloid cell leukemia 1 (MCL1), a member of the BCL2 
family of anti-apoptotic proteins [64]. 

59. Badoux X, Bueso-Ramos C, Harris D, Li P, Liu Z, Burger J, O’Brien S, Ferrajoli A, Keating MJ, Estrov Z. Cross-talk between chronic lymphocytic leukemia cells and bone marrow endothelial cells: role of signal transducer and activator of transcription 3. Human pathology. 2011;42:1989–2000. [PMC free article] [PubMed] [Google Scholar]
60. Cols M, Barra CM, He B, Puga I, Xu W, Chiu A, Tam W, Knowles DM, Dillon SR, Leonard JP, Furman RR, Chen K, Cerutti A. Stromal endothelial cells establish a bidirectional crosstalk with chronic lymphocytic leukemia cells through the TNF-related factors BAFF, APRIL, and CD40L. Journal of immunology. 2012;188:6071–6083. [PMC free article] [PubMed] [Google Scholar]
61. Hamilton E, Pearce L, Morgan L, Robinson S, Ware V, Brennan P, Thomas NS, Yallop D, Devereux S, Fegan C, Buggins AG, Pepper C. Mimicking the tumour microenvironment: three different co-culture systems induce a similar phenotype but distinct proliferative signals in primary chronic lymphocytic leukaemia cells. British journal of haematology. 2012;158:589–599. [PubMed] [Google Scholar]
62. Maffei R, Fiorcari S, Bulgarelli J, Martinelli S, Castelli I, Deaglio S, Debbia G, Fontana M, Coluccio V, Bonacorsi G, Zucchini P, Narni F, Torelli G, Luppi M, Marasca R. Physical contact with endothelial cells through beta1- and beta2- integrins rescues chronic lymphocytic leukemia cells from spontaneous and drug-induced apoptosis and induces a peculiar gene expression profile in leukemic cells. Haematologica. 2012;97:952–960. [PMC free article] [PubMed] [Google Scholar]
63. Maffei R, Bulgarelli J, Fiorcari S, Martinelli S, Castelli I, Valenti V, Rossi D, Bonacorsi G, Zucchini P, Potenza L, Vallisa D, Gattei V, Poeta GD, Forconi F, Gaidano G, Narni F, Luppi M, Marasca R. Endothelin-1 Promotes Survival and Chemoresistance in Chronic Lymphocytic Leukemia B Cells through ETA Receptor. PloS one. 2014;9:e98818. [PMC free article] [PubMed] [Google Scholar]
64. Pedersen IM, Kitada S, Leoni LM, Zapata JM, Karras JG, Tsukada N, Kipps TJ, Choi YS, Bennett F, Reed JC. Protection of CLL B cells by a follicular dendritic cell line is dependent on induction of Mcl-1. Blood. 2002;100:1795–1801. [PubMed] [Google Scholar]
65. Granziero L, Circosta P, Scielzo C, Frisaldi E, Stella S, Geuna M, Giordano S, Ghia P, Caligaris-Cappio F. CD100/Plexin-B1 interactions sustain proliferation and survival of normal and leukemic CD5+ B lymphocytes. Blood. 2003;101:1962–1969. 


D) NK CELLS AND T CELLS 
Interactions between CD40-expressing B cells and CD40 ligand (CD40L) on activated 
CD4+ T cells are critical in the context of antigen presentation and induction of normal B-
cell responses [67]. Similarly, activation of malignant B cells by CD40 ligation promotes 
survival of CLL cells [68]
[68] 68. Kitada S, Zapata JM, Andreeff M, Reed JC. Bryostatin and CD40-ligand enhance apoptosis resistance and induce expression of cell survival genes in B-cell chronic lymphocytic leukaemia. British journal of haematology. 1999;106:995–1004.

ver the years, several studies have also reported defective NK-cell 
function. Human leukocyte antigen G (HLA-G) molecule overexpression in the plasma of 
CLL patients [78] induces NK-cell apoptosis and impairs NK-cell mediated cytotoxicity. 
Reduced NK-cell cytotoxicity has been associated to low expression levels of the activating 
receptors natural killer cell p30-related protein (NKp30) [79, 80] and natural killer group 2 
member D (NKGD2) [81]; NK cells can also produce soluble BAFF, which interferes with 
NK-cell mediated CLL cell lysis after rituximab administration [82], and show reduced 
responses to the activating soluble BCL2-associated athanogene 6 (BAG6) ligand produced 
by CLL cells [79]. Taken together, these findings indicate that both the T and NK-cell 
compartments have overall reduced effector activities, which can explain the evasion of 
CLL cells from immune-mediated destruction.

78. Rizzo R, Audrito V, Vacca P, Rossi D, Brusa D, Stignani M, Bortolotti D, D’Arena G, Coscia M, Laurenti L, Forconi F, Gaidano G, Mingari MC, Moretta L, Malavasi F, Deaglio S. HLA-G is a component of the chronic lymphocytic leukemia escape repertoire to generate immune suppression: impact of the HLA-G 14 base pair (rs66554220) polymorphism. Haematologica. 2014;99:888–896. [PMC free article] [PubMed] [Google Scholar]
79. Reiners KS, Topolar D, Henke A, Simhadri VR, Kessler J, Sauer M, Bessler M, Hansen HP, Tawadros S, Herling M, Kronke M, Hallek M, Pogge von Strandmann E. Soluble ligands for NK cell receptors promote evasion of chronic lymphocytic leukemia cells from NK cell anti-tumor activity. Blood. 2013;121:3658–3665. [PMC free article] [PubMed] [Google Scholar]
80. Veuillen C, Aurran-Schleinitz T, Castellano R, Rey J, Mallet F, Orlanducci F, Pouyet L, Just-Landi S, Coso D, Ivanov V, Carcopino X, Bouabdallah R, Collette Y, Fauriat C, Olive D. Primary B-CLL resistance to NK cell cytotoxicity can be overcome in vitro and in vivo by priming NK cells and monoclonal antibody therapy. J Clin Immunol. 2012;32:632–646. [PubMed] [Google Scholar]
81. Huergo-Zapico L, Acebes-Huerta A, Gonzalez-Rodriguez AP, Contesti J, Gonzalez-Garcia E, Payer AR, Villa-Alvarez M, Fernandez-Guizan A, Lopez-Soto A, Gonzalez S. Expansion of NK cells and reduction of NKG2D expression in chronic lymphocytic leukemia. Correlation with progressive disease. PloS one. 2014;9:e108326. [PMC free article] [PubMed] [Google Scholar]
82. Wild J, Schmiedel BJ, Maurer A, Raab S, Prokop L, Stevanovic S, Dorfel D, Schneider P, Salih HR. Neutralization of (NK-cell-derived) B-cell activating factor by Belimumab restores sensitivity of chronic lymphoid leukemia cells to direct and Rituximab-induced NK lysis. Leukemia. 2015 Epub ahead of print. 



_cutting edge of whats known about how it operates in vivo_ 
NLC cells in vivo:[@tenHacken2019]
cNLCs an be found in situ in lymphoid organs from CLL patients [33, 34]. Gene expression 
profile analyses of CLL cells after CLL-NLC co-culture showed that NLCs activate the 
BCR and nuclear factor kappa B (NF-κB) signaling pathways in CLL cells [35]; similar 
gene signatures were identified in CLL cells isolated from lymph nodes of patients [36], 
demonstrating that NLCs are a valid model for studying the CLL microenvironment

and they promote CLL disease progression in vivo in mouse models of CLL [38, 39].
38. Troeger A, Johnson AJ, Wood J, Blum WG, Andritsos LA, Byrd JC, Williams DA. RhoH is critical for cell-microenvironment interactions in chronic lymphocytic leukemia in mice and humans. Blood. 2012;119:4708–4718. [PMC free article] [PubMed] [Google Scholar]
39. Reinart N, Nguyen PH, Boucas J, Rosen N, Kvasnicka HM, Heukamp L, Rudolph C, Ristovska V, Velmans T, Mueller C, Reiners KS, Pogge von Strandmann E, Krause G, Montesinos-Rongen M, Schlegelberger B, Herling M, Hallek M, Fingerle-Rowson G. Delayed development of chronic lymphocytic leukemia in the absence of macrophage migration inhibitory factor. Blood. 2013;121:812–821. [PubMed] [Google Scholar]


endothelial cells and FDCs:
[@tenHacken2016]:
Reciprocal cross-talk between CLL cells and FDCs via the 
ten Hacken and Burger Page 4
Biochim Biophys Acta. Author manuscript; available in PMC 2017 March 01.
Author Manuscript Author Manuscript Author Manuscript Author Manuscript
CXCR5-CXCL13 and the lymphotoxin beta receptor (LTβR)/lymphotoxin alpha beta 
(LTαβ) signaling pathways is essential for CLL positioning within lymphoid follicles and 
for leukemia progression in vivo in the EμTCL1 mouse model of CLL [66]

[66]66. Heinig K, Gatjen M, Grau M, Stache V, Anagnostopoulos I, Gerlach K, Niesner RA, Cseresnyes Z, Hauser AE, Lenz P, Hehlgans T, Brink R, Westermann J, Dorken B, Lipp M, Lenz G, Rehm A, Hopken UE. Access to follicular dendritic cells is a pivotal step in murine chronic lymphocytic leukemia B-cell activation and proliferation. Cancer discovery. 2014;4:1448–1465.

signals: 
High levels of CXCR4+ cells have been associated with higher risk of lymphoid organ infiltration and poorer disease outcome [84], as well as higher responsiveness to BCR stimulation [85].

84. Calissano C, Damle RN, Hayes G, Murphy EJ, Hellerstein MK, Moreno C, Sison C, Kaufman MS, Kolitz JE, Allen SL, Rai KR, Chiorazzi N. In vivo intraclonal and interclonal kinetic heterogeneity in B-cell chronic lymphocytic leukemia. Blood. 2009;114:4832–4842. [PMC free article] [PubMed] [Google Scholar]
85. Coelho V, Krysov S, Steele A, Sanchez Hidalgo M, Johnson PW, Chana PS, Packham G, Stevenson FK, Forconi F. Identification in CLL of circulating intraclonal subgroups with varying B-cell receptor expression and function. Blood. 2013;122:2664–2672. [PubMed] [Google Scholar]

the impact of ibrutinib on the cross talk: 
Disruption of in vivo Chronic Lymphocytic Leukemia Tumor–Microenvironment Interactions by Ibrutinib – Findings from an Investigator-Initiated Phase II Study
Carsten U. Niemann, Sarah E.M. Herman, Irina Maric, Julio Gomez-Rodriguez, Angelique Biancotto, Betty Y. Chang, Sabrina Martyr, Maryalice Stetler-Stevenson, Constance M. Yuan, Katherine R. Calvo, Raul C. Braylan, Janet Valdez, Yuh Shan Lee, Deanna H. Wong, Jade Jones, Clare Sun, Gerald E. Marti, Mohammed Z.H. Farooqui and Adrian Wiestner
DOI: 10.1158/1078-0432.CCR-15-1965 Published April 2016

evidence of CD14 activity in vivo
 Soluble CD14 is a novel monocyte-derived survival factor for chronic lymphocytic leukemia cells, which is induced by CLL cells in vitro and present at abnormally high levels in vivo
Martina Seiffert,
Angela Schulz,
Sibylle Ohl,
Hartmut Döhner,
Stephan Stilgenbauer,
Peter Lichter
Crossmark: Check for Updates
Blood (2010) 116 (20): 4223–4230.
https://doi.org/10.1182/blood-2010-05-284505

Chronic Lymphocytic Leukemia Cells in a Lymph Node Microenvironment Depict Molecular Signature Associated with an Aggressive Disease

    Amit K. Mittal, Nagendra K. Chaturvedi, Karan J. Rai, Christine E. Gilling-Cutucache, Tara M. Nordgren, Margaret Moragues, Runqing Lu, Rene Opavsky, Greg R. Bociek, Dennis D. Weisenburger, Javeed Iqbal & Shantaram S. Joshi 

Molecular Medicine volume 20, pages 290–301 (2014)


TLR Signaling Is Activated in Lymph Node–Resident CLL Cells and Is Only Partially Inhibited by Ibrutinib
Eman L. Dadashian,

Direct in vivo evidence for increased proliferation of CLL cells in lymph nodes compared to bone marrow and peripheral blood

    T M Herndon, 


\@ref(fig:microenvironmentTissues) _Adapted from Davids et al 2012_

```{r microenvironmentTissues, fig.cap='(ref:microenvironmentTissues)',out.width = "60%",  echo = FALSE}

knitr::include_graphics("figures/microenvironmentTissues.eps")

```

Post-lunch:
add remaining papers below 
Read and copy and paste things
Start writing tomorrow 

### Microenvironmental pathways and cross-talk {#intro-ME-pathways}
Review of pathways and what is missing: (use anything from this)
Proliferative Signals in Chronic Lymphocytic Leukemia; What Are We Missing?
Marco V. Haselager, 1 , 2 , 3 , 4 Arnon P. Kater, 3 , 4 , 5 and Eric Eldering 

_these then interact via cell cell contacts and pathways_
Importance of Crosstalk Between Chronic Lymphocytic Leukemia Cells and the Stromal Microenvironment: Direct Contact, Soluble Factors, and Extracellular Vesicles
Nathan Dubois,1 Emerence Crompot,1 Nathalie Meuleman,1,2 Dominique Bron,1,2 Laurence Lagneaux,1 and Basile Stamatopoulos
review of soluble factors and cell cell contacts coming from stomral cells 

_cell-cell contacts_
gneaux, L., Delforge, A., Bron, D., De Bruyn, C. and Stryckmans, P. (1998). Chronic lym-
phocytic leukemic B cells but not normal B cells are rescued from apoptosis by contact with
normal bone marrow stromal cells. Blood 91, 2387–96

anayiotidis, P., Jones, D., Ganeshaguru, K., Foroni, L. and Hoffbrand, A. V. (1996). Human
bone marrow stromal cells prevent apoptosis and support the survival of chronic lymphocytic
leukaemia cells in vitro. British Journal of Haematology 92, 97–103

Herbst, S. A., Kim, V., Schitter, E., Roider, T., Bruch, P., Kolb, C., Knoll, M., Lu, J., H ̈ullein,
J., von Kalle, C., Lutz, C., Dreger, P., M ̈uller-Tidow, C., Zenz, T., Huber, W. and Dietrich,
S. (2020a). The Influence of the Bone Marrow Niche on Drug Response Phenotypes of Blood
Cancers. Manuscript in preparation.

By providing these signals from the microenvironment to CLL cells cultured ex-vivo spontaneous
apoptosis can be prevented. Thus, it has been shown that BMSCs, one of the most important
interaction partners of B-cells in the bone marrow niche (Murphy, 2012), can rescue CLL cells
from spontaneous apoptosis (Herbst et al., 2020a; Lagneaux et al., 1998; Panayiotidis et al.,1996

best reveiw dividing cell-cell contacts and soluble factors: 
https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7466743/

_soluble factors_ 
a refernce or two on each of the pathways included <- get from Peter 
This paper covers whats known about each soluble factor: 
ne strategy is to supplement CLL cell mono-cultures with specific signalling molecules known to be present in the respective niches to analyse their impact
on cancer cells (Crassini et al., 2017). However, the addition of selected stimuli only probes the
influence of specific pathways and cannot recapitulate the complexity of the microenvironment.

within proliferation centres, all CLL cells are exposed to chemokines, integrins, cytokines and survival factors (such as tumour necrosis factor (TNF) ligand super-family member 13B (also known as BAFF) or TNF ligand superfamily member 13 (also known as APRIL)), which activate canonical nuclear factor-κB (NF-κB)90, before they exit to the blood. Activation of NF-κB can induce the expression of mir-155, which enhances BCR signalling and activation by reducing the expression of INPP5D, which encodes SHIP1 (REF. 65). Cytokines that are secreted by T cells, such as IL-4, can upregulate surface IgM, which potentially facilitates the interaction of the CLL cell with autoantigen91. In addition, the elaboration of various WNT proteins by cells in the microenvironment can activate canonical and non-canonical WNT signalling pathways92,93. Activation of the tyrosine-kinase-like transmembrane receptor ROR1 by WNT5A can induce the activation of RAC1 and RHOA, and thereby enhance CLL cell proliferation and promote migration in response to chemokines93; in part, for this reason, high-level CLL cell expression of ROR1 is associated with accelerated disease progression94. Finally, Notch signalling in response to Jagged, or Hedgehog signalling in response to Sonic Hedgehog or Desert Hedgehog, can provide pro-survival stimulation for at least a subset of patients with CLL, particularly those with trisomy 12 (REFS 95–97).

[90] BADD and APRIL activate nfkb:  Endo T, et al. BAFF and APRIL support chronic lymphocytic leukemia B-cell survival through activation of the canonical NF-kappaB pathway. Blood. 2007;109:703–710.

[65.] ACtivation of nkb enhances bcr signalling: Cui B, et al. MicroRNA-155 influences B-cell receptor signaling and associates with aggressive disease in chronic lymphocytic leukemia. Blood. 2014;124:546–554. 

[91] Aguilar-Hernandez MM, et al. IL-4 enhances expression and function of surface IgM in CLL cells. Blood. 2016;127:3015–3025

Finally, Notch signalling in response to Jagged, or Hedgehog signalling in response to Sonic Hedgehog or Desert Hedgehog, can provide pro-survival stimulation for at least a subset of patients with CLL, particularly those with trisomy 12: 
95. Seke Etet PF, Vecchio L, Nwabo Kamdje AH. Interactions between bone marrow stromal microenvironment and B-chronic lymphocytic leukemia cells: any role for Notch, Wnt and Hh signaling pathways? Cell Signal. 2012;24:1433–1443. 

96. Decker S, et al. Trisomy 12 and elevated GLI1 and PTCH1 transcript levels are biomarkers for Hedgehog-inhibitor responsiveness in CLL. Blood. 2012;119:997–1007. [PubMed] [Google Scholar]
97. De Falco F, et al. Notch signaling sustains the expression of Mcl-1 and the activity of eIF4E to promote cell survival in CLL. Oncotarget. 2015;6:16559–16572. [PMC free article] [PubMed] [Google Scholar]

[77] The lymph node microenvironment promotes B-cell receptor signaling, NF-κB activation, and tumor proliferation in chronic lymphocytic leukemia 2011 10.1182/blood-2010-05-284984
_cell-cell contacts_



_soluble factors_
what are the pahtwyas, what studies have been doen 
Whats knwonc about each pathway: Modeling the chronic lymphocytic leukemia microenvironment in vitro
Kyle Crassini, Yandong Shen, Stephen Mulligan & O. Giles Best (2016): Modeling the chronic lymphocytic leukemia microenvironment in vitro, Leukemia & Lymphoma, DOI: 10.1080/10428194.2016.1204654 To link to this article: http://dx.doi.org/10.1080/10428194.2016.1204654

pathways esp IL4 and TLR    
The microenvironment in CLL constitutes cross talk via soluble factors and cell-cell contacts, in the blood, bone marrow and lymph nodes.
_BCr signalling is central_ 
_other pathways_
_cellcell contacts_

_soluble factors_ 

For each - cover pathways activated and source 
within proliferation centres, all CLL cells are exposed to chemokines, integrins, cytokines and survival factors (such as tumour necrosis factor (TNF) ligand super-family member 13B (also known as BAFF) or TNF ligand superfamily member 13 (also known as APRIL)), which activate canonical nuclear factor-κB (NF-κB)90, before they exit to the blood. Activation of NF-κB can induce the expression of mir-155, which enhances BCR signalling and activation by reducing the expression of INPP5D, which encodes SHIP1 (REF. 65). Cytokines that are secreted by T cells, such as IL-4, can upregulate surface IgM, which potentially facilitates the interaction of the CLL cell with autoantigen91. In addition, the elaboration of various WNT proteins by cells in the microenvironment can activate canonical and non-canonical WNT signalling pathways92,93. Activation of the tyrosine-kinase-like transmembrane receptor ROR1 by WNT5A can induce the activation of RAC1 and RHOA, and thereby enhance CLL cell proliferation and promote migration in response to chemokines93; in part, for this reason, high-level CLL cell expression of ROR1 is associated with accelerated disease progression94. Finally, Notch signalling in response to Jagged, or Hedgehog signalling in response to Sonic Hedgehog or Desert Hedgehog, can provide pro-survival stimulation for at least a subset of patients with CLL, particularly those with trisomy 12 (REFS 95–97).

[90] BADD and APRIL activate nfkb:  Endo T, et al. BAFF and APRIL support chronic lymphocytic leukemia B-cell survival through activation of the canonical NF-kappaB pathway. Blood. 2007;109:703–710.

[65.] ACtivation of nkb enhances bcr signalling: Cui B, et al. MicroRNA-155 influences B-cell receptor signaling and associates with aggressive disease in chronic lymphocytic leukemia. Blood. 2014;124:546–554. 

Potentially generate a table 

IL4
[91] Aguilar-Hernandez MM, et al. IL-4 enhances expression and function of surface IgM in CLL cells. Blood. 2016;127:3015–3025

SDF 
[@tenHacken2016]
CLL chemotaxis towards stromal cells is promoted by the chemokine CXCL12 (previously called stromal cell derived factor 1 or SDF-1), which is secreted both by BMSCs [83] and by NLCs [32] (Figure 1 and Table 1)
[@tenHAcken2016]
CXCR4 stimulation is associated to prolonged CLL cell survival in vitro [32, 37] and activation of extracellular signal-regulated kinase (ERK) kinase and signal transducer and activator of transcription 3 (STAT-3) signaling [93, 94]
32. Burger JA, Tsukada N, Burger M, Zvaifler NJ, Dell’Aquila M, Kipps TJ. Blood-derived nurse-like cells protect chronic lymphocytic leukemia B cells from spontaneous apoptosis through stromal cell-derived factor-1. Blood. 2000;96:2655–2663. 
37. Nishio M, Endo T, Tsukada N, Ohata J, Kitada S, Reed JC, Zvaifler NJ, Kipps TJ. Nurselike cells express BAFF and APRIL, which can promote survival of chronic lymphocytic leukemia cells via a paracrine pathway distinct from that of SDF-1alpha. Blood. 2005;106:1012–1020.
93. Burger M, Hartmann T, Krome M, Rawluk J, Tamamura H, Fujii N, Kipps TJ, Burger JA. Small peptide inhibitors of the CXCR4 chemokine receptor (CD184) antagonize the activation, migration, and antiapoptotic responses of CXCL12 in chronic lymphocytic leukemia B cells. Blood. 2005;106:1824–1830.
94. O’Hayre M, Salanga CL, Kipps TJ, Messmer D, Dorrestein PC, Handel TM. Elucidating the CXCL12/CXCR4 signaling network in chronic lymphocytic leukemia through phosphoproteomics analysis. PloS one. 2010;5:e11716.

Notch
Finally, Notch signalling in response to Jagged, or Hedgehog signalling in response to Sonic Hedgehog or Desert Hedgehog, can provide pro-survival stimulation for at least a subset of patients with CLL, particularly those with trisomy 12: 
95. Seke Etet PF, Vecchio L, Nwabo Kamdje AH. Interactions between bone marrow stromal microenvironment and B-chronic lymphocytic leukemia cells: any role for Notch, Wnt and Hh signaling pathways? Cell Signal. 2012;24:1433–1443. 

96. Decker S, et al. Trisomy 12 and elevated GLI1 and PTCH1 transcript levels are biomarkers for Hedgehog-inhibitor responsiveness in CLL. Blood. 2012;119:997–1007. [PubMed] [Google Scholar]
97. De Falco F, et al. Notch signaling sustains the expression of Mcl-1 and the activity of eIF4E to promote cell survival in CLL. Oncotarget. 2015;6:16559–16572. [PMC free article] [PubMed] [Google Scholar]

[77] The lymph node microenvironment promotes B-cell receptor signaling, NF-κB activation, and tumor proliferation in chronic lymphocytic leukemia 2011 10.1182/blood-2010-05-284984

 CXCR4: a key receptor in the crosstalk between tumor cells and their microenvironment
Jan A. Burger,
Thomas J. Kipps
Crossmark: Check for Updates
Blood (2006) 107 (5): 1761–1767.
https://doi.org/10.1182/blood-2005-08-3182

Chemokines and chemokine receptors in chronic lymphocytic leukemia (CLL): From understanding the basics towards therapeutic targeting
Jan A.Burge
https://www.sciencedirect.com/science/article/pii/S1044579X10000830?casa_token=YINpgp597fgAAAAA:L0u76l0aqJILy3bnDTbe5wOUS7W7SCCEixqQQD49dbTYeRbO_4_nk_hqE0ctDQM7rvkTgYic6A


Chronic lymphocytic leukaemia – the role of the microenvironment pathogenesis and therapy (read section on ant-apototic signals)  https://doi.org/10.1111/bjh.12344


The important role of these niches in disease
pathogenesis is demonstrated by the fact that CLL cells undergo spontaneous apoptosis ex-
vivo (Collins et al., 1989).

in vitro CD40L and survival
The functional in vitro response to CD40 ligation reflects a different clinical outcome in patients with chronic lymphocytic leukemia https://www.nature.com/articles/leu2011149


TLR Signaling Is Activated in Lymph Node–Resident CLL Cells and Is Only Partially Inhibited by Ibrutinib
Eman L. Dadashian
https://cancerres.aacrjournals.org/content/79/2/360.short




_Studying / modelling these pathways: _
Review of ways to modeL : 
https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7773937/ MUST READ 
Modeling the Leukemia Microenviroment In Vitro
Cristina Scielzo and Paolo Ghia

What's the studies that have been done? what other approachs have people used
1. individual pathways 
This paper has study of each indificual pathway
One strategy is to supplement CLL cell mono-cultures with specific signalling molecules known to be present in the respective niches to analyse their impact
on cancer cells (Crassini et al., 2017). However, the addition of selected stimuli only probes the
influence of specific pathways and cannot recapitulate the complexity of the microenvironment.

Kyle Crassini, Yandong Shen, Stephen Mulligan & O. Giles Best (2016): Modeling the chronic lymphocytic leukemia microenvironment in vitro, Leukemia & Lymphoma, DOI: 10.1080/10428194.2016.1204654 To link to this article: http://dx.doi.org/10.1080/10428194.2016.1204654

What similar screening studies have been performed: 

AML paper :

We hypothesized that a systematic analysis of factors within the leukemia microenvironment might provide an opportunity to find more broadly effective therapeutic targets in AML. By using a unique ex vivo screen, we evaluated cytokine-driven pathways that are critical in promoting or suppressing AML cell growth. Among the 94 cytokines tested, only a few—including IL-1, GM-CSF, IL-3, M-CSF, G-CSF, and TNF-α—profoundly stimulated growth; members of the TGF-β family suppressed the growth of AML cells. These results suggest that select secreted proteins in the microenvironment influence the expansion of leukemic progenitors. In addition, this study offers a platform to systematically evaluate a large number of extrinsic factors to identify pathways critical for disease survival using primary AML patient specimens.

In a malignancy as genetically heterogeneous as AML, our use of in vitro functional screening to identify a unifying pattern of cytokine and growth-factor-driven pathway activation underscores the value of this type of assay in interrogating similar dilemmas in other genetically and molecularly complex malignancies. Furthermore, targeting aberrantly activated pathways in the tumor extrinsic environment may offer strategies to circumvent or mitigate the disease.

siRNA KDs of cytokine genes to see impact on cross talk, secpficlaly T cell synapse formation: 
https://www.sciencedirect.com/science/article/pii/S0006497119427663
Functional Screening Studies Identify Combinational Activity of PD-L1 and CD200 In Mediating Dysfunctional T Cell Immunological Synapse Formation In Chronic Lymphocytic Leukemia


NExt: 
2. entire soluble factor repetoire: 
study the influence of soluble components on CLL cells.
For this, conditioned medium is produced from cells from the microenvironment and transferred
to CLL mono-cultures. This model focuses on understanding the influence of the entire soluble
microenvironment, rather than on dissecting individual pathways in an isolated setting.

Studies using this: 

HS5CM and drug response : 

More systematically in AML : https://pubmed.ncbi.nlm.nih.gov/28355571/

3. Cell-cell contacts
Models using conditioned media take soluble factors into account, yet, they neglect the direct
interactions through cell-cell contact, which are essential for leukemia cells (Bakker et al., 2016).
Therefore, models using co-cultures of CLL and microenvironmental cells are indispensable to
gain a more complete picture on the impact of non-malignant cells on leukemia cells. 

Use different cell types: 
a) T cells One major interaction partner of CLL cells are T-cells. Such studies have
revealed that especially stimulation with CD40L and IL-21 contributes to T-cell induced prolif-
eration of CLL cells (Pascutti et al., 2013). Instead of the direct usage of T-cells, CD40L can
be expressed on the surface of fibroblasts to induce CD40-mediated signalling in CLL (Crassini
et al., 2017). Even though the CD40/CD40L system is commonly used to simulate the microen-
vironment, it again only relies on a single pathway to activate CLL cells.

b) Nurse-like cells 
Nurse-like cells have been first described by Burger et al. (2000) and protect CLL cells
from spontaneous apoptosis. The cells spontaneously outgrow from cultures of peripheral blood
mononuclear cells (PBMCs) and are of monocytic origin (Ten Hacken and Burger, 2016). Their
protective effect on CLL cells has especially been attributed to the activation of BCR and NFκB
within CLL cells upon co-culturing with nurse-like cells (Burger et al., 2009)
cite: 
Burger, J. A., Quiroga, M. P., Hartmann, E., B ̈urkle, A., Wierda, W. G., Keating, M. J. and
Rosenwald, A. (2009). High-level expression of the T-cell chemokines CCL3 and CCL4 by
chronic lymphocytic leukemia B cells in nurselike cell cocultures and after BCR stimulation.
Blood 113, 3050–8.

d) mimic the bone marrow wtih BMSCs: 
To recapitulate the interactions of CLL with the bone marrow microenvironment, especially
the influence of BMSCs needs to be taken into account. The first approaches into this direction
have been taken by co-culturing primary mesenchymal stem cells (MSCs), isolated from bone
marrow aspirates of donors, with CLL cells. These experiments have shown that MSCs generally
protect CLL cells from spontaneous apoptosis ex-vivo through soluble factors and direct cell-cell
contact (Lagneaux et al., 1998; Panayiotidis et al., 1996; Plander et al., 2011). As primary
material is not readily available for all research groups and primary cells are in general more
challenging to handle, multiple bone marrow derived stroma cell lines have been established from
murine or human donors (Kurtova et al., 2009). Two of the most commonly used human bone
marrow stroma cell lines are HS-5 and NKTert (Kawano et al., 2003; Roecklein and Torok-Storb,
1995). Both cell lines have been shown to provide protection to CLL cells from spontaneous
apoptosis, however, the effect of NKTert cells was especially pronounced (Balakrishnan et al.,
2015; Cheng et al., 2014; Fiorcari et al., 2013; Guo et al., 2017; Kurtova et al., 2009; Vangapandu
et al., 2018; Zhang et al., 2012). The cell lines have been widely used in many research studies
and helped to gain valuable insights into how CLL and the bone marrow microenvironment
interact. However, investigators should always be aware that these cell lines only represent
artificial systems and that the obtained results should always be validated in other models

cell - cell contact is essential for leukemia cells: 
Bakker, E., Qattan, M., Mutti, L., Demonacos, C. and Krstic-Demonacos, M. (2016). The role
of microenvironment and immunity in drug response in leukemia. Biochimica et Biophysica
Acta 1863, 414–426

Mimicking the tumour microenvironment: three different co-culture systems induce a similar phenotype but distinct proliferative signals in primary chronic lymphocytic leukaemia cells
Emma Hamilton, Laurence Pearce, Liam Morgan, Sophie Robinson, Vicki Ware, Paul Brennan, N. Shaun B. Thomas, Deborah Yallop, Stephen Devereux, Chris Fegan, Andrea G. S. Buggins, Chris Pepper
First published: 19 June 2012
https://doi.org/10.1111/j.1365-2141.2012.09191.x



https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4747215/
Cross-talk between chronic lymphocytic leukemia (CLL) tumor B cells and mesenchymal stromal cells (MSCs): implications for neoplastic cell survival
Valentina Trimarco,#1,2 Elisa Ave,#1,2 Monica Facco,1,2 Giorgia Chiodin,1,2 Federica Frezzato,1,2 Veronica Martini,1,2 Cristina Gattazzo,1,2 Federica Lessi,1 Carlo Alberto Giorgi,1 Andrea Visentin,1 Monica Castelli,1 Filippo Severin,1,2 Renato Zambello,1,2 Francesco Piazza,1,2 Gianpietro Semenzato,1,2 and Livio Trentin1

with scaffolds: 
Three-dimensional porous composite scaffolds for in vitro marrow microenvironment simulation to screen leukemia drug

Zhi-Yu Shang5,1, Hao Sun5,2, Hong-Fei Xiang5,3, Jun Zhang6,1, Miao Yu1,4, Huai-Song Zhao1, Chun-Ting Zhao2 and Yun-Ze Long6,1

in vivo: 
in vivo in mice: 
Modeling tumor–host interactions of chronic lymphocytic leukemia in xenografted mice to study tumor biology and evaluate targeted therapy

 S E M Herman, X Sun, E M McAuley, M M Hsieh, S Pittaluga, M Raffeld, D Liu, K Keyvanfar, C M Chapman, J Chen, J J Buggy, G Aue, J F Tisdale, P Pérez-Galán & A Wiestner 

See sectionn on mouse models of ME: 
https://ashpublications.org/blood/article/124/7/1010/33626/Mouse-models-in-the-study-of-chronic-lymphocytic

https://ashpublications.org/blood/article/119/21/4971/29934/Human-extramedullary-bone-marrow-in-mice-a-novel
Human extramedullary bone marrow in mice: a novel in vivo model of genetically controlled hematopoietic microenvironment 

THe type of BCR signalling is important: 
Mimicking the tumour microenvironment of chronic lymphocytic leukaemia in vitro critically depends on the type of B-cell receptor stimulation

    Ans Rombout, Sofie Lust, Fritz Offner, Evelien Naessens, Bruno Verhasselt & Jan Philippé

whats misisng here is a one by one dissection. 

with genetics
NF-κB activation in chronic lymphocytic leukemia: A point of convergence of external triggers and intrinsic lesions
https://www.sciencedirect.com/science/article/pii/S1044579X16300311?casa_token=IfC4bm0pVrYAAAAA:AaCfeETGlSJRllAPIzQOcv8a48vvnFmr9V125WI6_9ggFeUTbq3dr-1628CN3NXw3li9oQlgVw

\@ref(fig:microenvironmentOverview) _Adapted from: Weistner et al 2015_

```{r microenvironmentCrosstalk, fig.cap='(ref:microenvironmentOverview)', out.width = "60%", echo = FALSE}

knitr::include_graphics("figures/microenvironmentCrosstalk.eps")

```



\@ref(fig:microenvironmentSignalling) Taken from Kipps et al 2017; "Migration of chronic lymphocytic leukaemia (CLL) cells into the lyphoid tissu is primarily mediated through CXC-chemokine receptor 4 (CXCR4) in response to CXC-chemokine ligand 12 (CXCL12), which is secreted mainly by nurse-like cells (NLCs) and mesenchymal-derived stromal cells. Migration of CLL cells into lymph nodes also occurs via CC-chemokine receptor 7 (CCR7) in response to CC-chemokine ligand 19 (CCL19) and CCL21, which are produced by the endothelial cells of high endothelial venules (HEVs). HEV endothelial cells also express hyaluronan, which can interact with CD44, to facilitate B cell signalling and might enhance the production of active matrix metalloproteinase 9 (MMP9). Once in tissues, several chemokines promote B cell survival, including CXCL12, B cell-activating factor (BAFF; also known as TNFSF13B) and a proliferation-inducing ligand (APRIL; also known as TNFSF13). In addition, CLL cell survival can be promoted through cognate interactions between CD31 and CD38, and the production by stromal cells of WNT factors, which can interact with ROR1, ROR2 and/or various Frizzled receptors. CLL cell contact with mesenchymal stromal cells can also be established through vascular cell adhesion protein 1 (VCAM1)– α4β 1 integrin interactions that contribute to CLL cell survival. In turn, CLL cells can secrete chemokines, such as CCL3 and CCL4, which can recruit T cells and NLC-precursor cells (monocytes) to the CLL microenvironment. Activated T cells can provide CLL cells with proliferative signals through CD40 ligand (CD40L)-CD40 interactions and the secretion of several cytokines, such as IL-2, IL-4 and IL-10. In turn, activated CLL cells secrete CCL12 and CCL22, which attract more T cells into the CLL microenvironment. In tissues, CLL cells can be exposed to environmental and/or self-antigens that might trigger B cell activation through interactions with the surface immunoglobulin; this could amplify the responsiveness of CLL cells to the signals and factors that are provided by the CLL microenvironment. BAFFR, BAFF receptor (also known as TNFRSF13C); BCMA, B cell maturation protein (also known as TNFRSF17); BCR, B cell receptor; TACI, transmembrane activator and CAML interactor (also known as TNFRSF13B). _Figure from @Kipps2017 ._ "

```{r microenvironmentSignalling, fig.cap='(ref:microenvironmentSignalling)', out.width = "60%", echo = FALSE}

knitr::include_graphics("figures/microenivironmental_signalling.eps")

```



### The impact of the microenvironment on drug response 

Reviews
Treatment of chronic lymphocytic leukemia requires targeting of the protective lymph node environment with novel therapeutic approaches
Rachel E. Hayden
2012
https://doi.org/10.3109/10428194.2011.610014 

_intro_
Underlying mechanisms and drug intervention strategies for the tumour microenvironment 2021
https://jeccr.biomedcentral.com/articles/10.1186/s13046-021-01893-y

Updates on the hematologic tumor microenvironment and its therapeutic targeting
Dorian Forte,1,2 Daniela S. Krause,3 Michael Andreeff,4 Dominique Bonnet,5 and Simón Méndez-Ferrer2019

Addressing the tumour microenvironment in early drug discovery: a strategy to overcome drug resistance and identify novel targets for cancer therapy
https://doi.org/10.1016/j.drudis.2020.11.030

THERAPEUTIC MODALITIES: Edited by Steven T. Rosen
Targeting the microenvironment in chronic lymphocytic leukemia is changing the therapeutic landscape
Burger, Jan A.
https://journals.lww.com/co-oncology/FullText/2012/11000/Targeting_the_microenvironment_in_chronic.7.aspx

_cutting edge_
As well as providing protective signals that support tumour growth, the microenvironment is also believed to reduce treatment efficacy within the protective niche [@Ahn2018] 

Bakker, E., Qattan, M., Mutti, L., Demonacos, C. and Krstic-Demonacos, M. (2016). The role
of microenvironment and immunity in drug response in leukemia. Biochimica et Biophysica
Acta 1863, 414–426

During therapy, especially with new targeted agents, clearance of leukemia cells from the blood
can often be achieved. However, in most cases minimal residual disease, particularly in the bone
marrow, persists (Burger and O’Brien, 2018). Due to this, it is assumed that the niche of the
bone marrow protects CLL cells from drug-induced apoptosis and, thereby, contributes to the
emergence of resistant clones (Ten Hacken and Burger, 2016).

[@tenHacken2016]
CLL cells are protected from spontaneous and drug-induced 
apoptosis by direct contact with BMSCs [48, 49]

48. Lagneaux L, Delforge A, Bron D, De Bruyn C, Stryckmans P. Chronic lymphocytic leukemic B cells but not normal B cells are rescued from apoptosis by contact with normal bone marrow stromal cells. Blood. 1998;91:2387–2396. [PubMed] [Google Scholar]
49. Kurtova AV, Balakrishnan K, Chen R, Ding W, Schnabl S, Quiroga MP, Sivina M, Wierda WG, Estrov Z, Keating MJ, Shehata M, Jager U, Gandhi V, Kay NE, Plunkett W, Burger JA. Diverse marrow stromal cells protect CLL cells from spontaneous and drug-induced apoptosis: development of a reliable and reproducible system to assess stromal cell adhesion-mediated drug resistance. Blood. 2009;114:4441–4450.


Direct in vivo evidence for increased proliferation of CLL cells in lymph nodes compared to bone marrow and peripheral blood
T M Herndon

in vitro: 
Few studies exist which have investigated the influence of the bone marrow on the response to selected drugs to further understand the mechanisms of drug resistance.  

Kay, N. E., Shanafelt, T. D., Strege, A. K., Lee, Y. K., Bone, N. D. and Raza, A. (2007). Bone
biopsy derived marrow stromal elements rescue chronic lymphocytic leukemia B-cells from
spontaneous and drug induced cell death and facilitates an ”angiogenic switch”. Leukemia
Research 31, 899–906.


Nurse-like cells mediate ibrutinib resistance in chronic lymphocytic leukemia patients
F Boissard

_examples_
The chemotherapeutics fludarabine, oxaliplatin, chlo-rambucil, cyclophosphamide and doxorubicine have been described to have diminished effects in co-cultures with stroma cells (Kay et al., 2007; Kurtova et al., 2009; Mraz et al., 2011; Zhang et al., 2012):

Mraz, M., Zent, C. S., Church, A. K., Jelinek, D. F., Wu, X., Pospisilova, S., Ansell, S. M.,
Novak, A. J., Kay, N. E., Witzig, T. E. and Nowakowski, G. S. (2011). Bone marrow stromal
cells protect lymphoma B-cells from rituximab-induced apoptosis and targeting integrin α-4-
β-1 (VLA-4) with natalizumab can overcome this resistance. British Journal of Haematology
155, 53–64

Zhang, W., Trachootham, D., Liu, J., Chen, G., Pelicano, H., Garcia-Prieto, C., Lu, W., Burger,
J. A., Croce, C. M., Plunkett, W., Keating, M. J. and Huang, P. (2012). Stromal control of
cystine metabolism promotes cancer cell survival in chronic lymphocytic leukaemia. Nature
Cell Biology 14, 276–286.

reduced efficiency for the corticosteroid medications hydrocortisone and dexamethasone (Kurtova et al., 2009; Panayiotidis et al., 1996). 

Kurtova, A. V., Balakrishnan, K., Chen, R., Ding, W., Schnabl, S., Quiroga, M. P., Sivina, M.,
Wierda, W. G., Estrov, Z., Keating, M. J., Shehata, M., J ̈ager, U., Gandhi, V., Kay, N. E.,
Plunkett, W. and Burger, J. A. (2009). Diverse marrow stromal cells protect CLL cells from
spontaneous and drug-induced apoptosis: Development of a reliable and reproducible system
to assess stromal cell adhesion-mediated drug resistance. Blood 114, 4441–4450

protection of CLL cells from carilzomib-induced apoptosis has been mentioned (Gupta
et al., 2013). 

Gupta, S. V., Hertlein, E., Lu, Y., Sass, E. J., Lapalombella, R., Chen, T. L., Davis, M. E.,
Woyach, J. A., Lehman, A., Jarjoura, D., Byrd, J. C. and Lucas, D. M. (2013). The Pro-
teasome Inhibitor Carfilzomib Functions Independently of p53 to Induce Cytotoxicity and an
Atypical NF- B Response in Chronic Lymphocytic Leukemia Cells. Clinical Cancer Research
19, 2406–2419.

IL4 induced resistance to ibrutinib - aguilar - hernandes: 

TLr and fludarabine: 
n Vitro Sensitivity of CLL Cells to Fludarabine May Be Modulated by the Stimulation of Toll-like Receptors
Eleonora Fonte

TLR induced resistance to venetoclax: 
 Microenvironmental agonists generate de novo phenotypic resistance to combined ibrutinib plus venetoclax in CLL and MCL
Kallesh D. Jayappa,
Craig A. Portell,
Vicki L. Gordon,
Brian J. Capaldo,
Stefan Bekiranov,
Mark J. Axelrod,
L. Kyle Brett,
Julia D. Wulfkuhle,
Rosa I. Gallagher,
Emanuel F. Petricoin,
Timothy P. Bender,
Michael E. Williams,
Michael J. Weber

BAFF and ibrutinib: 
McWilliams, E.M.; Lucas, C.R.; Chen, T.; Harrington, B.K.; Wasmuth, R.; Campbell, A.; Rogers, K.A.; Cheney, C.M.; Mo, X.; Andritsos, L.A.; et al. Anti-BAFF-R antibody VAY-736 demonstrates promising preclinical activity in CLL and enhances effectiveness of ibrutinib. Blood Adv. 2019, 3, 447–460.

 stroma cell mediated resistances to ibrutinib
treatment (Cheng et al., 2014; Guo et al., 2017), which might explain why complete eradication
of leukemia cells from the bone marrow is rare during therapy with ibrutinib (Byrd et al., 2013).
also enalrged lymph nodes with ibrutinib treatment: 

Cheng, S., Ma, J., Guo, A., Lu, P., Leonard, J. P., Coleman, M., Liu, M., Buggy, J. J., Furman,
R. R. and Wang, Y. L. (2014). BTK inhibition targets in vivo CLL proliferation through its
effects on B-cell receptor signaling activity. Leukemia 28, 649–57

Guo, A., Lu, P., Coffey, G., Conley, P., Pandey, A. and Wang, Y. L. (2017). Dual SYK/JAK
inhibition overcomes ibrutinib resistance in chronic lymphocytic leukemia: Cerdulatinib, but
not ibrutinib, induces apoptosis of tumor cells protected by the microenvironment. Oncotarget
8, 12953–12967.

Byrd, J. C., Furman, R. R., Coutre, S. E., Flinn, I. W., Burger, J. A., Blum, K. A., Grant,
B., Sharman, J. P., Coleman, M., Wierda, W. G., Jones, J. A., Zhao, W., Heerema, N. A.,
Johnson, A. J., Sukbuntherng, J., Chang, B. Y., Clow, F., Hedrick, E., Buggy, J. J., James,
D. F. and O’Brien, S. (2013). Targeting BTK with ibrutinib in relapsed chronic lymphocytic
leukemia. The New England Journal of Medicine 369, 32–42.


[@tenHacken2016]
 NK cells can also produce soluble BAFF, which interferes with 
NK-cell mediated CLL cell lysis after rituximab administration [82]

dexamehtasone and chlorambucil 
Guilty bystanders: nurse-like cells as a model of microenvironmental support for leukemic lymphocytes

    Agata A. Filip
    
_cutting edge of drug - microenvironment interactopns_
(are there any other papers like this)
A Multi-Drug Tolerant Phenotype Is Induced By Diverse Microenvironmental Agonists in Chronic Lymphocytic Leukemia and Mantle Cell Lymphoma
Kallesh D.JayappaDVM, PhD*12Konrad J.CiosBSc (2019)*12Vicki L.GordonBSc*12Puja C.AroraMD3Timothy P.BenderPhD*12Michael E.WilliamsMD34Craig A.PortellMD45Michael J.WeberPhD
https://www.sciencedirect.com/science/article/pii/S0006497119402206

drug screen of stomral cells to identify best drugs to inhhibiti pathways: 
https://www.nature.com/articles/s41598-020-78315-0
Systems biology drug screening identifies statins as enhancers of current therapies in chronic lymphocytic leukemia2020

soe  studies taht look for individaul cases where certain drugs act to increase otehr drug efficacy in te hcontext of ME signals

Screening of patient spceicfc pathways that induce Venetoclax resistance: 
https://ashpublications.org/blood/article/128/7/934/35889/High-content-screening-identifies-kinase
High-content screening identifies kinase inhibitors that overcome venetoclax resistance in activated CLL cells 


in vivo: 
https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4747215/
Cross-talk between chronic lymphocytic leukemia (CLL) tumor B cells and mesenchymal stromal cells (MSCs): implications for neoplastic cell survival
Valentina Trimarco,#1,2 Elisa Ave,#1,2 Monica Facco,1,2 Giorgia Chiodin,1,2 Federica Frezzato,1,2 Veronica Martini,1,2 Cristina Gattazzo,1,2 Federica Lessi,1 Carlo Alberto Giorgi,1 Andrea Visentin,1 Monica Castelli,1 Filippo Severin,1,2 Renato Zambello,1,2 Francesco Piazza,1,2 Gianpietro Semenzato,1,2 and Livio Trentin1
look at section: 
MSCs from CLL patients protect leukemic B cells after in vivo treatment with fludarabine and cyclophosphamide containing regimen

There are models to observed genetic ibrutinibresistance in vivo models, but not for ME: 
 Development and characterization of prototypes for in vitro and in vivo mouse models of ibrutinib-resistant CLL
Burcu Aslan

_strategies to overcome_

https://www.sciencedirect.com/science/article/pii/S0006497119386987
Drug Combinations Co-Targeting Myeloid Cell Leukemia-1 (Mcl-1) Protein Can Overcome Microenvironmentally-Induced Multi-Drug Tolerance in Non-Hodgkin Lymphomas

One suggestion for overcoming these resistances is the simultaneous targeting of SYK and JAK
with the drug cerdulatinib (Guo et al., 2017). 


Guo, A., Lu, P., Coffey, G., Conley, P., Pandey, A. and Wang, Y. L. (2017). Dual SYK/JAK
inhibition overcomes ibrutinib resistance in chronic lymphocytic leukemia: Cerdulatinib, but
not ibrutinib, induces apoptosis of tumor cells protected by the microenvironment. Oncotarget
8, 12953–12967.

the BCL-2 inhibitor AT-101 was ble to overcome stroma mediated protection, which might represent an interesting option for avoiding resistance (Balakrishnan et al., 2009). 

Balakrishnan, K., Burger, J. A., Wierda, W. G. and Gandhi, V. (2009). AT-101 induces apopto-
sis in CLL B cells and overcomes stromal cell-mediated Mcl-1 induction and drug resistance.
Blood 113, 149–53.

With lenalidomide: 

[@tenHacken]
Targeting the microenvironment wtih lenalidomid and and CXCR4/CXCL12 inhibitors
Lenalidomide is an immunomodulatory agent, which interferes with multiple components of the CLL microenvironment. Lenalidomide only induces mild apoptosis of leukemic cells, but reduces CLL proliferation through a cereblon/p21 dependent mechanism [142] and interferes with NLC-mediated [143] and endothelium-mediated [111] survival support. Lenalidomide has pleiotropic effects on the CLL microenvironment: it increases CD4+ T-mediated antigen presentation, proliferation and activity [144, 145], and enhances NK and CD4+ T-cell mediated anti-tumor immune responses [146, 147]. Lenalidomide restores functional immune synapse formation between T and CLL cells, down-regulates the immunosuppressive axis PD-1/PD-L1, and enhances T-cell motility [71, 73]; it also causes B-cell activation [148], and interferes with the activity and proliferation of T-regulatory cells [149]. Lenalidomide is active alone, in CLL relapsed/refractory patients [150, 151], or as initial treatment for elderly patients [152, 153] or in combination with rituximab [150, 154] and is currently tested in combination with ibrutinib [155].

142. Fecteau JF, Corral LG, Ghia EM, Gaidarova S, Futalan D, Bharati IS, Cathers B, Schwaederle M, Cui B, Lopez-Girona A, Messmer D, Kipps TJ. Lenalidomide inhibits the proliferation of CLL cells via a cereblon/p21(WAF1/Cip1)-dependent mechanism independent of functional p53. Blood. 2014;124:1637–1644. [PMC free article] [PubMed] [Google Scholar]
143. Schulz A, Durr C, Zenz T, Dohner H, Stilgenbauer S, Lichter P, Seiffert M. Lenalidomide reduces survival of chronic lymphocytic leukemia cells in primary cocultures by altering the myeloid microenvironment. Blood. 2013;121:2503–2511. [PubMed] [Google Scholar]
144. Aue G, Njuguna N, Tian X, Soto S, Hughes T, Vire B, Keyvanfar K, Gibellini F, Valdez J, Boss C, Samsel L, McCoy JP, Jr, Wilson WH, Pittaluga S, Wiestner A. Lenalidomide-induced upregulation of CD80 on tumor cells correlates with T-cell activation, the rapid onset of a cytokine release syndrome and leukemic cell clearance in chronic lymphocytic leukemia. Haematologica. 2009;94:1266–1273. [PMC free article] [PubMed] [Google Scholar]
145. Lee BN, Gao H, Cohen EN, Badoux X, Wierda WG, Estrov Z, Faderl SH, Keating MJ, Ferrajoli A, Reuben JM. Treatment with lenalidomide modulates T-cell immunophenotype and cytokine production in patients with chronic lymphocytic leukemia. Cancer. 2011;117:3999–4008. [PMC free article] [PubMed] [Google Scholar]
146. Acebes-Huerta A, Huergo-Zapico L, Gonzalez-Rodriguez AP, Fernandez-Guizan A, Payer AR, Lopez-Soto A, Gonzalez S. Lenalidomide induces immunomodulation in chronic lymphocytic leukemia and enhances antitumor immune responses mediated by NK and CD4 T cells. BioMed research international. 2014;2014:265840. [PMC free article] [PubMed] [Google Scholar]
147. Wu L, Adams M, Carter T, Chen R, Muller G, Stirling D, Schafer P, Bartlett JB. lenalidomide enhances natural killer cell and monocyte-mediated antibody-dependent cellular cytotoxicity of rituximab-treated CD20+ tumor cells. Clinical cancer research: an official journal of the American Association for Cancer Research. 2008;14:4650–4657. [PubMed] [Google Scholar]
148. Lapalombella R, Andritsos L, Liu Q, May SE, Browning R, Pham LV, Blum KA, Blum W, Ramanunni A, Raymond CA, Smith LL, Lehman A, Mo X, Jarjoura D, Chen CS, Ford R, Jr, Rader C, Muthusamy N, Johnson AJ, Byrd JC. Lenalidomide treatment promotes CD154 expression on CLL cells and enhances production of antibodies by normal B cells through a PI3-kinase-dependent pathway. Blood. 2010;115:2619–2629. [PMC free article] [PubMed] [Google Scholar]
149. Galustian C, Meyer B, Labarthe MC, Dredge K, Klaschka D, Henry J, Todryk S, Chen R, Muller G, Stirling D, Schafer P, Bartlett JB, Dalgleish AG. The anti-cancer agents lenalidomide and pomalidomide inhibit the proliferation and function of T regulatory cells. Cancer Immunol Immunother. 2009;58:1033–1045. [PubMed] [Google Scholar]
150. Ferrajoli A, Lee BN, Schlette EJ, O’Brien SM, Gao H, Wen S, Wierda WG, Estrov Z, Faderl S, Cohen EN, Li C, Reuben JM, Keating MJ. Lenalidomide induces complete and partial remissions in patients with relapsed and refractory chronic lymphocytic leukemia. Blood. 2008;111:5291–5297. [PMC free article] [PubMed] [Google Scholar]
151. Chanan-Khan A, Miller KC, Lawrence D, Padmanabhan S, Miller A, Hernandez-Illatazurri F, Czuczman MS, Wallace PK, Zeldis JB, Lee K. Tumor flare reaction associated with lenalidomide treatment in patients with chronic lymphocytic leukemia predicts clinical response. Cancer. 2011;117:2127–2135. [PubMed] [Google Scholar]
152. Badoux XC, Keating MJ, Wen S, Lee BN, Sivina M, Reuben J, Wierda WG, O’Brien SM, Faderl S, Kornblau SM, Burger JA, Ferrajoli A. Lenalidomide as initial therapy of elderly patients with chronic lymphocytic leukemia. Blood. 2011;118:3489–3498. [PMC free article] [PubMed] [Google Scholar]
153. Chen CI, Bergsagel PL, Paul H, Xu W, Lau A, Dave N, Kukreti V, Wei E, Leung-Hagesteijn C, Li ZH, Brandwein J, Pantoja M, Johnston J, Gibson S, Hernandez T, Spaner D, Trudel S. Single-agent lenalidomide in the treatment of previously untreated chronic lymphocytic leukemia. Journal of clinical oncology: official journal of the American Society of Clinical Oncology. 2011;29:1175–1181. [PMC free article] [PubMed] [Google Scholar]
154. James DF, Werner L, Brown JR, Wierda WG, Barrientos JC, Castro JE, Greaves A, Johnson AJ, Rassenti LZ, Rai KR, Neuberg D, Kipps TJ. Lenalidomide and rituximab for the initial treatment of patients with chronic lymphocytic leukemia: a multicenter clinical-translational study from the chronic lymphocytic leukemia research consortium. Journal of clinical oncology: official journal of the American Society of Clinical Oncology. 2014;32:2067–2073. [PMC free article] [PubMed] [Google Scholar]
155. Pollyea DA, Coutre S, Gore L, Adler N, Harris P, Phelps MA, Johnson AJ, Ling Y, Li H, Gutman JA, Byrd JC. A Dose Escalation Study of Ibrutinib with Lenalidomide for Relapsed and Refractory Chronic Lymphocytic Leukemia/Small Lymphocytic Lymphoma. Blood (ASH annual meeting abstracts) 2014;124:1987. [Google Scholar]

With ibrutinib and idelalaisib:
Targeting with ibrutiniband BCR kinase inhibtiros: 
[@tenHacken2016]
Irbutinib: 
everal small molecule BCR signaling inhibitors, mainly targeting BTK, PI3K and SYK kinases, have been generated and have shown excellent clinical activity (
brutinib is capable of overcoming pro-survival signals derived from the CLL microenvironment in vitro, including those from NLC- contact, CD40 ligation, BAFF, fibronectin (FN), IL-6, IL-4, TNFα [165] and BCR stimulation [166]. Ibrutinib inhibits CLL-cell proliferation [166], chemotaxis towards CXCL12 and CXCL13 [166, 167], integrin-mediated adhesion [167], CLL-cell release of exosomes [168] and of CCL3 and CCL4 chemokines, in vitro and in CLL patients receiving ibrutinib therapy [166]. 

165. Herman SE, Gordon AL, Hertlein E, Ramanunni A, Zhang X, Jaglowski S, Flynn J, Jones J, Blum KA, Buggy JJ, Hamdy A, Johnson AJ, Byrd JC. Bruton’s tyrosine kinase represents a promising therapeutic target for treatment of chronic lymphocytic leukemia and is effectively targeted by PCI-32765. Blood. 2011;117:6287–6296. [PMC free article] [PubMed] [Google Scholar]
166. Ponader S, Chen SS, Buggy JJ, Balakrishnan K, Gandhi V, Wierda WG, Keating MJ, O’Brien S, Chiorazzi N, Burger JA. The Bruton tyrosine kinase inhibitor PCI-32765 thwarts chronic lymphocytic leukemia cell survival and tissue homing in vitro and in vivo. Blood. 2012;119:1182–1189. [PMC free article] [PubMed] [Google Scholar]
167. de Rooij MF, Kuil A, Geest CR, Eldering E, Chang BY, Buggy JJ, Pals ST, Spaargaren M. The clinically active BTK inhibitor PCI-32765 targets B-cell receptor- and chemokine-controlled adhesion and migration in chronic lymphocytic leukemia. Blood. 2012;119:2590–2594

Idelalisib: 
[@tenHacken2016]
Idelalisib is a highly selective PI3Kδ inhibitor [183], which antagonizes CLL-survival signals coming from the microenvironment, such as NLC-contact [184], CD40 ligation, TNFα, fibronectin and BCR stimulation [165, 183]. Idelalisib reduces CLL-cell chemotaxis towards CXCL12 and CXCL13 [184], CCL3 and CCL4 release by CLL cells in vitro and in patients receiving idelalisib therapy [184], and synergizes with ibrutinib in reducing CLL adhesion to vascular cell-adhesion molecule-1 (VCAM-1) and fibronectin [185].

183. Lannutti BJ, Meadows SA, Herman SE, Kashishian A, Steiner B, Johnson AJ, Byrd JC, Tyner JW, Loriaux MM, Deininger M, Druker BJ, Puri KD, Ulrich RG, Giese NA. CAL-101, a p110{delta} selective phosphatidylinositol-3-kinase inhibitor for the treatment of B-cell malignancies, inhibits PI3K signaling and cellular viability. Blood. 2011;117:591–594. [PMC free article] [PubMed] [Google Scholar]
184. Hoellenriegel J, Meadows SA, Sivina M, Wierda WG, Kantarjian H, Keating MJ, Giese N, O’Brien S, Yu A, Miller LL, Lannutti BJ, Burger JA. The phosphoinositide 3′-kinase delta inhibitor, CAL-101, inhibits B-cell receptor signaling and chemokine networks in chronic lymphocytic leukemia. Blood. 2011;118:3603–3612. [PMC free article] [PubMed] [Google Scholar]
185. de Rooij MF, Kuil A, Kater AP, Kersten MJ, Pals ST, Spaargaren M. Ibrutinib and idelalisib synergistically target BCR-controlled adhesion in MCL and CLL: a rationale for combination therapy. Blood. 2015;125:2306–2309. 


SYK inhibtion: 

[@tenHacken2016]
SYK kinase belongs to the SYK/ZAP70 family of non-receptor kinases, and activates signaling pathways downstream of the BCR, chemokine and integrin receptors, suggesting that SYK participates in tissue homing and retention of activated B cells [199]. Fostamatinib (FosD, R788) is an orally available inhibitor of SYK, which induced partial responses in relapsed CLL patients in phase I/II study [200], but further development of this drug focused on rheumatoid arthritis [201]. Additional SYK-specific inhibitors are under development, including GS-9973, which has been tested alone [202] or in combination with idelalisib, and PRT-2070.

_how this can be modelled_
Kurtova, A. V., Balakrishnan, K., Chen, R., Ding, W., Schnabl, S., Quiroga, M. P., Sivina, M.,
Wierda, W. G., Estrov, Z., Keating, M. J., Shehata, M., J ̈ager, U., Gandhi, V., Kay, N. E.,
Plunkett, W. and Burger, J. A. (2009). Diverse marrow stromal cells protect CLL cells from
spontaneous and drug-induced apoptosis: Development of a reliable and reproducible system
to assess stromal cell adhesion-mediated drug resistance. Blood 114, 4441–4450


get again from sophie thesis 

However, to date a systematic investigation of the
effect of stroma cells on drug response is lacking. Only a comprehensive evaluation will enable
a better understanding of these interactions and allow the development of efficient strategies to
overcome the emergence of drug resistances in the bone marro
 
 in an in vivo model: 
82. Wild J, Schmiedel BJ, Maurer A, Raab S, Prokop L, Stevanovic S, Dorfel D, Schneider P, Salih HR. Neutralization of (NK-cell-derived) B-cell activating factor by Belimumab restores sensitivity of chronic lymphoid leukemia cells to direct and Rituximab-induced NK lysis. Leukemia. 2015 Epub ahead of print. 


## Therapies in CLL 
Most up to date review: https://www.nejm.org/doi/full/10.1056/NEJMra1908213 Treatment of Chronic Lymphocytic Leukemia

SYK kinase belongs to the SYK/ZAP70 family of non-receptor kinases, and activates signaling pathways downstream of the BCR, chemokine and integrin receptors, suggesting that SYK participates in tissue homing and retention of activated B cells [199]. Fostamatinib (FosD, R788) is an orally available inhibitor of SYK, which induced partial responses in relapsed CLL patients in phase I/II study [200], but further development of this drug focused on rheumatoid arthritis [201]. Additional SYK-specific inhibitors are under development, including GS-9973, which has been tested alone [202] or in combination with idelalisib, and PRT-2070.### Clinical regimes 
[@Kipps2017]
The treatment of patients with CLL can include chemotherapy, a combination of chemotherapy and immunotherapy, or drugs that target the signalling pathways that promote the growth and/or survival of CLL cells (for example, BCR signalling and BCL-2)128,129.

Chemotherapy—Chemotherapy has been the mainstay of therapy for the past 50 years. Purine analogues (most commonly fludarabine, but also pentostatin or cladribine) and alkylating agents (including chlorambucil, cyclophosphamide or bendamustine) are used in the treatment of CLL130–132. Chemotherapy-based regimens can cause myelosuppression, an increased risk of infections and, in a small subset of patients, post-therapy myelodysplasia or secondary cancers, such as acute myeloid leukaemia (see Secondary cancers).

Chemoimmunotherapy—Phase III clinical trials have validated the benefit of anti-CD20 monoclonal antibodies, such as rituximab, obinutuzumab or ofatumumab, in combination with chemotherapy for the treatment of patients with CLL. In one trial (the CLL8 trial of the German CLL Study Group), patients who received fludarabine and cyclophosphamide with rituximab had higher response rates and a longer median progression-free survival (PFS) than patients who were treated with fludarabine and cyclophosphamide133. In a separate study (the CLL11 trial), patients >65 years of age with medical comorbidities who were treated with chlorambucil and either obinutuzumab or rituximab had improved response rates and longer median PFS than patients who were treated with chlorambucil alone134. However, the median PFS was significantly longer for patients who received obinutuzumab (26.7 months) than in those who received rituximab (11.1 months). In a third phase III trial, median PFS significantly improved from 13.1 months for patients treated with chlorambucil to only 22.4 months for patients treated with chlorambucil and ofatumumab135. As a consequence of these three trials, the US FDA approved the use of rituximab, obinutuzumab or ofatumumab in combination with chemotherapy for the first-line treatment of patients with CLL. The FDA also approved the use of ofatumumab as a single agent for the treatment of patients with relapsed or refractory disease based on data from a phase II study136.

_cutting edge of drug - cyt gene interactions_ 


### New Targeted therapies
in particular Ibrutinib and idelalisib 

[@Kipps2017]
Three main classes of drugs that each can inhibit BCR signalling have been evaluated in patients with CLL: BTK inhibitors, PI3K inhibitors and spleen tyrosine kinase (SYK) inhibitors86,143. CLL cells with unmutated IGHV seem to be more sensitive to inhibitors of BCR signalling than CLL cells with mutated IGHV144, but whether inhibitors, such as ibrutinib, are more effective in patients with CLL and unmutated IGHV, remains to be validated in clinical trials.
Ibrutinib has been approved in the United States and Europe for use as initial therapy, as well as in patients with relapsed disease, which followed results from a randomized trial that showed a significantly higher response rate to therapy with ibrutinib than with ofatumumab145. In addition, with continuous therapy, patients treated with ibrutinib had a significantly longer median PFS and overall survival than patients treated for 8 months with ofatumumab. Approval of ibrutinib as initial therapy was based on the results of a randomized trial that showed a significant improvement in median PFS and overall survival in patients 65 years of age without del(17p) who were treated indefinitely with ibrutinib than in patients treated for up to 48 weeks with chlorambucil146.
Upon initiation of treatment with ibrutinib, lymphadenopathy is rapidly reduced, which is associated with a concomitant increase in absolute lymphocyte count147. The rise in absolute lymphocyte count is related to the inhibition of chemokine receptor signalling, which inhibits the migration of CLL cells from the blood into the lymphoid tissues. This resulting lymphocytosis should not be considered a sign of progression; over time, the lymphocytosis subsides as the overall tumour burden decreases with continued therapy.

[@Kipps2017]
PI3K inhibitors include idelalisib, duvelisib (also known as IPI-145), TGR-1022 and ACP-319 (also known as AMG-319)151; the latter three drugs are being evaluated in clinical trials, whereas idelalisib was approved in the United States and Europe for the treatment of patients with relapsed CLL

[@Kipps2017]
Phase I/II clinical trials of fostamatinib, an oral SYK inhibitor, caused reduction in lymphadenopathy with concomitant lymphocytosis, an improvement in disease-related cytopenias and relief of disease-related symptoms in most of the treated patients with CLL156. However, dose-limiting toxicities of fostamatinib treatment include neutropenia,thrombocytopenia and diarrhoea. Other inhibitors of SYK, such as entospletinib, are being evaluated in preclinical and clinical studies.

[@Kipps2017]
BCL-2 inhibitors—Venetoclax is a small molecule that functions as a BH3 mimetic to inhibit BCL-2 (REF. 157). This drug is highly potent in inducing apoptosis in CLL cells, possibly by diminishing the capacity of BCL-2 to sequester the pro-apoptotic molecule BCL-2-interacting mediator of cell death (BIM; also known as BCL2L11)158. As such, venetoclax is effective in patients with relapsed and/or refractory disease159 or in patients with relapsed disease and del(17p)160. Indeed, the overall response rate for patients with relapsed disease and del(17p) was 79%, with 8% achieving a complete response. In addition, the estimated 12-month PFS was 72% and overall survival was 87%. On the basis of these results, the FDA approved the use of venetoclax for patients with relapsed disease and del(17p). Ongoing studies have shown that venetoclax can be safely combined with rituximab or obinutuzumab. Moreover, studies are examining the use of venetoclax with or without an anti-CD20 monoclonal antibody, and with or without ibrutinib161,162, which might provide higher response rates to therapy than that with venetoclax alone.

Idelalisib: 
Main: 
Brown, J. R., Byrd, J. C., Coutre, S. E., Benson, D. M., Flinn, I. W., Wagner-Johnston, N. D.,
Spurgeon, S. E., Kahl, B. S., Bello, C., Webb, H. K., Johnson, D. M., Peterman, S., Li, D.,
Jahn, T. M., Lannutti, B. J., Ulrich, R. G., Yu, A. S., Miller, L. L. and Furman, R. R. (2014).
Idelalisib, an inhibitor of phosphatidylinositol 3-kinase p110δ, for relapsed/refractory chronic
lymphocytic leukemia. Blood 123, 3390–7

Also: 
[@tenHacken2016]

PI3Ks are divided into 3 classes (I through III) and class I is further composed by four different isoforms (α,β,γ,δ). PI3Ks are responsible for the activation of AKT kinase along the BCR signaling pathway, as well as they exert effects on cell metabolism, migration, proliferation, and survival [182]. The predominant form expressed by hematopoietic cells is PI3Kδ, which plays a critical role in B-cell homeostasis and function. Idelalisib is a PI3Kδ inhibitor, which was approved by the FDA in 2014 for the treatment of previously treated CLL patients when used in combination with rituximab. Idelalisib is a highly selective PI3Kδ inhibitor 

182. Okkenhaug K, Vanhaesebroeck B. PI3K in lymphocyte development, differentiation and activation. Nat Rev Immunol. 2003;3:317–330. [PubMed] [Google Scholar


Ibrutinib: 
Main 
Byrd, J. C., Furman, R. R., Coutre, S. E., Flinn, I. W., Burger, J. A., Blum, K. A., Grant,
B., Sharman, J. P., Coleman, M., Wierda, W. G., Jones, J. A., Zhao, W., Heerema, N. A.,
Johnson, A. J., Sukbuntherng, J., Chang, B. Y., Clow, F., Hedrick, E., Buggy, J. J., James,
D. F. and O’Brien, S. (2013). Targeting BTK with ibrutinib in relapsed chronic lymphocytic
leukemia. The New England Journal of Medicine 369, 32–42.

Also: 
[@tenHacken2016]
BTK is a non-receptor tyrosine kinase of the Tec family, and is rapidly phosphorylated by both LYN and SYK kinases upon BCR engagement, resulting in the activation of PLCγ2, AKT and ERK kinases, and NF-κB signaling [161]. BTK mutations in humans are associated to X-linked agammaglobulinemia, a primary immunodeficiency characterized by the absence of peripheral blood B cells and decreased levels of serum immunoglobulins [161]. In addition, BTK is also involved in the regulation of migration and adhesion via CXCR4/CXCR5 and integrin signaling [162]. Ibrutinib is an orally bioavailable inhibitor which was approved in 2014 for the treatment of mantle cell lymphoma and CLL; it blocks BTK kinase activity by forming a covalent bond with cysteine-481 residue [163, 164]. 

161. Takata M, Kurosaki T. A role for Bruton’s tyrosine kinase in B cell antigen receptor-mediated activation of phospholipase C-gamma 2. The Journal of experimental medicine. 1996;184:31–40. [PMC free article] [PubMed] [Google Scholar]
162. de Gorter DJ, Beuling EA, Kersseboom R, Middendorp S, van Gils JM, Hendriks RW, Pals ST, Spaargaren M. Bruton’s tyrosine kinase and phospholipase Cgamma2 mediate chemokine-controlled B cell migration and homing. Immunity. 2007;26:93–104. [PubMed] [Google Scholar]
163. Rushworth SA, Bowles KM, Barrera LN, Murray MY, Zaitseva L, MacEwan DJ. BTK inhibitor ibrutinib is cytotoxic to myeloma and potently enhances bortezomib and lenalidomide activities through NF-kappaB. Cellular signalling. 2013;25:106–112. [PubMed] [Google Scholar]
164. Honigberg LA, Smith AM, Sirisawad M, Verner E, Loury D, Chang B, Li S, Pan Z, Thamm DH, Miller RA, Buggy JJ. The Bruton tyrosine kinase inhibitor PCI-32765 blocks B-cell activation and is efficacious in models of autoimmune disease and B-cell malignancy. Proceedings of the National Academy of Sciences of the United States of America. 2010;107:13075–13080. [PMC free article] [PubMed] [Google Scholar]

also impacts on ME response,
Disruption of in vivo Chronic Lymphocytic Leukemia Tumor–Microenvironment Interactions by Ibrutinib – Findings from an Investigator-Initiated Phase II Study
Carsten U. Niemann, Sarah E.M. Herman, Irina Maric, Julio Gomez-Rodriguez, Angelique Biancotto, Betty Y. Chang, Sabrina Martyr, Maryalice Stetler-Stevenson, Constance M. Yuan, Katherine R. Calvo, Raul C. Braylan, Janet Valdez, Yuh Shan Lee, Deanna H. Wong, Jade Jones, Clare Sun, Gerald E. Marti, Mohammed Z.H. Farooqui and Adrian Wiestner
DOI: 10.1158/1078-0432.CCR-15-1965 Published April 2016


SYK inhbition: 
SYK kinase belongs to the SYK/ZAP70 family of non-receptor kinases, and activates signaling pathways downstream of the BCR, chemokine and integrin receptors, suggesting that SYK participates in tissue homing and retention of activated B cells [199]. Fostamatinib (FosD, R788) is an orally available inhibitor of SYK, which induced partial responses in relapsed CLL patients in phase I/II study [200], but further development of this drug focused on rheumatoid arthritis [201]. Additional SYK-specific inhibitors are under development, including GS-9973, which has been tested alone [202] or in combination with idelalisib, and PRT-2070.

### Drug resistances in CLL
incl ME -mediated and ibrutinib - IL4 

### BET inhibitors, and their potential for the treatment of leukemia

## Background to the approache used in this thesis
### Ex-vivo drug sensitivity screens

### Studies of the microenivornment, and attempts to replicate the impact of the microenvironment 
cancer biology is integratively influenced by cell-intrinsic and cell-extrinsic features<- evidence, and jsutify why we need to knwo more about this 
### ATAC sequncing as a proxy for TF activity justification {#intro-ATACseq}

### Measurement of outcomes? 
LDT, TTT, TFT, OS 

[@Kipps2017]
Currently, the most reliable prognostic models are those developed for treatment-free survival, as evolving treatments have yet to change the indications for therapy. Predictive models to define overall survival with a given type of therapy are challenged by the chronicity of CLL and the fact that patients often receive serial treatments, each of which can affect outcome; moreover, death might be due to an indirect or unrelated cause. Furthermore, treatment options are changing, with newly identified, highly effective agents that are clearly prolonging survival and have activity among patients who would have been considered high risk when the only option was conventional chemotherapy.

[@Kipps2017]
Historically, a favourable response to therapy has been defined as a partial remission or complete remission. Partial remission requires a 50% reduction in tumour bulk (for example, lymphadenopathy and splenomegaly), a 50% reduction in lymphocytosis, and platelet counts of >100,000 cells per µl (or 50% improvement over baseline) or a haemoglobin level of >11 g per dl (or 50% improvement over baseline) without requiring transfusions or exogenous growth factors107. Complete remission requires the normalization of blood counts, resolution in lymphadenopathy and splenomegaly, and normal marrow function. The use of CT to assess response in CLL is becoming more common, particularly in clinical trials. However, the benefit of using repeated CT scans to monitor disease is uncertain, and seems unlikely to change patient outcome. Because of the distinct pattern of response observed with BCR inhibitors, a new response category, namely, partial response with lymphocytosis, has been described. Partial response with lymphocytosis is defined as a >50% reduction in lymphadenopathy and splenomegaly, with persistent lymphocytosis; often the blood lymphocyte counts are equal to or greater than those observed prior to therapy.

In clinical trials, it is becoming more common to evaluate for MRD with 0.01% of CLL cells among the total population of mononuclear cells in the blood or marrow. MRD can be measured by flow cytometry or PCR with next-generation sequencing of the clonal immunoglobulin variable region gene rearrangements163. In most clinical trials for patients with CLL, particularly those conducted in Europe, evaluation of MRD has been performed by flow cytometry of mononuclear cells from the marrow aspirate (the preferred method) or from the peripheral blood. In the 6 months following anti-CD20 monoclonal antibody treatment, the assessment of MRD is more sensitive on the mononuclear cells of the marrow aspirate than on cells that are isolated from the blood, which will often lack detectable CLL even when they are readily found in the marrow. Beyond a complete response, the best predictor of long-term PFS and overall survival is the achievement of a complete remission without evidence for MRD.

### Multiomics datasets to study CLL 
The power of multi-omics approaches especially the use of ATAC sequencing and combining ATAC and RNA
 
### Mathematical modelling {#intro-multivariate-modelling}
Mathematical background in particular, lasso-regularised regression, and the caveats associated with feature selection
wrt cluster assignment 

Also use of bliss and other interactin methods? 

use method, explain how it works, use line plots: https://stattrek.com/multiple-regression/interaction.aspx
we need mor ltieracy about these things 
discuovery of mulitple layers of biology rather thatn jsut using t tests

how this can be used to build classfiers


caviats - screenign approach, wont find everything, rigid (as in drug - cyt -gene interactions), if two things are coerrelated, one could be srhuk to ze or both share weighting 
### CLL as a model for studying tumour biology
it makes a good model to dissect interply between extrinc triggers and internal lesions: 
https://www.sciencedirect.com/science/article/pii/S1044579X16300311?casa_token=IfC4bm0pVrYAAAAA:AaCfeETGlSJRllAPIzQOcv8a48vvnFmr9V125WI6_9ggFeUTbq3dr-1628CN3NXw3li9oQlgVw

NF-κB activation in chronic lymphocytic leukemia: A point of convergence of external triggers and intrinsic lesions

## Aims of this thesis

## Overview of the approach
Including diagrams - drug targets and cytokine targets and sources 

## Note on contributions
Many of the figures, adn their cpations are taken directly from the amunscript. Where this si the case, it is indicated in the figure caption. 


Reading list before defence0: 
