# Introduction

```{r loadLibraries, echo = FALSE}
library(ggplot2)
```

(theoretical background)
# CLL Disease biology of CLL

## Disease Characteristcs
@JCIpaper shows that...  [@JCIpaper].

## Disease biology
### IGHV status

### The role of BCR signalling 

###Recurrent genetic features in CLL  
-mutations

-copy number variants

### The incompletely understood role of trisomy12 

### Stratification of CLL patients

# The tumour microenvironment in CLL 
## The concept of the protective niche 
Lymph nodes etc

## Microenvironmeental pathways and cross-talk 
pathways esp IL4 and TLR    
The microenvironment in CLL constitutes cross talk via soluble factors and cell-cell contacts, in the blood, bone marrow and lymph nodes. \@ref(fig:microenvironmentOverview)
```{r microenvironmentCrosstalk, fig.cap= "Summary of microenvironmental in CLL", echo = FALSE}

#knitr::include_graphics("figures/microenvironmentCrosstalk.eps")

```
_Figure adapted from ..._ 


```{r microenvironmentTissues, fig.cap= "The tissues that constitute the microenvironment in CLL", echo = FALSE}

#knitr::include_graphics("figures/microenvironmentTissues.eps")

```
_Figure adapted from ..._ 

# Therapies in CLL 
## Clinical regimes 

## New Targeted therapies
in particular Ibrutinib and idelalisib  

## Drug resistances in CLL
incl ME -mediated and ibrutinib - IL4 

## BET inhibitors, and their potential for the treatment of leukemia

# Background to the approache used in this thesis
## Ex-vivo drug sensitivity screens

## Studies of the microenivornment, and attempts to replicate the impact of the microenvironment 

## Measurement of outcomes? 
LDT, TTT, TFT, OS 

## Multiomics datasets to study CLL 
The power of multi-omics approaches especially the use of ATAC sequencing and combining ATAC and RNA
 
## Mathematical modelling 
Mathematical background in particular, lasso-regularised regression, and the caveats associated with feature selection
wrt cluster assignment 

## CLL as a model for studying tumour biology

# Aims of this thesis

# Overview of the approach
Including diagrams 
