# Introduction {#introduction}

```{r loadLibraries, echo = FALSE}
library(ggplot2)
```

## CLL
### Disease characteristics {#intro-disease-characteristics}
Chronic lymphocytic leukaemia (CLL) is a malignancy of mature B cells, characterised by the progressive accumulation of malignant lymphocytes in the blood, bone marrow and lymph nodes [@Kipps2017]. The malignant cells can be distinguished by expression of CD5, CD19, and CD23, and by lower levels of membrane IgM, IgD, and CD79B [@Matutes1994; @Moreau1997; @Ternynck1974; @Chiorazzi2005] (Figure \@ref(fig:CLLmorphology)).

The disease pathogenesis is driven by multiple factors, including molecular features, signalling via the B cell receptor (BCR) and interactions with non-neoplastic cells within the lymphoid tissues (known as the tumour microenvironment) [@Kipps2017].  CLL cells accumulate due to a defect in apoptosis [@Schena1992; @Kitada1998] and the majority are in a resting state [@Messmer2005; @Defoiche2008]. Evidence suggests that CLL cells migrate towards lymph nodes where they form proliferation centres [@Granziero2001; @Messmer2005], similar to germinal centres in healthy lymph nodes, which can show a daily birth rate of up to 3.3% of the tumour [@Herndon2017]. 

ref(fig:CLLmorphology) Caption from Kipps _Figure from @Kipps2017_. 

```{r CLLmorphology, fig.cap='(ref:CLLmorphology)', out.width="60%", echo = FALSE}

knitr::include_graphics("figures/CLL_morphology.eps")

```

The disease is the most common leukaemia in the West, accounting for 37% of leukaemia cases and ~19,000 newly detected cancers in the US in 2016 [@Kipps2017;@Dubois2020]. The risk of developing CLL is twice as high for men than women, and more likely with increasing age [@Siegel2012; @Nabhan2014; @Li2015; @Pulte2015]. Chemotherapy [@Robak2005; @Chang2012; @Lukenbill2013] and chemoimmunotherapy [@Hallek2010; @Goede2014; @Hillmen2015] has been the mainstay of therapy for many years. More recently, the central role of BCR signalling in disease pathogenesis has been appreciated and new drugs targeting this pathway have improved patient outcomes.  The overall 5-years relative survival for CLL is 84% [@Miller2019]. 

In spite of this, CLL is widely considered to be incurable [@Bosch2019], and treatment regimens remain harsh. Moreover, the disease is characterised by its clinical heterogeneity [@Miller2019] and there is significant variation in survival amongst patients. Some patients harbour indolent disease, and can be followed by a watch-and-wait approach, sometimes going decades without requiring treatment. Others may need immediate treatment and survival can be short [@Miller2019]. There is a clinical need to understand the underlying biology of the disease, with a view to understanding the causes of this clinical heterogeneity and to improving patient outcomes and experiences. 

### Cell of Origin {#intro-cell-of-origin}
Identifying the cell of origin of a cancer can help explain disease pathogenesis, and understand the different subtypes, and their associated prognosis. Various studies have established mature CD5+ B cells as the cell of origin in CLL [@Campo2017; @Bosch2019]. 

In healthy tissues, these mature B cells are derived from haematopoietic stem cells in the bone marrow [@Kondo2010; @Fischer2020]. These stem cells develop in multiple stages, where each stage is defined by rearrangements within the immunoglobulin heavy chain and light chain loci [@Pelanda2012] that encode components of the BCR. The aim of this process is to generate a large BCR repertoire, capable of recognising a range of foreign antigens. 

Developing B cells in the bone marrow undergo positive and negative selection, ensuring that each BCR binds effectively to foreign antigen, whilst eliminating those that bind strongly to self-antigen [@Lebien2008; @Martensson2010]. These immature B cells then migrate to the lymph nodes and spleen where they differentiate into mature cells that are considered to be antigen “na$\"i$ve”[@Chung2003].

B cells are later activated when they encounter their respective antigen. When activated, the B cells form germinal centres, which are specialised microenvironments within the lymph node that facilitate extensive proliferation. Here they undergo a process called affinity maturation, in which the loci encoding components of the BCR undergo somatic hypermutation to optimise BCR antigen specificity [@Shlomchik2012]. This process generates short-lived plasmablasts which provide immediate protection, along with plasma cells and memory B cells procuring longer-term immunity [@Nutt2015]. 

### IGHV status {#intro-ighv-status}
The two major subtypes of CLL can be defined by whether the B cell of origin has undergone this process of somatic hypermutation within a germinal centre. This is reflected in the degree of mutation of the immunoglobulin heavy-chain variable region (IGHV) genes. Evidence suggests that IGHV-mutated (IGHV-M) CLLs (60% of cases) are derived from antigen-experienced B cells. On the other hand, there is continued debate as to whether IGHV-unmutated (IGHV-U) CLLs derivate from na$\"i$ve B cells that have yet to transit through the germinal centre, or germinal-centre, antigen-experienced B cells [@Klein2001] (Figure \@ref(fig:IGHVstatus)). 

The two subtypes show distinct clinical and molecular properties. In contrast to IGHV-M CLLs, IGHV-U CLLs have a higher proportion of high-risk genetic lesions, more commonly undergo clonal evolution and consequently have a shorter time to first treatment (TTFT) and less favourable overall survival (OS) [@Hamblin1999; @Damle1999; @Landau2013; @Puente2015; @Bosch2019]. Their differential antigen experience also affects how they respond to signals from the microenvironment, including stimulation of the BCR. 

(ref:IGHVstatus) Taken from Kipps et al 2017: "Normal naive B cells that have undergone successful V(D)J recombination and express functional B cell receptors that are capable of binding to antigen interact with CD4+ T cells and accessory cells, which aggregate to form follicles that become germinal centres. Germinal cells each have a dark zone, comprising rapidly dividing B cells, and a light zone, comprising B cells mixed with follicular dendritic cells (FDCs), macrophages and helper T cells (TH cells). The B cells enter the dark zone of the germinal centre where they experience rapid proliferation and somatic hypermutation (SHM) in the genes encoding the immunoglobulin variable regions of the heavy chain (IGHV) and the light chain (IGVL). As they pass through to the light zone, the B cells that express the fittest B cell receptors for binding antigen are selected and may undergo immunoglobulin class-switch recombination. Chronic lymphocytic leukaemia (CLL) cells that use unmutated IGHV apparently originate from CD5+ B cells prior to experiencing SHM, whereas CLL cells that use mutated IGHV most likely originate from CD5+ B cells that have passed through and differentiated in the germinal centre. Some CLL cells might be derived from B cells that also have undergone immunoglobulin class-switch recombination and express immunoglobulin isotypes other than IgM and IgD, for example, IgG or IgA. Another subset is one with CLL cells that express immunoglobulin with only modest somatic mutations, such as CLL cells that use IGHV3-21 with ~97% homology to the inherited IGHV3-21 gene and an immunoglobulin light chain encoded by an unmutated IGLV3-21; these cells might derive from a B cell that has had constrained SHM, possibly owing to a limited need for immunoglobulin somatic diveresification and selection. Dashed arrows indicate speculated pathways." _Figure from @Kipps2017_. 

```{r IGHVstatus, fig.cap='(ref:IGHVstatus)', out.width="60%", echo = FALSE}

knitr::include_graphics("figures/IGHV_status.eps")

```



### The central role of BCR signalling {#intro-bcr-signalling}

The BCR pathway is central to the process of selection, development, proliferation and survival of CLL clones [@Chiorazzi2003; @Stevenson2004; @Agathangelidis2012; @Burger2012; @Iacovelli2015; @Dubois2020]. Evidence of the importance of this pathway comes from the observation that BCRs of CLL cells show a biased selection of IGHV and IGLV$\kappa$/$\lambda$ genes. This generates BCRs that are remarkably similar across patients [@Kipps1989; @Fais1998; @Widhopf2004; @Messmer2004; @Stamatopoulos2007],  suggesting that BCR binding to certain antigens may drive selection and proliferation of CLL clones. Other compelling studies have shown that BCR signalling genes are upregulated in CLL cells taken from lymph nodes [@Herishanu2011] and that cells from IGHV-U CLL patients (with poorer outcomes), also show activation of BCR-related genes [@Rosenwald2001]. The success of BCR inhibitors in the clinic [@Byrd2013], also underline the critical importance of this pathway in CLL.

Figure \@ref(fig:BCRsignalling) shows a schematic of the BCR pathway itself. The BCR is a multimeric complex consisting of surface immunoglobulin (Ig), which recognises antigen, plus the Ig-$\alpha£/Ig-$\beta$ hetero-dimers, known as CD79A and CD79B. The BCR may bind to external antigens within the microenvironment [@Binder2010] or intra-BCR self-antigens 
[@Minden], which in turn recruits SYK and the Src kinase LYN. These kinases phosphorylate motifs located on the cytoplasmic tails of CD79A and CD79B, initiating a signalling cascade involving a number of proteins and pathways. These include the proteins BTK [@Herman2011] and PI3K [@Longo2006] which activate a number of downstream pathways and players including PLC-$\gamma$2, calcium signalling, PKC, NF$\kappa$B signalling, ERK and MAPKs, and nuclear transcription [@Burger2012; @Dubois2020]. 

There are two types of BCR signalling in healthy and CLL B cells: ligand-dependent “active” signalling that relies on antigen-binding, and ligand-independent “tonic” signalling [@Lam1997; @Kraus2004]. Whilst active signalling engages the entire signalling cascade described above, tonic signalling activates only a subset of these. @Kraus2004 showed that tonic signalling is important for prolonged B cell survival whereby PI3K signalling is thought to play a key role in delivering survival signals [@Srinivasan2009; @Pula2019]. Both these modes of BCR signalling are believed to impact of the survival and growth of the tumour, although the dominant mode remains a matter of debate [@Burger2012]. 

The different clinical properties of IGHV-M and IGHV-U cell is believed to be determined by in part by their differential response to BCR stimulation, which is influenced by their cell of origin [@Burger2012]. IGHV-U CLLs have not undergone somatic hypermutation and consequently express low-affinity BCRs that are frequently activated by numerous antigens and auto-antigens in the microenvironment [@Borche1990; @Broker1988; @Sthoeger1989; @Herve2005; @Lanemo2008; @Chu2008; @Binder2010; @Krysov2010; @Kostareli2012; @Hoogeboom2012]. In contrast, IGHV-M BCRs only recognise highly specific antigens, which either occur infrequently or induce anergy due to high-affinity binding [@Chiorazzi2003; @Stevenson2004; @Chiorazzi2005; @Burger2012]. IGHV-M CLL clones therefore are more stable and expand at a slower rate. 


\@ref(fig:BCRsignalling) Adapted from: Kipps et al 2017: "B cell receptor (BCR) signalling is initiated by SRC-family kinase-dependent phosphorylation (mainly LYN) of CD79A and CD79B that creates a docking site for the binding and activation of spleen tyrosine kinase (SYK). SYK then triggers the formation of a multi-component ‘signalosome’, comprising Bruton tyrosine kinase (BTK), AKT, phosphoinositide 3-kinase (PI3K), phospholipase Cγ2 (PLCγ2) and B cell-linker protein (BLNK), among others. CD19 is a co-receptor for BCR and is important for PI3K activation, which recruits and activates PLCγ2, BTK and AKT. PLCγ2 generates diacylglycerol (DAG) and inositol-1,4,5-trisphosphate (Ins(1,4,5)P3), which triggers Ca2+ release from the endoplasmic reticulum, leading to the activation of the MEK–extracellular signal-regulated kinase (ERK) and nuclear factor-κB (NF-κB) signalling pathways. Other effects of BCR signalling include activation of mechanistic target of rapamycin complex 1 (mTORC1) and of Rho-family GTPases, RAC1 and RHOA, which can affect the cytoskeleton. Inhibitors of SYK, PI3K and BTK are shown. Note that this figure describes the main molecules and interactions that are involved in positive BCR signalling, but is not an exhaustive description of all signalling pathways or molecules activated. IKK, IκB kinase; PKC, protein kinase C."

```{r BCRsignalling, fig.cap='(ref:BCRsignalling)', out.width = "60%",  echo = FALSE}

knitr::include_graphics("figures/BCR_signalling.eps")

```



### Recurrent genetic features in CLL {#intro-genetic-features}
The genomic landscape of CLL has been thoroughly characterised in a number of studies, including two seminal papers involving >500 CLL samples [@Landau2015; @Puente2015]. Such studies have indicated that CLL shows a lower mutational load than other lymphoid neoplasms [@Vogelstein2013; @Alexandrov2013; @Bosch2019], in which a relatively large number of genes are rarely mutated [@Fabbri2016]. A small number of driver genes are mutated in a significant proportion of cases, though a common genetic event which can accounts for most cases of CLL has not been identified [@Fabbri2016]. These genetic alterations encompass chromosomal alterations, mutations, alterations in miRNA expression and epigenetic modifications [@Kipps2017].


#### Somatic Mutations {#intro-mutations}
There are many recurrent somatic mutations in CLL, and these centre on several major pathways and functions that are frequently altered in patients (Figure \@ref(fig:CLLmutations)). [@Puente2015; @Fabbri2016; @Kipps2017]. These pathways include Notch signalling, DNA damage response, RNA processing, Nf$/kappa$B signalling, BCR signalling, WNT signalling and chromatin modification [@Kipps2017; @Bosch2019]. 

Within these pathways, most mutations are rare and only a few occur at a frequency >5% [@Bosch2019]. One landmark study indicated the most frequent mutations occur in *NOTCH1* (12.6% of patients), *ATM* (11%), *BIRC3* (8.8%) and *SF3B1* (8.6%), although these frequencies are dependent disease stage and treatment status [@Puente2015]. The functional role and prognostic importance of a number of these putative driver mutations has been established, and are discussed below. 

**Notch signalling**
The Notch pathway activates genes required for proliferation, metabolism and survival, including *MYC*, via the activation of the NOTCH1 transmembrane receptor [@Guruharsha2012]. Mutations in *NOTCH1* are very common in CLL (~4-20%) [@Fabbri2011; @Puente2011; @Landau2013; @Puente2015; @Landau2015]. A number of other recurrent mutations also centre on Notch deregulation, indicating that this pathway is disrupted in many CLL cases [@Fabbri2016]. *NOTCH1* mutations occur more often in IGVH-U CLLs and are associated with a less favourable OS [@Fabbri2011]. *NOTCH1* mutants also respond less-well to anti-CD20-based therapies, due to decreased surface expression of CD20 [@Fabbri2001; @Rossi2012]. 

**DNA Damage Response**
A number of frequently occurring mutations disrupt the DNA damage response, the most clinically important of which is *TP53*. *TP53* is known as the “guardian of the genome”, owing to its role as a tumour suppressor gene protecting genome integrity by preventing mutation. Despite being a tumour suppressor, *TP53* mutations can have a dominant negative effect on function, such that loss of genomic stability occurs even when a single allele is mutated [@Zenz2008].  *TP53* is an important prognostic marker, as dysregulation of its function is associated with resistance to DNA-damaging agents (chemotherapy and radiotherapy) [@Zenz2008; @Dicker2009; @Rossi2009], and patients who present with mutations at diagnosis often show shorter TTFT and a less favourable OS [@Dohner2000; @Zenz2008; @Rossi2013; @Puente2015]. 

Upstream of *TP53*, the tumour suppressor *ATM* is also frequently mutated in CLL [@Bosch2019]. *ATM* activates the DNA damage response upon recognition of DNA double strand breaks [@Austen2005; @Shiloh2013]. Similarly to *TP53* mutants, *ATM*-disrupted CLLs also show genomic instability and these mutations are associated with shorter TTT and OS, and chemoresistance [@Austen2005; @Stankovic2014]. 

*POT1* is also involved in genomic stability, and is mutated in around 3 – 7% of cases [@Bosch2019; @Ramsay2012]. *POT1* mutations disrupt telomere protection, leading to a increased in structural aberrations and chromosomal breaks [@Ramsay2012]. Mutations in *POT1* are associated with IGHV-U CLL and advanced clinical stage [@Ramsay2012].

**RNA processing**
30% of CLL patients harbour mutations disrupting RNA processing and the spliceosome machinery [@Puente2011; @Wang2011; @Quesada2012; @Puente2015; @Fabbri2016].
*SF3B1* mutations are the most frequent to occur, and are found in ∼10% of cases, usually in IGHV-U patients [ @Puente2011; @Wang2011; @Quesada2012; @Puente2015; @Fabbri2016]. The *SF3B1* gene encodes part of the U2 snRNP complex which is involved in RNA splicing [@Shin2004], though the functional implication of *SF3B1* mutations are yet to be established, and many transcripts show abnormal splicing in*SF3B1 cases [@Quesada2012]. The presence of *SF3B1* mutations is associated with a decreased TTFT and unfavourable OS [@Bosch2019].

**NF$\kappa$B signalling**
A wide range of mutations across various pathways converge on the activation of NF$\kappa$B [@Fabbri2011; @Puente2011; @Wang2011; @Quesada2012; @Puente2015; @Landau2015], including *BIRC3* and *MYD88*. For example, certain mutations in *MYD88* result in increased binding to IRAK1 and higher activation of NF$\kappa$B [@Puente2011]. However, the role and prognostic importance of NF$\kappa$B activation in CLL is still unclear [@Bosch2019].

(ref:CLLmutations) Taken from Kipps et al 2017: "Genes that are mutated in chronic lymphocytic leukaemia (CLL) are involved in several cellular pathways (blue boxes). As such, mutations in these genes can lead to a range of cellular consequences, such as aberrant DNA repair and B cell receptor (BCR) signalling, among others51,213. The minus sign from GBN1 to the MAPK–ERK pathway indicates negative regulation. *For more detail of the BCR and its associated signalling, see FIG. 3. ASXL1, additional sex combs-like protein 1; ATM, ataxia telangiectasia mutated; BAZ2A, bromodomain adjacent to zinc-finger domain protein 2A; BCOR, BCL-6 co-repressor; BIRC3, baculoviral IAP repeat-containing protein 3; BRCC3, BRCA1/BRCA2-containing complex subunit 3; C-NOTCH, carboxy-terminal domain of NOTCH; CARD11, caspase recruitment domain-containing protein 11; CHD2, chromodomainhelicase-DNA-binding protein 2; CHK2, checkpoint kinase 2; Co-A, co-activator; CSL, CBF1–Suppressor of Hairless–LAG1 (also known as RBPJ); DDX3X, ATP-dependent RNA helicase DDX3X; DYRK1A, dual-specificity tyrosine-phosphorylation-regulated kinase 1A; EGR2, early growth response 2; ELF4, ETS-related transcription factor Elf-4; ERK, extracellular signal-regulated kinase; EWSR1, Ewing sarcoma breakpoint region 1 protein; FBXW7, F-box/WD repeat-containing protein 7; FUBP1, far upstream element-binding protein 1; GNB1, guanine nucleotide-binding protein G(I)/G(S)/G(T) subunit β1; H3K4, histone H3 lysine 4; IC, intracellular domain; IKZF3, Ikaros family zinc-finger protein 3; IL-1R, IL-1 receptor; IRF4, interferon regulatory factor 4; ITPKB, inositol-trisphosphate 3-kinase B; LRP, low-density lipoprotein receptor-related protein; MAP2K1, dual-specificity mitogen-activated protein kinase kinase 1; MAPK, mitogen-activated protein kinase; MED12, Mediator of RNA polymerase II transcription subunit 12; MGA, MAX gene-associated protein; MYD88, myeloid differentiation primary response protein MyD88; NF-κB, nuclear factor-κB; NXF1, nuclear RNA export factor 1; P, phosphate; POT1, protection of telomeres protein 1; PTPN11, tyrosine-protein phosphatase non-receptor type 11; RIPK1, receptor-interacting serine/threonine-protein kinase 1; RPS15, 40S ribosomal protein S15; SAMHD1, SAM domain and HD domain-containing protein 1; SF3B1, splicing factor 3B subunit 1; SHP1, Src homology region 2 domain-containing phosphatase 1 (also known as PTPN6); SYK, spleen tyrosine kinase; TCF/LEF, T cell factor/lymphoid enhancer factor; TLR8, Toll-like receptor 8; TNFR1, tumour necrosis factor receptor 1 (also known as TNFRSF1A); TRAF, TNFR-associated factor; XPO, exportin; ZMYM3, zinc-finger MYM-type protein 3. Adapted with permission from REF. 51, Macmillan Publishers Limited." _Figure from @Kipps2017_

```{r CLLmutations, fig.cap='(ref:CLLmutations)', out.width="60%", echo = FALSE}

knitr::include_graphics("figures/genetic_mutations.eps")

```

#### Structural Aberrations {#intro-structural-abberations}
In addition to the aforementioned mutations, a number of common structural aberrations confer similar disruption to normal B cell function. The key structural aberrations and their prognostic value were set out in a landmark paper by @Dohner2000, as follows. 

**del(11q)**
Deletions of chromosome 11q (del(11q)) are fairly common (~10%) in CLL, and are believed to target the *ATM* gene in the 11q22-23 region [@Dohner2000; @Austen2005; @Bosch2019]. del(11q) is usually monoallelic, but can also be associated with mutations in the remaining ATM allele (~30% of cases) [@Austen2005]. In certain instances, the deleted region does not include *ATM* but rather *BIRC3*, a negative regulator of the NF$\kappa$B pathway [@Rossi2012]. Patients with del(11q) or *ATM* lesions have a shorter TTFT and OS, especially if the lesion is biallelic [@Austen2005; @Skowronska2012; @Stankovic2014; @Nadeu2016]. 

**del(17p)**
Deletion of chromosomal region 17p13 (del(17p)) is found in 1 – 20% of cases, depending on the stage of the disease and most common in chemo-refractory cases [@Dohner2000; @Zenz2008; @Zenz2010; @Stilgenbauer2014]. The target of this lesion is thought to be *TP53*; the deleted region consistently includes the *TP53* locus [@Dohner1995], and around 80% del(17p) cases also have missense mutations in the second *TP53* allele [@Zenz2008; @Gonzalez2011; @Trbusek2011; @ Rossi2014]. Del(17p) CLLs show increased genomic instability [@Yu2017], resistance to chemotherapy and, correspondingly, a shorter TTFT and a less favourable OS [@Dohner2000; @Zenz2008; @Rossi2013; @Puente2015]. 
**del(13q)**
Deletion in the 13q14 region (del(13q)) is the most common genetic lesion in CLL (~50–60% of patients) [@Dohner2000]. Experiments to determine the minimal deleted region identified that this invariably contains *DLEU1* and *DLEU2*, two long non-coding RNA genes, and the microRNA gene cluster *MIR15A–MIR16-1* [@Kalachikov1997; @Migliazza2001; @Calin2002; @Hammarsund2004; @Palamarchuk2010]. *in vitro* studies have demonstrated the role of these genes in regulation of the cell cycle and apoptosis [@Cimmino2005; @Klein2010; @Bosch2019]. In some CLL cases, del(13q) is the sole genetic abnormality, indicating that this lesion may be involved in early CLL development [@Dohner2000; @Klein2010; @Landau2015]. Moreover, conditional deletion of the equivalent minimal deleted region in mice recapitulated CLL initiation and progression, and these mice developed clonal lymphoproliferations [@Migliazza2001; @Klein2010].  del(13q) patients have the best prognosis, with prolonged TTFT, and OS compared to patients with other lesions [@Dohner2000; @Rossi2013]. 

#### The incompletely understood role of trisomy12 {#intro-trisomy12}
Complete duplication of chromosome 12 (trisomy 12) is observed in ~15% of CLL patients at diagnosis [@Dohner2000]. Despite its recurrence, there is currently no functional explanation for this lesion [@Bosch2019], although a number of features have been observed. Trisomy 12 is more common in IGHV-M patients than IGHV-U [@Hamblin1999], and is thought to confer an abnormal cellular morphology [@Bosch2019]. Previous work in our lab has also demonstrated that trisomy 12 CLLs show a specific signalling signature and distinct transcriptomic [@JCIpaper] and proteomic profiles [@HerbstThesis], including differential expression of genes within the BCR, PI3K, AKT, and mTOR signaling and chemokine signaling pathways. Moreover, trisomy 12 CLLs show higher sensitivity to BCR inhibitors, indicating that BCR signalling may be amplified in these cases [@JCIpaper]. 

Traditionally trisomy 12 has been classified as an intermediate-risk lesion [@Dohner2000]: these cases have a higher proliferative capacity, but are more treatable with chemotherapeutics and BCR inhibitors. However, *NOTCH1* mutations are frequently observed in trisomy 12 cases, and this is associated with poorer outcomes [@Balatti2012; @Giudice2012]. 
  
#### Epigenetic alterations {#intro-epigenetic-alterations}
In addition to genetic lesions, the epigenome is also modified in CLL and samples typically show global hypomethylation combined with local hypermethylation [@Wahlfors1992; @Cahill2013; @Ziller2013; @Kipps2017; @Bosch2019]. Studies investigating the epigenome in CLL have proven revealing, in particular, higher levels of intra-sample methylation heterogeneity have been shown to be associated with high-risk genetic lesions and poorer prognosis [@Landau2014]. 

Moreover, methylation signatures have been used to classify distinct clinical CLL subgroups [@Kulis2012; @Bhoi2016], as they are useful to trace the cell of origin. For example, CLL cells from distinct patients originate from many different B cell maturation states, possibly reflecting the biological and phenotypic heterogeneity of CLL [@Oakes2016]. Accordingly, IGHV-U CLLs have a distinct methylation signature to IGHV-M CLLs, and these patterns correspond approximately to those of pre-germinal centre or post-germinal centre memory B cells, respectively [@Kulis2012; @Oakes2016]. Epigenetic studies have also revealed the relationship between certain genetic lesions and specific epigenetic signatures, for example *MYD88* mutations and trisomy 12 [@Beekman2018].  The CLL epigenome can also be modulated by drugs and thus is of increasing clinical interest [@Timp2013; @Beekman2018; @Gaiti2019].

## Therapies in CLL {#intro-therapies}
Extensive work to uncover the molecular drivers of CLL has led to the development of several therapeutic strategies. CLL represents a successful example of how developing a complex understanding of the biological characteristics of a disease can lead to significantly improved patient outcomes [@Yosifov2019]. 
Treatment of CLL patients can be via chemotherapy, chemoimmunotherapy or targeted therapies that inhibit specific pathways [@Kipps2017; @Burger2020]. Additionally, allogeneic stem cell transplantation is increasingly considered as an alternative option in relapsed or refractory patients [@Kipps2017].  

### Chemotherapy and chemoimmunotherapy {#intro-chemotherapy}
Chemotherapy has been the standard of care for CLL for many decades, either with purine analogues such as fludarabine or alkylating agents such as chlorambucil [@Robak2005; @Lukenbill2013]. Chemoimmunotherapy has also benefited many patients owing to the rapid improvements to monoclonal antibody technology and the development of anti-CD20 treatments, such as rituximab [@Yosifov2019; @Robak2010]. However, patients with higher risk lesions such as *TP53* and del(17p) do not respond well to chemoimmunotherapy, and require alternative therapeutic options [@Zenz2010].


### Targeted therapies {#intro-targeted-therapies}
More recently, the importance of BCR signalling and upregulation of anti-apoptotic proteins in CLL expansion has been increasingly realised and led to the development of therapies targeting these pathways. Drugs targeting BCR signalling and BCL-2 have changed the treatment landscape dramatically [@Scheffold2020]. Three main drug classes that target BCR signalling have been developed for CLL: BTK inhibitors, PI3K inhibitors and SYK inhibitors [@Rooij2012; @Kipps2017]. 

Ibrutinib is a BTK inhibitor approved for use as an initial therapy and for patients who are refractory to chemoimmunotherapy [@Byrd2013; @Byrd2014]. Despite such success, complete remission is rare and many patients continue to harbour minimal residual disease within the bone marrow [@Byrd2013], requiring continued therapy for years [@Byrd2013; @Woyach2015]. Resistance can also occur via the acquisition of mutations in BTK or PLC$\gamma$2 [@Woyach2015] genes. Treatment initiation with ibrutinib is associated with a concomitant increase in the absolute lymphocyte count in the blood [@Woyach2014], thought to caused by the inhibition of chemokine receptor signalling leading to the release of malignant B cells from the lymph nodes into the peripheral blood. 

Idelalisib also acts to inhibits BCR signalling and chemokine signalling [@Hoellenriegel2011], via inhibition of PI3K. Whilst the drug is highly efficacious in CLL [@Furman2014; @Brown2014], idelalisib demonstrates more toxicities and lower efficacy than BTK inhibitors and thus is generally used as an alternative therapy in patients for whom BTK inhibitors are unsuitable [@Ghia2020; @Burger2020].

Other drugs targeting SYK, downstream of BTK, have also shown promise in Phase I/II clinical trials [@Friedberg2010], though none are licenced as yet.

Aside from BCR inhibitors, BCL-2 inhibitors such as venetoclax, have also shown good efficacy in CLL. Venetoclax is thought to induce apoptosis in CLL cells, by acting as a BH3 mimetic interfering with the ability of BCL-2 to sequester BIM [@Moore2007]. This makes venetoclax an attractive alternative in patients with relapsed or refractory disease [@Roberts2016], or who have del(17p) i.e. loss of *TP53 *[@Stilgenbauer2016]. Complete remissions are observed in ~20% patients [@Roberts2016] which is higher than with other targeted therapies, and minimal residual disease in the bone marrow is less common [@Roberts2016].

### Management algorithm {#intro-management-algorithm}
The decision to initiate therapy is guided by the stage of disease, evidence for rapid disease progression or disease-related symptoms [@Hallek2008; @Kipps2017]. Disease stage is determined by an index such as the Rai staging system, which categorises patients based on disease severity [@Kipps2017]. The choice of therapy accounts for certain mutations, in particular the presence of *TP53* or del(17p) mutations, the age of the patient and the objective of therapy. IGHV status is also increasingly used for patient stratification (Figure \@ref(fig:CLLstratification,  [@Kipps2017]).

CLL is a manageable disease, with a well-established arsenal of treatments and associated management algorithm. However, treatments are harsh, and the disease is still considered incurable [@Bosch2019]. Many cases develop resistance to therapy, both via acquired mutations and through survival signals provided by the microenvironment (see section @\ref(intro-microenvironment-drug-response)) and minimal residual disease within the bone marrow is common. Collectively, these lead to relapse or required prolonged therapy and its associated toxicities. There remains a clinical need to improve patient outcomes and experience.

(ref:CLLstratification) Taken from Kipps et al 2017: "Indications for therapy of patients with chronic lymphocytic leukaemia (CLL) include late-stage disease, evidence for rapid disease progression or disease-related symptoms. Patients with del(17p) or mutated TP53 should be treated with therapy that does not require functional TP53, such as ibrutinib (a Bruton tyrosine kinase (BTK) inhibitor), given the relatively poor outcome for such patients with chemotherapy. For patients without del(17p) or known mutations in TP53, immunoglobulin heavy-chain variable region (IGHV) mutational status can help to define the treatment strategy; patients with unmutated IGHV could be considered for therapy with a BTK inhibitor (such as ibrutinib) and patients with mutated IGHV might be good candidates for chemoimmunotherapy (CIT), if amenable. Indeed, patients with mutated IGHV can have excellent outcomes with CIT regimens, such as fludarabine, cyclophosphamide and rituximab, with >50% of patients having a median progression-free survival of >10 years, including the potential for cure. If the patient is amenable to CIT, age, medical comorbidities and myeloid reserve should be taken into consideration. Patients >65 years of age commonly have medical comorbidities and are less able to tolerate myelosuppressive regimens, such as fludarabine, cyclophosphamide and rituximab. Thus, considerations should be given to using reduced dose or less myelosuppressive chemotherapy regimens, such as chlorambucil or reduced-dose bendamustine and an anti-CD20 monoclonal antibody for patients with limited myeloid reserve. Patients who either do not respond, have a poor tolerance to CIT or relapse following CIT, should be re-evaluated for del(17p) or TP53 mutations. Patients who develop de novo del(17p) or TP53 mutations, or have known del(17p) and/or TP53 mutations, or who develop resistance or intolerance to ibrutinib, could be considered for therapy with idelalisib and rituximab or the BCL-2 inhibitor venetoclax. Patients treated with CIT who do not have del(17p) or TP53 mutations could be considered for repeat CIT if their progression-free survival after CIT is >2 years and the patient has sufficient myeloid reserve. Such patients also might be treated with a BTK inhibitor or a phosphoinositide 3-kinase (PI3K) inhibitor, which also could be considered for patients who develop intolerance or resistance to therapy with ibrutinib. Patients who develop resistance or intolerance to inhibitors of BTK, PI3K and/or BCL-2 should be considered for clinical trials or alternative agents. LDT, lymphocyte doubling time." _Figure from @Kipps2017 _.whcih provides five categories, stage 0 – stage IV, where IV is high risk


```{r CLLstratification, fig.cap='(ref:CLLstratification)', out.width = '60%', echo = FALSE}

knitr::include_graphics("figures/disease_stratification.eps")

```
 

### BET inhibitors, and their potential for the treatment of leukemia (remove? )

## The tumour microenvironment in CLL {#intro-tumour-microenvironment}

### The role of the tumour microenvironment in CLL {#intro-tumour-microenvironment-role}
In addition to genetic aberrations, the tumour microenvironment is an important driver of disease pathogenesis in CLL [@tenHacken2016]. The term microenvironment encompasses the set of non-neoplastic cells within the lymphoid tissues, including the bone marrow and lymph nodes, that provide survival signals to the tumour (Figure \@ref(fig:microenvironmentTissues)), leading to clonal expansion and drug resistance [@Burger2014].  Malignant B cells engage in a dialogue with the non-neoplastic cells, via cell-cell contacts and soluble factors, including chemokines, integrins, cytokines and survival factors, centring on a number of important pathways including BCR signalling and tissue homing chemokine receptors [@tenHacken2016].

The importance of the microenvironment in CLL pathogenesis was first recognised in studies that showed CLL cells rapidly undergo apoptosis *in vitro*, whilst their survival can be extended by stimulation or by co-culture with nurselike cells (NLCs) or mesenchymal bone marrow stromal cells (BMSCs) [@Collins1989; @Burger2000; @Kurtova2009; @DeaglioMalavasi2009; @Purroy2015]. This observation indicated that the ability of CLL to progressively accumulate *in vivo* may be highly dependent on external stimulation, rather than some cell-intrinsic feature of the tumour. Building on these observations, further *in vitro* studies have shown a number of cell types and soluble factors belonging to the microenvironment are also capable of protecting the tumour cells from drugs and chemotherapeutic agents. Many CLL patients continue to harbour minimal residual disease, which eventually leads to relapse: it is believed that the tumour microenvironment provides a sanctuary for the malignant B cells to shield from the effects of therapy [@Dubois2020]. 


(ref:microenvironmentTissues) _Adapted from Davids et al 2012_

```{r microenvironmentTissues, fig.cap='(ref:microenvironmentTissues)',out.width = "60%",  echo = FALSE}

knitr::include_graphics("figures/microenvironmentTissues.eps")

```

### Components of the tumour microenvironment
**Lymph Nodes**
Over the last two decades, significant progress has been made in unravelling this complex cross-talk and many of the important cellular and molecular components have been defined and studied. Malignant B cells circulate through the blood in a resting state, and follow chemokine gradients towards the lymph nodes to form "profileration centres" (Figure \@ref(fig:proliferationCentres)), similar to germinal centres [@Herishanu2011]. Studies using detuerated water labelling have shown that up to 3% of the clone is actively proliferating within the lymph node [@Messmer2005; @Herndon2017]. Cross-talk with the non-neoplastic cells in the lymph node  shapes the transcriptomic profile of the malignant B cells, and leads to upregulation of the BCR pathway [@Herishanu2011, @Mittal2014], a central driver of CLL pathogenesis. 

(ref:proliferationCentres)  _Figure from @Kipps2017_. 

```{r proliferationCentres, fig.cap='(ref:proliferationCentres)', out.width="60%", echo = FALSE}

knitr::include_graphics("figures/proliferation_centres.eps")

```

**Bone marrow**
The bone marrow is also known to be important, and several studies have shown *ex vivo* BMSCs to protect CLL cells against the drug toxicity  [@Kay2007; @Kurtova2009]. However, gene expression changes are less pronounced within the bone marrow compared to the lymph node [@Herishanu2011].

**Cellular components**
Within these compartments, the CLL cells engage in a dialogue with mesenchymal stromal cells (MSCs), NLCs and follicular dendritic cells (FDCs), in concert with T cells, natural killer cells (NK cells) and components of the extracellular matrix [@tenHacken2016]. NLCs are of monocytic origin: their critical role in CLL was first demonstrated by the observation that peripheral blood-derived monocytes differentiate into NLCs, and that these cells prolong CLL cell survival *ex vivo* [@Burger2000]. They are also found within the lymphoid tissues of CLL patients [@Tsukada2002; @Burkle2007]. MSCs, which include BMSCs, are frequently observed within the secondary lymphatic tissues of CLL patients. Many studies have demonstrated the ability of these cells to inhibit spontaneous and drug-induced apoptosis in *in vitro* CLL co-cultures [@Panayiotidis1996; @Lagneaux1998, @Kurtova2009; @Stamatopoulos2010; @Stamatopoulos2012]. 

FDCs are important for tissue homing and retention of CLL cells within tissues. In healthy tissues, they are usually found within germinal centres [@Allen2008], and present unprocessed antigen to B cells. In CLL they play an important role within the secondary lymphoid organs, having a protective effect on CLL via cytokine secreton, adhesion molecules and the activation of BCR signalling [@Dubois2020]. CLL co-culture with FDCs leads to inhibition of spontaneous apoptosis and upregulation of anti-apoptotic MCL [@Pedersen2002]. 

The T cell compartment is also altered in CLL, first described by @Scrivener2003. T cells have been observed to have pro-tumour and anti-tumour behaviour. On the one hand,  higher numbers of CD4+ T-helper (Th) cells are seen in CLL patient blood samples [@Palma2017; @Catakovic2017; @Elston2020], and in line with this, Th cell cytokines have been shown to provide pro-survival signals *in vitro*, for example IL4 from Th2 cells [@Dancescu1992; @Bhattacharya2015; @AguilarHernandez2016]. Activated CD4+ T-cells in murine xenograft models [@Bagnara2011; @Os2013] of CLL have also been shown to increase survival and growth of the tumour. On the other hand, there is evidence of increased antigen-experienced CD8+ T cells in CLL, which control tumour growth in a CLL mouse model [@Roessner2020; @Grioni2021].


### Microenvironmental pathways {#intro-microenvironmental-pathways}
Cross-talk between these non-neoplastic cells and the malignant B cells can occur directly, via cell-cell contacts and adhesion molecules, indirectly, via soluble factors that bind to receptors on the CLL cells, or through the exchange of material held in extracellular vesicles [@Guarini2008; @Oppezzo2013; @Crompot2017]. Collectively these induce pathway activation (most importantly BCR and NF$\kappa$B [@Herishanu2011]) and gene expression modifications with the CLL cells, leading to chemotaxis, homing to lymphoid tissues and survival of the tumour cells [@Dubois2020]. Figure \@ref(fig:microenvironmentSignalling) depicts an overview of this cross-talk.

(ref:microenvironmentSignalling) Taken from Kipps et al 2017; "Migration of chronic lymphocytic leukaemia (CLL) cells into the lyphoid tissu is primarily mediated through CXC-chemokine receptor 4 (CXCR4) in response to CXC-chemokine ligand 12 (CXCL12), which is secreted mainly by nurse-like cells (NLCs) and mesenchymal-derived stromal cells. Migration of CLL cells into lymph nodes also occurs via CC-chemokine receptor 7 (CCR7) in response to CC-chemokine ligand 19 (CCL19) and CCL21, which are produced by the endothelial cells of high endothelial venules (HEVs). HEV endothelial cells also express hyaluronan, which can interact with CD44, to facilitate B cell signalling and might enhance the production of active matrix metalloproteinase 9 (MMP9). Once in tissues, several chemokines promote B cell survival, including CXCL12, B cell-activating factor (BAFF; also known as TNFSF13B) and a proliferation-inducing ligand (APRIL; also known as TNFSF13). In addition, CLL cell survival can be promoted through cognate interactions between CD31 and CD38, and the production by stromal cells of WNT factors, which can interact with ROR1, ROR2 and/or various Frizzled receptors. CLL cell contact with mesenchymal stromal cells can also be established through vascular cell adhesion protein 1 (VCAM1)– α4β 1 integrin interactions that contribute to CLL cell survival. In turn, CLL cells can secrete chemokines, such as CCL3 and CCL4, which can recruit T cells and NLC-precursor cells (monocytes) to the CLL microenvironment. Activated T cells can provide CLL cells with proliferative signals through CD40 ligand (CD40L)-CD40 interactions and the secretion of several cytokines, such as IL-2, IL-4 and IL-10. In turn, activated CLL cells secrete CCL12 and CCL22, which attract more T cells into the CLL microenvironment. In tissues, CLL cells can be exposed to environmental and/or self-antigens that might trigger B cell activation through interactions with the surface immunoglobulin; this could amplify the responsiveness of CLL cells to the signals and factors that are provided by the CLL microenvironment. BAFFR, BAFF receptor (also known as TNFRSF13C); BCMA, B cell maturation protein (also known as TNFRSF17); BCR, B cell receptor; TACI, transmembrane activator and CAML interactor (also known as TNFRSF13B). _Figure from @Kipps2017 ._ "

```{r microenvironmentSignalling, fig.cap='(ref:microenvironmentSignalling)', out.width = "60%", echo = FALSE}

knitr::include_graphics("figures/microenivironmental_signalling.eps")

```

**Cell-cell contacts**
The importance of direct contact between cells became clear with observations that the ability of MScs to provide efficient rescue from spontaneous and drug-induced apoptosis is dependent on direct contact and can be blocked by separation through a filter [@Lagneaux1998; @Burger2000; @Kay2007; @Kurtova2009; @Ding2009]. Likewise, FDCs also operate through direct contact, as evidenced by the observation that contact with HK cells (an FDC cell line) protects against CLL cells from apoptosis [@Pedersen2002].

This direct contact operates through a number of receptor-receptor interactions. For example, CLL-stromal cell binding involves $\beta$1 integrin (ITGB1, or CD29) and $\beta$2 integrin (ITGB2, or CD18) [@Lagneaux1999; @Lee2001].  VLA-4 is also an important integrin for retention of CLL cells within the lymph nodes and bone marrow, by interacting with its ligand VCAM-1 (or CD106) on stromal cells [@Burger2001]. 

These cell-cell interactions then lead to pathway activation (including BCR [@Stamatopoulos2015] and TLR [@Schulz2011]),  gene expression changes and epigenetic changes  [@Vangapandu2017; @Xu2018] within the tumour cells. For example, contact between CLL B cells and MSCs alters the transcriptomic profile of the cells [@Schulz2011; @Mangolini2018], leading to increased expression of anti-apoptotic proteins such as BCL2 [@Nwabo2012; @Patel2013], BCL-XL [@Patel2013; @AmigoJimenez2015], MCL1[@Kurtova2009; @AmigoJimenez2015], and $\beta$-catenin [@Mangolini2018]. 

**Soluble Factors**
NLCs, MSCs, FDCs and T cells also secrete soluble factors that have a protective effect on the tumour (Figure \@ref(fig:microenvironmentOverview)). For example, MSCs secrete a number of cytokines. One of the most widely studied is SDF1-$\alpha$ (or CXCL12), which interacts with CXCR4 on CLL cells [@Burger1999; @Burger2001; @Kay2007], stimulating the PI3K [@Burger2005], STAT3 [@Burger1999], and p44/42 MAPK [@Burger2000] pathways which activates BTK [@Montresor2011], ERK [@Messmer2011], and AKT [@OHayre2010].  FDCs on the other hand secrete BAFF, which has been shown to increase survival of CLL cells through the activation of canonical NF$\kappa$B signalling [@Nishio2005]. A number of Th T-cell-derived cytokines have also been shown to increase CLL viability *in vitro*, including IFN$\gamma$ [@Buschle2003], IL15 [@Trentin1996], IL21 [@deTotero2006; @Pascutti2013], IL4 [@Dancescu1992; @Bhattacharya2015; @AguilarHernandez2016],  IL2 [@Decker2010] and CD40L [@Kitada1999; @Pascutti2013; @Bhattacharya2015]. Certain soluble factors can also increase apoptosis and act against the tumour, including TGF$\beta$ [@Lotz1994]. 

### The impact of the microenvironment on drug response {#intro-microenvironment-drug-response}

### Modelling the tumour microenvironment



## Background to the approaches used in this thesis
### Ex-vivo drug sensitivity screens

### Multiomics datasets to study CLL 
The power of multi-omics approaches especially the use of ATAC sequencing and combining ATAC and RNA
ATAC sequncing as a proxy for TF activity justification {#intro-ATACseq}
PUt with section on this? 

### CLL as a model for studying tumour biology

### Measurement of outcomes? 
Put with section on this? 
 
### Mathematical modelling {#intro-multivariate-modelling}

## Aims of this thesis

(Overview of approach is in data chapter)
## Note on contributions??




