# Molecular and microenvironmental modulators of drug response in CLL {#chapter6}

Chapters \@ref(chapter4) and \@ref(chapter5) provide a systematic exploration of the effect of microenvironmental stimulation on _ex vivo_ CLL biology, and how these pathways can be modulated by genetic features. Collectively, these results underline the fact that cancer biology is determined by an integrated network of cell-intrinsic molecular aberrations and cell-intrinsic signals generated by cell-cell contacts and soluble factors. The comprehensive nature of this dataset enables the study of these multiple axis. 

In this chapter, we examine the third axis of the dataset: drug response. As well as providing protective signals that increase CLL viability and support tumour growth, the microenvironment is also believed to interfere with drug action. Several studies have demonstrated the impact of individual microenvironmental signals on drug response _ex vivo_, and is also thought to occur within the protective niche [@Ahn2018]. Guided by this observation, we next wanted to investigate the impact of each of the stimuli on responses to the panel of drugs, and how this can be further modulated by genetic features. 

^sharpen this once more 

In section \@ref(mapping-interactions), the impact of the stimuli on drug response is quantified by identifying drug - stimulus interactions. In section \@ref(mapping-genetic-modulators), the impact of gneetic features on drug response is explored. Finally, in section \@ref(drug-stimulus-gene-interactions),  multivariate modelling is apploed to integrate all three axes of the dataset, to investigate how drug - microenvironment interactions can be further modulated by genetic features. These analyses benefit from linear regression and generalised linear modelling approaches, described in more detail in section \@ref(intro-multivariate-modelling). 


```{r, echo = FALSE}

knitr::opts_chunk$set(
   echo = FALSE, 
   message = FALSE, 
   warning = FALSE,
   fig.align="center"
)

```

```{r setup06, echo = FALSE}

#libraries
library(patchwork)
library(ggplot2)
library(data.table)
library(magrittr)
library(pheatmap)
library(tidyr)
library(dplyr)
library(plyr)
library(ggpubr)
library(ggbeeswarm)
library(gtable)
library(glmnet)
library(RColorBrewer)
library(gridExtra)
library(cowplot)
library(tidyverse)

```

```{r loadData06, eval = TRUE}

#data("df", "patMeta", "LDT", "survT", "dds_smp")
#df_complete <- df

#Data
#df: tibble containing all screening viability data
load("data/df.RData")
df_complete <- df

#patMeta: tibble containing all patient genetic data
load("data/patMeta.RData")


#data for drug multivariate models 
#lasso coefficients
load("data/coefficients_lasso.RData") #from shiny app data
#gene Matrix
load("data/geneMatrix.RData") #from shiny app data
#viability Matrix
load("data/viabMatrix.RData") #from shiny app data


#data for drug - cyt - gene int analysis
#coefficients from lasso - linear modelling
load("data/coeff_mat_drug_cyt_gene_int.RData") #from figure 6 data script 

#predictor profiles 
load("data/Lasso_Plots_Fig6.RData") #from supplemental data




```

```{r themes_functions06, echo = FALSE, eval = TRUE}

### ggplot themes
fontsize = 11

## theme for ggplots
t1 <- 
  theme(                              
  plot.background = element_blank(), 
  panel.grid.major = element_line(),
  panel.grid.major.x = element_line(linetype = "dotted", colour = "grey"),
  panel.grid.minor = element_blank(), 
  panel.border = element_blank(), 
  panel.background = element_blank(),
  axis.line = element_line(size=.4),
  axis.line.x = element_line(),
  axis.line.y = element_line(),
  axis.text.x  = element_text(angle=90, size=11, hjust = 1, vjust = 0.4),
  axis.text.y = element_text(size = 11),
  axis.ticks.length = unit(0.3,"cm"),
  axis.title.x = element_text(face="bold", size=13), 
  axis.title.y = element_text(face="bold", size=13),
  plot.title = element_text(face="bold", size=14, hjust = 0.5),
  strip.text = element_text(size = fontsize)
)

t2 <- t1+
  theme( axis.text.x  = element_text(angle=0, size=11, hjust = 0.5, vjust = 1))

## theme for legends
t.leg <-  theme(legend.title = element_text(face='bold', 
                                            hjust = 1, size=11),
                legend.key = element_blank(),
                legend.text = element_text(size=11),
                legend.background = element_rect(color = "black"))



### Set colour palettes

#For Categorical: 
colors <- c("#A1BE1F", #green
            "#F4C61F", #yellow
            "#734595", #purple
            "#D41645", #red
            "#3B6FB6", #blue
            "#B65417", #orange
            "#E2E868", #light green
            "#CBA3D8", #light purple
            "#E58F9E", #light purple
            "#8BB8E8", #light blue
            "#F49E17", #light orange
            "#303030", #black
            "#A8A99E", #grey
            "#007B53") #dark green



#For Divergent: 
Divergent <- c("#003DA5", "#2055B0", "#406EBC", "#6086C7", "#809ED2", "#9FB6DD", "#BFCFE9", "#DFE7F4", "white", "white", "white","#F4E0E7", "#E9C2CF", "#DEA3B6", "#D3849E", "#C76586", "#BC476E", "#B12855", "#A6093D")

#for negatives only: 
palblues <- c("#003DA5", "#2055B0", "#406EBC", "#6086C7", "#809ED2", "#9FB6DD", "#BFCFE9", "#DFE7F4")

#for positives only:
palreds <- c("#F4E0E7", "#E9C2CF", "#DEA3B6", "#D3849E", "#C76586", "#BC476E", "#B12855", "#A6093D")

#For mutations: 
Mutant <- c("#b5b5b5","#373A36")
Sex <- c("#707372","#D0D0CE")
IGHV <- c("#373A36","#D0D0CE")
Methylation_cluster <- c("#373A36","#A8A99E","#D0D0CE")

#For drugs: 
drugpal <- c("#734595", "#CBA3D8") #purples

#For cytokines: 
cytpal <- c("#F49E17", "#EFC06E") #yellows


#neutral
offwhite <- "#f8f8ff"
lightergrey <- "#D0D0CE"
darkergrey <- "#707372"


#Function for multi-variant regression
runGlm06 <- function(X, y, method = "lasso", repeats = 20, folds = 3) {
  #set up objects to store results
  modelList <- list()
  lambdaList <- c()
  varExplain <- c()
  coefMat <- matrix(NA, ncol(X), repeats)
  rownames(coefMat) <- colnames(X)

  #set alpha
  if (method == "lasso"){
    alpha = 1
  } else if (method == "ridge") {
    alpha = 0
  }
  
  #for the set number of repeats, run the regression
  for (i in seq(repeats)) {
    if (ncol(X) > 2) {
      #run cross validated generalised linear model with given parameters
      res <- cv.glmnet(X,y, type.measure = "mse", family="gaussian", 
                       nfolds = folds, alpha = alpha, standardize = FALSE)
      
      #store lamdas and min lambda value
      lambdaList <- c(lambdaList, res$lambda.min)
      
      #store result of cv.glmnet
      modelList[[i]] <- res
      
      #get coefficents with min lambda value 
      coefModel <- coef(res, s = "lambda.min")[-1] #remove intercept row
      
      #store coefficients for this repeat
      coefMat[,i] <- coefModel
      
      #calculate variance explained
      if(sum(coefModel !=0)){
      y.pred <- predict(res, s = "lambda.min", newx = X)
      #if there are no predictors, all y.pred will be the same so its not possible to calculate variance explained because the SD is 0
      varExp <- cor(as.vector(y),as.vector(y.pred))^2
      }else{ varExp <- NA}
      varExplain[i] <- ifelse(is.na(varExp), 0, varExp) 
      
     
      
    } else {
      fitlm<-lm(y~., data.frame(X))
      varExp <- summary(fitlm)$r.squared
      varExplain <- c(varExplain, varExp)
      
    }

  }
  #store all results 
  list(modelList = modelList, lambdaList = lambdaList, varExplain = varExplain, coefMat = coefMat)
}


lassoPlot06 <- function(lassoOut, geneMatrix, betaMatrix, freqCut = 1, coefCut = 0.01) {
  #object to hold all plots
  plotList <- list()
  
  #for each drug - stimuli combination, run the following:
  for (seaName in names(lassoOut)) {
    ###FOR THE BAR PLOT
    #extract mean coefficients for each drug - stimuli combination
    barValue <- rowMeans(lassoOut[[seaName]]$coefMat)
    #extract proportion of repeats for which each coefficient is significant
    freqValue <- rowMeans(abs(sign(lassoOut[[seaName]]$coefMat)))
    #filter out coefficients that don't meet freqCut and coefCut thresholds
    barValue <- barValue[abs(barValue) >= coefCut & freqValue >= freqCut] 
    #arrange the bar values in numerical order
    barValue <- barValue[order(barValue)]
    #if there are no sig coefficients, don't plot
    if(length(barValue) == 0) {
      plotList[[seaName]] <- NA
      next
    }
    
   
    ###FOR THE HEATMAP AND SCATTER PLOT
    #get feature matrix and response matrix to plot
    allData <- geneMatrix
    betaValue <- unlist(betaMatrix[seaName,])
    
    #get feature matrix for features with significant coefficients only
    tabValue <- allData[, names(barValue),drop=FALSE]
    ord <- order(betaValue)
    betaValue <- betaValue[ord]
    tabValue <- tabValue[ord, ,drop=FALSE]
    sampleIDs <- rownames(tabValue)
    tabValue <- as_tibble(tabValue)
    tabValue$Sample <- sampleIDs
    
    #annotate features as mutations, methylation cluster or IGHV, and apply different scaling     so that different colours can be used in plotting 
    
    #for mutations:
    matValue <- gather(tabValue, key = "Var",value = "Value", -Sample)
    matValue$Type <- "mut"
    
    #for methylation cluster
    matValue$Type[grep("Methylation",matValue$Var)] <- "meth"
    
    #for IGHV status
    matValue$Type[grep("IGHV.status",matValue$Var)] <- "ighv"
    
    #change the scale of the value so that IGHV, Methylation and Mutation do not overlap
    matValue[matValue$Type == "mut",]$Value = matValue[matValue$Type == "mut",]$Value + 10
    matValue[matValue$Type == "meth",]$Value = matValue[matValue$Type == "meth",]$Value + 20
    matValue[matValue$Type == "ighv",]$Value = matValue[matValue$Type == "ighv",]$Value + 30
    
    #change continous to catagorical
    matValue$Value <- factor(matValue$Value,levels = sort(unique(matValue$Value)))
    
    #arrange order of feature rows and columns in heatmap
    #heatmap rows should align with order of genetic coefficients
    matValue$Var <- factor(matValue$Var, levels = names(barValue))
    
    #sample columns should be ascending order according to value of coefficient of each patient, so that heatmap aligns with scatter plot below
    matValue$Sample <- factor(matValue$Sample, levels = names(betaValue))
    
    #update labels, keep factor levels
    matValue$Var <- revalue(matValue$Var, c("IGHV.status" = "IGHV status", "del11q" = "del(11q)", "del13q" = "del(13q)", "del17p" = "del(17p)", "trisomy12" = "trisomy 12")) 
    
    #MAKE PLOTS
    #plot the heatmap of genetic feature values
    p1 <- ggplot(matValue, aes(x=Sample, y=Var)) + 
      geom_tile(aes(fill=Value), color = "white") + #ghost white
      theme_bw()+
      scale_y_discrete(expand=c(0,0)) + 
      theme(axis.title.x = element_text( size=fontsize+4),
            axis.text.y=element_text(hjust=0, size=18, face="bold"), 
            axis.ticks=element_blank(),
            panel.border=element_rect(colour="gainsboro"),  
            plot.title=element_text(face="bold", size = 18, margin = margin(t = -5, b = 1)), 
            panel.background=element_blank(),
            panel.grid.major=element_blank(), 
            panel.grid.minor=element_blank()) + 
      xlab("Mutation status for each patient") + 
      ylab("") + 
      scale_fill_manual(name="Mutated", 
                              values=c(`10`= offwhite,  #WT
                                       `11`="#373A36", #Mutant
                                       `20`= offwhite, #LP
                                       `20.5`= "#707372", #IP
                                       `21` = "#A8A99E", #HP
                                       `30` = offwhite, #IGHV-U
                                       `31` = "#707372"), #IGHV-M
                                       guide="none") + 
            ggtitle(seaName)
    
    
    #Plot the bar plot on the left of the heatmap 
    barDF = data.frame(barValue, nm=factor(names(barValue),levels=names(barValue)))
    
    p2 <- ggplot(data=barDF, aes(x=nm, y=barValue)) + 
      geom_bar(stat="identity", 
               fill=ifelse(barValue<0,
                           palblues[6],palreds[8]), 
               colour="black", 
               size=0.3) +
      scale_x_discrete(expand=c(0,0.5))+ 
      scale_y_continuous(expand=c(0,0), n.breaks = 4)+ 
      coord_flip(ylim=c(-0.3,0.35))+ 
      theme(panel.grid.major=element_blank(), 
            panel.background=element_blank(), 
            axis.ticks.y = element_blank(),
            panel.grid.minor = element_blank(), 
            axis.text=element_text(size=fontsize, angle = 0, hjust = 0),
                axis.title = element_text(size=fontsize+4), 
            panel.border=element_blank()) +
      ylab("Size of coefficient") + 
      geom_vline(xintercept=c(0.5), 
                 color="black", 
                 size=0.6)
    
    #Plot the scatter plot of patient coefficient values under the heatmap
    scatterDF = data.frame(X=factor(names(betaValue), 
                                    levels=names(betaValue)), 
                           Y=unlist(betaValue))
    
    p3 <- ggplot(scatterDF, aes(x=X, y=Y)) + 
          geom_point(shape=21, 
                     fill="dimgrey", 
                     colour="#707372", #dark grey
                     size=1.2) + 
          theme_bw() +
          theme(panel.grid.minor=element_blank(), 
                panel.grid.major.x=element_blank(), 
                axis.ticks.x=element_blank(), 
                axis.text.y=element_text(size=fontsize),
                axis.title = element_text(size=fontsize+4),
                panel.border=element_rect(colour="dimgrey", size=0.1),
                panel.background=element_rect(fill="white")) +
  xlab(expression(paste("Patient-specific ", beta["int"])))
    
    
    #Assemble all the plots together

    # construct the gtable
    wdths = c(0.2, 1.5, 0.4, 1.3*ncol(matValue), 1.7, 0.2)
    hghts = c(0.3, 0.3, 0.0020*nrow(matValue), 0.2, 0.8, 0.3)*1.5
    gt = gtable(widths=unit(wdths, "in"), heights=unit(hghts, "in"))
    
    ## make grobs
    gg1 = ggplotGrob(p1)
    gg2 = ggplotGrob(p2)
    gg3 = ggplotGrob(p3)

    ## fill in the gtable
   
    #HEATMAP
    #5:1 = "PREDICTORS"
    gt = gtable_add_grob(gt, gtable_filter(gg1, "panel"), 3, 4) # add heatmap
    gt = gtable_add_grob(gt, gtable_filter(gg1, "panel"), 3, 4) #add legend
    gt = gtable_add_grob(gt, gtable_filter(gg1, "title"), 1, 4) #add title to plot
    gt = gtable_add_grob(gt, gtable_filter(gg1, "axis-l"), 3, 5) # variable names
    gt = gtable_add_grob(gt, gtable_filter(gg1, "xlab-b"), 2, 4) # axis title
    
    #BARPLOT
    gt = gtable_add_grob(gt, gtable_filter(gg2, "panel"), 3, 2) # add barplot
    gt = gtable_add_grob(gt, gtable_filter(gg2, "axis-b"), 4, 2) # y axis for barplot
    gt = gtable_add_grob(gt, gtable_filter(gg2, "xlab-b"), 2, 2) # y lab for barplot

    
    #SCATTER PLOT
    gt = gtable_add_grob(gt, gtable_filter(gg3, "panel"), 5, 4) # add scatterplot
    gt = gtable_add_grob(gt, gtable_filter(gg3, "xlab-b"), 6, 4) # x label for scatter plot
    gt = gtable_add_grob(gt, gtable_filter(gg3, "axis-l"), 5, 3) #  axis for scatter plot
    
   

    
    #plot
    plotList[[seaName]] <- gt
  }
  return(plotList)
}

makelegends <- function (legendFor, colors) 
{
    x = NULL
    y = NULL
    colors = colors[names(colors) %in% legendFor]
    nleg = length(colors)
    
    #edit these widths to change alignment with lasso plots in patchwork code
    wdths = c(0.4,2,2,2,1.5)
    hghts = c(2)
    
    gtl = gtable(widths=unit(wdths, "in"), heights=unit(hghts, "in"))
    n = 2
    if ("M" %in% names(colors)) {
        Mgg = ggplot(data = data.frame(x = 1, 
                                       y = factor(c("LP", "IP", "HP"), 
                                                  levels = c("LP", "IP", "HP"))), 
                     aes(x = x, y = y, fill = y)) + 
              geom_tile() + 
              scale_fill_manual(name = "Methylation cluster", 
                                values = setNames(colors[["M"]], 
                                                  nm = c("LP", "IP","HP"))) + 
              theme(legend.title = element_text(size = 12), 
                    legend.text = element_text(size = 12))
        
        gtl = gtable_add_grob(gtl, gtable_filter(ggplotGrob(Mgg), "guide-box"), 1, n)
        n = n + 1
    }
    
    if ("I" %in% names(colors)) {
        Igg = ggplot(data = data.frame(x = 1, y = factor(c("Unmutated", 
            "Mutated"), levels = c("Unmutated", "Mutated"))), 
            aes(x = x, y = y, fill = y)) + geom_tile() + scale_fill_manual(name = "IGHV", 
            values = setNames(colors[["I"]], nm = c("Unmutated", "Mutated"))) + 
            theme(legend.title = element_text(size = 12), 
                  legend.text = element_text(size = 12))
        gtl = gtable_add_grob(gtl, gtable_filter(ggplotGrob(Igg), 
            "guide-box"), 1, n)
        n = n + 1
    }
    
    if ("G" %in% names(colors)) {
        Ggg = ggplot(data = data.frame(x = 1, y = factor(c("Wild Type", 
            "Mutated"), levels = c("Wild Type", "Mutated"))), 
            aes(x = x, y = y, fill = y)) + geom_tile() + scale_fill_manual(name = "Gene", 
            values = setNames(colors[["G"]], nm = c("Wild Type", "Mutated"))) + 
            theme(legend.title = element_text(size = 12), 
                  legend.text = element_text(size = 12))
        gtl = gtable_add_grob(gtl, gtable_filter(ggplotGrob(Ggg), 
            "guide-box"), 1, n)
        n = n + 1
    }
    
    return(list(plot = gtl, width = sum(wdths), height = sum(hghts)))
}


```

```{r processData06}
# List of Cytokines, Drugs and treatments 
  ##Drugs
  thedrugs <- unique(df$Drug) 

  ##Cytokines
  thecytokines <- unique(df$Cytokine) 

  #Treatments
  #drugs
  drugtreatments <- list()
  drugtreatments <- 
    lapply(thedrugs, function(x){
      paste("treatment_drug", x, sep='')
      })
  
  
  #cytokines
  cytokinetreatments <- list()
  cytokinetreatments <- 
    lapply(thecytokines, function(y){
      paste("treatment_cytokine", y, sep='')
      })

df_complete <- df 

# Tidy up data frame to use in linear model
df <- 
  dplyr::filter(df,
                #use high concentrations only
                Drug_Concentration %in% c("High","None")) %>% 
  #select columns for linear model
  dplyr::select(PatientID, Drug, Cytokine, Log) %>%  
        
  #rename columns 
  plyr::rename(c("Drug"="treatment_drug", "Cytokine"="treatment_cytokine", "Log"="Viability"))


#patMeta
patMeta$treatment <- as.factor(patMeta$treatment)
```

```{r preWork06, message = FALSE}

# define the base level per treatment_type as the "no"-treatment
df$treatment_drug <- as.factor(df$treatment_drug)
df$treatment_cytokine <- as.factor(df$treatment_cytokine)

df$treatment_drug %<>% relevel("DMSO")
df$treatment_cytokine %<>% relevel("No Cytokine")


# inspect design matrix of interaction model
# model.matrix(~treatment_drug + treatment_cytokine + treatment_drug:treatment_cytokine, df)

# fit model with interaction
fit <- lm(Viability ~ treatment_drug * treatment_cytokine, df)



#extract p values

pvalues <- summary(fit)$coefficients[,4]

#order as a dataframe
pvaldf <- as.data.frame(as.matrix(pvalues)) %>% 
          setDT(keep.rownames = TRUE)
          
#rename columns
colnames(pvaldf) <- c("treatment", "pvalue")

#filter out single agent treatments
singletreatments <- c(drugtreatments, cytokinetreatments, "(Intercept)")

pvaldf <- dplyr::filter(pvaldf, !treatment %in% singletreatments)


pvaldf <- 
  pvaldf %>% 
  #remove "treatment" prefix
  mutate_at(vars(treatment), funs(as.character(gsub("treatment_drug", "", .)))) %>% 
  mutate_at(vars(treatment), funs(as.character(gsub("treatment_cytokine", "", .))))
  
  #split drug:cytokine into two columns
pvaldf <-   
  data.frame(pvaldf, do.call(rbind, strsplit(pvaldf$treatment, split = ":", fixed = TRUE)))

#renove treatment column
pvaldf <- pvaldf[,c("pvalue","X1", "X2")]

#rename columns
colnames(pvaldf) <- c("pvalue","drug", "Cytokine")



```

## Mapping drug - stimulus interactions {#mapping-interactions}
### Linear modelling quantifies interactions between drugs and stimuli {#drug-stimulus-linear-model}
To begin, I evaluated computational methods for quantifying the impact of the stimuli on drug action. The aim was to screen for cases where the stimulus pathway specifically interacted with the target pathway of the drug. 

When a stimulus is individually applied to a CLL sample, the stimulus activates signalling cascades that modulate CLl viability and impact the rate of spontaneous apoptosis. When a stimulus is co-applied with a drug, the stimulus will continue to impact upon baseline viability, at the same time as the drug inhibits baseline viability. Assuming the drug and stimulus do not interact, the viability of the cells with the combinatorial treatment will equate to the additive impact of both these compounds. In the case that there is some interaction between the stimulus pathway and drug target pathway, the resulting viability will be more complex. The difference between the additive, or expected viability, and the true measured viability, can be quantified by an interaction factor. 

I fitted a linear model (Equation \@ref(eq:drugCytInt)) to quantify how each stimulus modulates drug efficacy beyond its individual effect [@Giles2021]. I used the `lm` function implemented in the `R` package `stats`[@R-base]. The model accounts for the individual effects of the drug and stimulus treatment on spontaneous apoptosis and baseline viability, in addition to an interaction term, named $\beta{int}$.  $\beta{int}$, the interaction factor,  quantifies how the combined treatment effect differs from the sum of the individual treatments. 

Equation \@ref(eq:drugCytInt) quantifies how the viability with any combination can be predicted:
    \begin{equation}
            log(V) = \beta_{drug}X_{drug} + \beta_{stimulus}X_{stimulus} + \beta_{int}X_{drug}X_{stimulus} + \epsilon
                                       (\#eq:drugCytInt)
    \end{equation}
    
_where $V$ is the predicted viability with a given treatment,_ $\beta_{drug}$, $\beta_{stimulus}$ _and_ $\beta_{int}$ _are coefficients for the drug, stimulus and combinatorial terms and_ $X_{drug}$ _and_ $X_{stimulus}$  _are indicator variables (0 or 1) for the presence or absence of a drug/stimulus._ $\epsilon$ _is the vector of model residuals. See also Methods section \@ref(drug-stimulus-linear-model-method). Equation from @Giles2021._
    
This approach identified  45 drug-stimulus combinations (out of 204), where $\beta{int}$ had p<0.05, i.e. the stimulus had a specific effect on the efficacy of the drug,  highlighting the extent to which drug action can be modulated by cell-external signals _ex vivo_. 

These interactions are of potential clinical interest, to support our understanding of  _in vivo_ drug response, and the development of strategies to target drug resistance pathways. They are also relevant to laboratory studies of drug action,  which are very often performed in the absence of such signals. 

To establish which interactions may be important in these contexts, we began by classifying the drug - stimulus interactions into four categories [@Giles2021].  The categories were firstly defined based on whether the interaction was antagonistic or synergistic, i.e. the stimulus acted to oppose or reinforce the activity of the drug. These were further divided  based on the sign of $\beta{int}$ i.e. whether the combinatorial viability was higher (postive) or lower (negative) than would be expected based on additive effects (Figure \@ref(fig:drugStimulusDummyPLots)A). 

(ref:drugStimulusDummyPLots) (A) Graphical line plots representing typical response patterns for each of the four drug - stimulus interaction categories (types I - IV). x-axis shows treatment, y-axis shows log transformed viability values, facet labels indicate interaction type. Blue and black horizontal lines demonstrate effect of interaction on viability with combinatorial treatment. Blue line shows predicted viability based on additive effects alone and black line shows viability accounting for additive effects and interaction. (B) Histogram showing number of drug-stimulus interactions within each category, where p value for $\beta_{int}$ is <0.05.  See Methods section \@ref(drug-stimulus-linear-model-method). _Figure and caption from @Giles2021._


```{r drugStimulusDummyPLots, fig.cap='(ref:drugStimulusDummyPLots)', message = FALSE,  echo = FALSE, fig.height=5, fig.width = 8, fig.align="center", out.width = '100%', dev = 'cairo_pdf', fig.pos="H"}


dummydata <-
  as.data.frame(list(c(1,2,3,4),
                     c(0,0,0,0),
                     c(-0.6,-0.6,0,-0.1), 
                     c(0.6,0.6,0.1,-0.1), 
                     c(0.5,-0.5,0.8,-0.8)), 
                col.names=c("plot",
                            "coord1", 
                            "coord2", 
                            "coord3", 
                            "coord4")) %>% 
  as_tibble() %>% 
  
  mutate(expected = coord2 + coord3, measured = coord4) %>% 
  
  pivot_longer(cols = coord1:measured, names_to = c("plot"), values_to="y.value", names_repair="unique" ) %>% 
  
  dplyr::rename(Plot = plot...1, x.value = plot...2) %>% 
  
  mutate(x.value = factor(x.value, levels = c("coord1", "coord2", "coord3", "coord4", "expected", "measured"))) %>% 
  
  #add annotation of interaction type
  mutate(Plot = factor(Plot, 
                       labels = c("I", 
                                  "II", 
                                  "III", 
                                  "IV"))) %>%
  mutate(type = ifelse(Plot %in% c("I", "II"), "antagonism", "synergism")) %>%
  mutate(direction = ifelse(Plot %in% c("I", "III"), "positive", "negative"))

my_labeller <- as_labeller(c("positive"="Positive~\u03B2[int]", "negative"="Negative~\u03B2[int]", "antagonism"="Antagonism", "synergism" = "Synergism"),
                           default = label_parsed)

dummydata$direction <- factor(dummydata$direction, levels = c("positive","negative"))
dummydata$type <- factor(dummydata$type, levels = c("antagonism","synergism"))
                              

#plot dummy plot
dummyPlot<- 
  ggplot(dplyr::filter(dummydata, grepl('coord', x.value)), 
       aes(x=x.value, y=y.value, group = Plot, color = direction))+
  geom_hline(yintercept = 0, linetype=2, color="black")+
  geom_point(size=1)+
  geom_line(size=1)+
  facet_grid(type~direction, labeller = my_labeller )+
  geom_segment(data=dplyr::filter(dummydata, grepl('expected', x.value)), aes(x = 3.5, y = y.value, xend = 4.5, yend = y.value), color="blue", size=2) +
  geom_segment(data=dplyr::filter(dummydata, grepl('measured', x.value)), aes(x = 3.5, y = y.value, xend = 4.5, yend = y.value), color="black", size=2)+
  theme_bw() +
  t1 +
  xlab("") + 
  ylab("Log(Viability)") +
  coord_cartesian(ylim = c(-1, 1)) +
  scale_x_discrete(labels = c( "coord1" = "DMSO", "coord2" = "Drug","coord3" = "Stimulus","coord4" = "Drug +\nStimulus")) +
    theme(axis.ticks.x = element_blank(),axis.ticks.y = element_blank(),
          plot.tag=element_text(size = 20), 
          axis.text.x = element_text(angle = 45,
                                   hjust = 1, 
                                   vjust = 1)) +
  scale_y_continuous(breaks = c(-1,0, 1)) +
  scale_color_manual(values = c("#A6093D","#003DA5"))  + theme(legend.position = "none") +
  geom_text(x = 1, y = 0.75, aes(label=Plot))


fit_as_df <- 
  fit$coefficients %>% 
  as.matrix() %>% 
  as.data.frame() %>%
  setDT(keep.rownames = TRUE) %>% 
  as_tibble() %>% 
  dplyr::rename(Condition = rn, comb_coefficient = V1) %>% 
  dplyr::filter(grepl("drug", Condition)&grepl("cytokine", Condition)) %>% 
  mutate(Condition = gsub("treatment_drug", "", Condition)) %>% 
  mutate(Condition = gsub("treatment_cytokine", "", Condition)) %>% 
  separate(Condition, c("Drug", "Cytokine"), sep=":")

drug_fit_as_df <- 
  fit$coefficients %>% 
  as.matrix() %>%
  as.data.frame() %>%
  setDT(keep.rownames = TRUE) %>% 
  as_tibble() %>% 
  dplyr::rename(Drug = rn, drug_coefficent=V1) %>% 
  dplyr::filter(grepl("drug", Drug)&!grepl("cytokine", Drug)) %>% 
  mutate(Drug = gsub("treatment_drug", "", Drug))


cyt_fit_as_df <-
  fit$coefficients %>% 
  as.matrix() %>% 
  as.data.frame() %>%
  setDT(keep.rownames = TRUE) %>% 
  as_tibble() %>% 
  dplyr::rename(Cytokine=rn, cyt_coefficent=V1) %>% 
  dplyr::filter(!grepl("drug", Cytokine)&grepl("cytokine", Cytokine)) %>% 
  mutate(Cytokine = gsub("treatment_cytokine", "", Cytokine))


fit_combinations <-
  left_join(fit_as_df, drug_fit_as_df, by = "Drug") %>% 
  left_join(cyt_fit_as_df, by = "Cytokine") %>% 
  mutate(predicted_coefficient=cyt_coefficent+drug_coefficent) %>% 
  mutate(obs_coef=predicted_coefficient+comb_coefficient) %>%
  left_join(mutate(pvaldf,drug=as.character(drug),Cytokine=as.character(Cytokine)), by=c("Drug"="drug", "Cytokine"="Cytokine")) %>% 
  dplyr::filter(pvalue<=0.05) %>%
  mutate(Category_Symbol=case_when(
    comb_coefficient>=0&(obs_coef<cyt_coefficent|obs_coef<drug_coefficent)~ "I",  
    comb_coefficient>=0&(obs_coef>cyt_coefficent&obs_coef>drug_coefficent)~ "III",
    comb_coefficient<0&(obs_coef>cyt_coefficent|obs_coef>drug_coefficent)~ "II",  
    comb_coefficient<0&(obs_coef<cyt_coefficent&obs_coef<drug_coefficent)~ "IV"
    )) %>% 
  mutate(Category=case_when(
    Category_Symbol=="I"~ "Positive \nAntagonistic",  
    Category_Symbol=="III"~ "Positive \nSynergistic",
    Category_Symbol=="II"~ "Negative \nAntagonistic",  
    Category_Symbol=="IV"~ "Negative \nSynergistic"
    )) %>% 
  mutate( Category=factor(Category, levels=c("Positive \nAntagonistic","Negative \nAntagonistic","Positive \nSynergistic","Negative \nSynergistic")))
  
barPlot <- 
fit_combinations %>% 
  dplyr::group_by(Category) %>% 
  tally() %>% 
 
  
 
ggplot(aes(x=Category, y=n, fill=Category))+
  geom_bar(width = 1, stat = "identity")+
  t1+
  guides(fill="none")+
  xlab("")+ ylab("Number of drug - stimulus pairs")+
  scale_fill_manual(values = c("#A6093D","#003DA5","#A6093D","#003DA5")) + 
  theme(axis.text.x = element_text(angle = 45,
                                   hjust = 1, 
                                   vjust = 1))

dummyPlot + barPlot + plot_annotation(tag_levels = c("A"))

```

These categories encompassed the following interaction types. Type I defined positive antagonisms in which stimuli reversed drug action, leading to decreased drug efficacy and increased CLL viability.  This group highlights drug - stimulus interactions that could be relevant to treatment resistance pathways _in vivo_. Type II negative antagonisms highlighted cases where the drug inhibited the stimulus, reducing the effect of the stimulus on baseline viability, and decreasing CLL viability. These interactions include drugs that provide potential therapies to target treatment - resistance pathways _in vivo_. Type III and type IV interactions cover synergies, in which the drug and stimulus increase the effect of the other. Type III positive synergies lead to increased CLL viability with the combinatorial treatment, whilst type IV negative synergies lead to increased drug efficacy and lower viability. In the latter case, these may highlight cases where drug efficacy is dependent on external pathways, and may be useful for understanding _ex vivo_ drug action. Figure \@ref(fig:drugStimulusDummyPLots)B quantifies the number of interactions in each category. 


### INF$\gamma$ induces resistance to ibrutinib _in vitro_  {#drug-stimulus-interaction-examples}
Using the linear modelling approach in combination with our classification system, we attempted to generate the first comprehensive map of drug - stimulus interactions in CLL [@Giles2021]. This map provides a resource to explore the impact of the microenvironment on the majority of drugs that are clinically-licensed in CLL (Figure \@ref(fig:drugStimulusIntMapt)). We use this resource to investigate a number of drug - stimulus interactions here, and in section \@ref(chapter7).  

(ref:drugStimulusIntMapt) Heatmap of all $\beta_{int}$ values for all drug - stimulus combinations where p for $\beta_{int}$ <0.05, annotated with interaction type. Scale indicates size and sign of $\beta_{int}$. Rows and columns clustered according to hierarchical clustering.  See Methods section \@ref(drug-stimulus-linear-model-method). _Figure and caption from @Giles2021._

```{r drugStimulusIntMapt, fig.cap='(ref:drugStimulusIntMapt)', message = FALSE,  echo = FALSE, fig.width=8, fig.height = 6, fig.align="center", out.width = '60%', dev = 'cairo_pdf'}

#extract beta interaction values
betavalues <- summary(fit)$coefficients[,1]

#create matrix of beta values 
betadf <- as.data.frame(as.matrix(betavalues)) %>%
          setDT(keep.rownames = TRUE)

#rename columns
colnames(betadf) <- c("treatment", "betavalue")

#remove beta values for drugs and cytokines alone
betadf <- dplyr::filter(betadf, !treatment %in% singletreatments)

#remove treatment prefix
betadf <- betadf %>% 
  mutate_at(vars(treatment), funs(as.character(gsub("treatment_drug", "", .)))) %>% 
  mutate_at(vars(treatment), funs(as.character(gsub("treatment_cytokine", "", .))))

#split drug:cytokine by colon
betadf <- data.frame(betadf, do.call(rbind, strsplit(betadf$treatment, split = ":", fixed = TRUE)))

#select columns of interest
betadf <- betadf[,c("betavalue","X1", "X2")]

#rename columns 
colnames(betadf) <- c("betavalue","drug", "Cytokine")

#set significance
a <- 0.05

#make a matrix of beta values, where  beta = 0 if corresponding p val is > than significance threshold 
pvalmat <- xtabs(pvalue~drug+Cytokine, data=pvaldf)
bvalmat <- xtabs(betavalue~drug+Cytokine, data=betadf)

#check in same order
bvalmat <- bvalmat[,colnames(pvalmat)] 
bvalmat <- bvalmat[rownames(pvalmat),]

bvalmat[pvalmat >= a] = 0

#transform matrix 
bvalmat <- t(bvalmat)

tree <- 
  pheatmap(bvalmat, silent = TRUE)


cyt_order <- rownames(bvalmat[tree$tree_row[["order"]],])

drug_order <- colnames(bvalmat[,tree$tree_col[["order"]]])



  #append p values
  left_join(betadf, pvaldf, by = c("drug", "Cytokine")) %>% 
  #make beta value 0 if p value is < sig
  mutate(betavalue = ifelse(pvalue<a, betavalue, 0)) %>% 
  #put drugs in order of clustered heatmap
  mutate(drug=factor(drug, levels = drug_order)) %>% 
  
  #put stimuli in order of clustered heatmap
  mutate(Cytokine=factor(Cytokine, levels = rev(cyt_order))) %>% 
  
  #make heatmap with ggplot
  ggplot(aes(x=drug, y=Cytokine))+
  
  geom_tile(aes(fill=betavalue),color = "grey")+
  #set colour scale
  scale_fill_gradientn(colors=c("#003DA5",  "white",  "#A6093D"), limits=c(-0.6,0.6))+
  #set theme
  t2+
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks = element_blank(),
        axis.ticks.x =element_blank(),
        panel.background = element_blank(), 
        axis.line.x = element_blank(),
        axis.line.y = element_blank(),
        legend.key.height=unit(2.5, "cm"))+
  labs(fill = expression(paste(beta["int"], "-value")))+
  #add category
  geom_text(data = fit_combinations, aes(x=Drug, label=Category_Symbol) )+
  scale_y_discrete(labels=c("TGF-b1"="TGF-\u03B21", "sCD40L+IL-4"="sCD40L + IL4", "IL-1b"="IL1\u03B2", "IL-4"="IL4", "IL-6"="IL6","IL-15"="IL15","IL-10"="IL10", "IL-21"="IL21","IL-2"="IL2", "Interferon gamma"= "Interferon \u03B3", "SDF-1a"="SDF-1\u03B1"))
  


```

To complement this broader approach, I next examined each interaction individually, to select the most clinically interesting examples (Figure \@ref(fig:drugStimulusIntEx)). 

Positive antagonistic interactions were the most common, in which stimuli reversed drug action. This group included resistance mechanisms such as the inactivation of ibrutinib by IL4 stimulation (Figure \@ref(fig:drugStimulusIntEx)A)[@Aguilar-Hernandez2016], indicating that the modelling and classification approach  successfully recapitulated known drug - stimulus interactions.  Our approach demonstrated that IL4 induced resistance to a range of drugs, beyond ibrutinib, and showed the highest number of positive antagonistic interactions of all of the stimuli. 

Amongst the positive antagonistic interactions,  we noted reduced efficacy of ibrutinib in the context of IFN$\gamma$ stimulation (Figure \@ref(fig:drugStimulusIntEx)B, [@Giles2021]). [ADD FOLLOW UP / STAT1 experiment].

There were six cases of negative antagonisms, whereby drug action inhibited the pro-survival effect of the stimulus. As IL4 appeared to be the most potent conferrer of drug resistance, it was of greatest interest to identify drugs that may inhibit this pathway.  For example, the Pan-JAK inhibitor pyridone-6 inhibited the increase in viability with sCD40L + IL4 stimulation  (Figure \@ref(fig:drugStimulusIntEx)C, [@Giles2021]).

The model identified a single positive synergism, in which IFN$\gamma$ treatment in combination with ralimetinib, a p38 MAPK inhibitor, stimulated a large increase in viability that was not observed in each of the single treatments (Figure \@ref(fig:drugStimulusIntEx)D, [@Giles2021]). This finding suggests  a potential inhibitory effect of p38 MAPK activity on signalling via IFN$\gamma$. [ADD TO THIS NOS THEORY]

16 drug - stimulus combinations demonstrated negative synergistic interactions. This implied that the drug efficacy was somehow dependent upon, or increased by, activation of the stimulus pathway. For example, luminespib, a HSP90 inhibitor, showed higher efficacy with a number of stimuli, including soluble anti-IgM, indicating that _ex vivo_ measurements of luminespib action are likely to be affected by the presence or absence of microenvironmental signals (Figure \@ref(fig:drugStimulusIntEx)F, [@Giles2021]). 

Collectively, these results represent a comprehensive and systematic study of the influence of the microenvironment on drug efficacy in CLL. This work highlights key resistance pathways, strategies for targeting microenvironmental resistance, and underlines important pathways in _ex vivo_ studies of drug efficacy. In this thesis, I have selected some of the more interesting examples, such as resistance to Ibrutinib induced by IL4 and IFN$\gamma$, to investigate in more detail here and in Chapter \@ref(chapter7). This work also aims to serve as a resource for others, to provide a basis for further follow-up studies beyond the ones in this thesis. It also aims to provide a proof-of-principle highlighting the value of combinatorial screening, and linear modelling approaches, to study the impact of the microenvironment in cancers in general. All drug - stimulus combinations and interactions can be explored on the [online shiny app](https://www.imbi.uni-heidelberg.de/dietrichlab/CLL_Microenvironment/), and the whole analysis can be replicated, or adapted, from the [online code repository](https://github.com/Huber-group-EMBL/CLLCytokineScreen2021), published alongside [@Giles2021].  


(ref:drugStimulusIntEx)  Line plots showing examples of drug-stimulus interactions. x axis indicates treatment, y axis shows log transformed viability values, with matching samples linked across treatments. Black horizontal lines represent  predicted viability for each treatment, using coefficients from linear model fit. For combinatorial treatment, blue line indicates predicted viability based on additive effect alone, and black line indicates predicted viability accounting for additive effects plus interaction. See Methods section \@ref(drug-stimulus-linear-model-method). _Figure and caption from [@Giles2021]._

```{r drugStimulusIntEx, fig.cap='(ref:drugStimulusIntEx)', message = FALSE,  echo = FALSE, fig.height=10, fig.width = 13, fig.align="center", out.width = '80%', dev = 'cairo_pdf', fig.pos = "H"}


#get lists of drugs and cytokines without baseline treatments
thedrugs.only <- thedrugs %>% setdiff("DMSO") %>% sort()
thecytokines.only <- thecytokines %>% setdiff("No Cytokine")%>% sort()

# get a list of interactions for which p value of beta interaction is significant
sign_conditions <-
  pvaldf %>% 
  dplyr::filter(pvalue<=0.05) %>% 
  mutate(condition = paste0("treatment_drug:", 
                            drug, 
                            ":treatment_cytokine:",
                            Cytokine))


gg <- vector(mode="list", length=length(sign_conditions$condition))
names(gg) <- sign_conditions$condition



plotList <- 
lapply(names(gg), function(i){
  
  #preparations for the interaction plot
  drug <- dplyr::filter(sign_conditions, condition ==i)$drug
  drugTreat <- paste("treatment_drug", drug, sep='')
      
  cyt <- dplyr::filter(sign_conditions, condition ==i)$Cytokine
  cytokineTreat <-paste("treatment_cytokine", cyt, sep='')
      
      
  Category <- fit_combinations[which(fit_combinations$Drug==drug &
                               fit_combinations$Cytokine==cyt),]$Category
  
  #extract coefficients to plot

  ## baseline effect
  baseline <- fit$coefficients["(Intercept)"]

  ## single treatment effect for type_1 = "a"
  single_1 <- fit$coefficients[drugTreat]

    ## single treatment effect for type_2 = "b"
    single_2 <- fit$coefficients[cytokineTreat]

    # interaction Viability
    interaction <- fit$coefficients[paste(drugTreat, cytokineTreat, sep=':')]

      
    ## observed single treatment values
    observed_single_1 <- baseline + single_1 
    observed_single_2 <- baseline + single_2

    ## observed double treatment value
    observed_double <- baseline + single_1 + single_2 + interaction

    ## predicted double treatment value (based only on single effects)
    predicted_double <- baseline + single_1 + single_2


    #prepare labels and aesthetics for plotting
    # y_label <- case_when(cyt=="Interferon gamma"~"Interferon\ngamma",
    #                      cyt=="soluble anti-IgM"~"soluble\nanti-IgM",
    #                      TRUE~cyt)

    xlabels <- c("DMSO", drug, 
                 case_when(cyt=="IL-4"~"IL4",
                           cyt=="Interferon gamma"~"Interferon \u03B3",
                           cyt=="sCD40L+IL-4"~"sCD40L \n+ IL4",
                           cyt=="soluble anti-IgM"~"soluble\nanti-IgM",
                           TRUE~cyt),
                paste(drug, " +\n",
                       case_when(cyt=="IL-4"~"IL4",
                           cyt=="Interferon gamma"~"Interferon \u03B3",
                           cyt=="sCD40L+IL-4"~"sCD40L + IL4",
                           TRUE~cyt),
                sep=""))
      
    segment_size <- 2
    
    
    plotTab <-
      df %>%
      
      dplyr::filter(treatment_drug %in% c("DMSO" , drug)) %>%

      dplyr::filter(treatment_cytokine %in% c("No Cytokine", cyt)) %>%

      dplyr::mutate(treatment_combination = 
                  #if baseline treatment
                  ifelse(treatment_drug=="DMSO" & 
                         treatment_cytokine=="No Cytokine",
                         "DMSO",
                         #if drug only treatment
                         ifelse(treatment_drug == drug & 
                                treatment_cytokine =="No Cytokine",
                                paste0("Drug=", drug),
                                #if cytokine only treatment
                                ifelse(treatment_drug == "DMSO" & 
                                       treatment_cytokine == cyt,
                                       paste0("Cytokine=", cyt),
                                       #otherwise combinatorial treatment
                                       paste0("Drug=", drug, "\nCytokine=", cyt))))) %>%
  #make treatment_combination as a factor
  mutate(treatment_combination = factor(treatment_combination, 
                                        levels=c("DMSO", 
                                                 paste0("Drug=", drug),
                                                 paste0("Cytokine=", cyt),
                                                 paste0("Drug=", drug, "\nCytokine=", cyt)
                                                 )))
  #make plot
        
        

  ggplot(plotTab, aes(treatment_combination, Viability, group=PatientID)) +
  
  geom_hline(yintercept = 0, linetype=2, color="black")+

  lemon::geom_pointline( size=1,   
                         colour=ifelse(interaction > 0,
                                       "#A6093D","#003DA5"), alpha=0.3) +
  #add predicted viability with control treatment
  geom_segment( size=segment_size,
                aes(x=1-.2, xend=1+.2, y=baseline, yend=baseline),
                colour = "black") +
  #add predicted viability with drug only treatment
  geom_segment( size=segment_size,
                aes(x=2-.2, xend=2+.2, y=observed_single_1,yend=observed_single_1),
                colour="black") +
  #add predicted viability with cytokine only treatment
  geom_segment( size=segment_size,
                aes(x=3-.2, xend=3+.2, y=observed_single_2, yend=observed_single_2),
                colour="black") +
  #add predicted viability with both treatments
  geom_segment( size=segment_size,
                aes(x=4-.2, xend=4+.2, y=observed_double, yend=observed_double),
                colour="black") +
  #add predicted viability with both treatments, without interaction
  geom_segment( size=segment_size,
                aes(x=4-.2, xend=4+.2, y=predicted_double, 
                    yend=predicted_double), color="blue") + 
 
   #add category and combination as title
    
    
    
  ggtitle(paste( Category,": ", drug, " & ", 
                 case_when(cyt=="IL-4"~"IL4",
                           cyt=="Interferon gamma"~"Interferon \u03B3",
                           cyt=="sCD40L+IL-4"~"sCD40L + IL4",
                           TRUE~cyt), sep="")) + 
  
  #add theme
  t2 + theme(axis.text.x = element_text(size=fontsize+5)) +
    theme(plot.tag=element_text(size = 30))+
  scale_x_discrete(labels = xlabels) +
  ylab("Log (Viability)") + 
  xlab("")


  } ) 

names(plotList) <- names(gg)

design1 <-"
  12
  34
  56"
  
plotList$`treatment_drug:Ibrutinib:treatment_cytokine:IL-4` +
  
plotList$`treatment_drug:Ibrutinib:treatment_cytokine:Interferon gamma` + 
  
plotList$`treatment_drug:Pyridone-6:treatment_cytokine:sCD40L+IL-4` +
  
plotList$`treatment_drug:Ralimetinib:treatment_cytokine:Interferon gamma`+
  
plotList$`treatment_drug:Ibrutinib:treatment_cytokine:CpG ODN` +
  
plotList$`treatment_drug:Luminespib:treatment_cytokine:soluble anti-IgM` +  plot_annotation(tag_levels=c("A"))  + plot_layout(design=design1) 
  
```

## Genetic modulators of drug response {#mapping-genetic-modulators}
### Univariate analysis identifies genetic modulators of drug response {#univariate-gene-drug-associations}
Most mutations in cancer have not been linked to drug response [@JCIpaper]. CLL has provided an powerful model system in which to study the link between recurrent genetic features and drug response: liquid biopsies are easy to obtain and sequence, and screen _ex vivo_ and it is much easier than in many cancers to take samples over a time course. Previous work in our lab [@JCIpaper] has taken advantage of this, generating the largest survey of molecular determinants of response in CLL to date. 

Before investigating the impact of mutations on drug - stimulus interactions,  I performed a similar analysis of genetic determinants of drug response, to first establish the effects of mutations independently of stimuli on drug response. I used the genetic data, including all mutations, copy number variants and IGHV status for which there were greater than 3 cases in the cohort (n = 54), to perform t-tests to identify molecular features that impact on drug response (Figure \@ref(fig:drugGeneIntMap)). 

This analysis recapitulated the findings of previous work [@JCIpaper], including the impact of IGHV status on responses to BCR inhibitors ibrutinib (BTK) and idelalisib (Pi3K), and the impact of TP53 and del(17p) mutations on Nutlin-3a (MDM2) and Fludarabine (Purine analogue). 

(ref:drugGeneIntMap) Plot showing BH-adjusted p values from Student's t-tests (two-sided, with equal variance), for all tested drug-gene associations. Tests performed for IGHV status and somatic mutations and copy number aberrations (n=54). Each coloured circle represents a gene-stimulus association meeting 10% FDR cut off.  Method \@ref(univariate-gene-stimulus-associations-method) _Figure from [@Giles2021]._ 

```{r drugGeneIntMap, fig.cap='(ref:drugGeneIntMap)', message = FALSE,  echo = FALSE, fig.height=5, fig.width = 7, fig.align="center", out.width = '60%', dev = 'cairo_pdf'}

df_univariate <- df_complete %>% filter(Drug_Concentration=="High")

##################### List of mutations with >2 positive cases #############################
selected_mutations <- 
  patMeta %>% 
  mutate(Ras_Raf = as.factor(ifelse(BRAF == 1 | KRAS == 1| NRAS == 1, 1, 0))) %>% 
  dplyr::select(-BRAF, -KRAS, -NRAS) %>% 
  dplyr::select( -gender, 
                 -diagnosis, 
                 ) %>%
  pivot_longer(-PatientID, 
               names_to = "Genetic_Alteration", 
               values_to = "alteration_value") %>% 
  dplyr::filter(alteration_value %in% c(1, "U")) %>% 
  dplyr::group_by(Genetic_Alteration) %>% 
  dplyr::count() %>% 
  dplyr::filter(n>2) %>% 
  dplyr::select(Genetic_Alteration) %>% 
  unlist()

##################### t tests and p-value adjustment ###############################  
p_values <-
  filter(df_univariate, Cytokine == "No Cytokine", Drug != "DMSO") %>%
  ## Select columns from screening data 
  dplyr::select(PatientID, Log, Drug) %>% 
  
  ## Join Screening data with metadata
  left_join(patMeta, by = "PatientID") %>%
  
  ## Add Ras/Raf column, remove single columns
  mutate(Ras_Raf=as.factor(ifelse(BRAF == 1 | KRAS == 1 | NRAS == 1, 1, 0))) %>% 
  dplyr::select(-BRAF, -KRAS, -NRAS) %>% 
## remove unused columns from metadata
  dplyr::select( -gender, 
                 -diagnosis, 
                 -treatment) %>% 
  
  ## transform data to long format
  pivot_longer(cols=c(-PatientID, -Log, -Drug), 
               names_to = "Genetic_alt", 
               values_to = "alt_value") %>% 
  
  ## filter to selected mutations (see above)
  dplyr::filter(Genetic_alt %in% selected_mutations) %>% 
  
  
  ## group by Drug and Genetic alteration  
  dplyr::group_by(Drug, Genetic_alt) %>% 
  
  ## Perform t.test on every combination of Cytokine and genetic alteration
  do(tidy(t.test(Log ~ alt_value, data = ., var.equal = T))) %>% 
  
  ## ungroup before adjusting p-value
  ungroup() %>% 
  
  ## adjust p-values to multiple testing using BH method 
  mutate(adj.p.value=p.adjust(p.value, method = "BH"))
  
################################## Order of cytokines by descending significance ############################
Drug_order <-
  p_values %>% 
  dplyr::group_by(Drug) %>% 
  dplyr::arrange(adj.p.value) %>% 
  dplyr::filter(row_number() == 1) %>% 
  ungroup() %>% 
  dplyr::arrange(adj.p.value) %>% 
  dplyr::select(Drug) %>% 
  unlist()

Genetic_alt_order <-
  p_values %>% 
  dplyr::group_by(Genetic_alt) %>% 
  dplyr::arrange(adj.p.value) %>% 
  dplyr::filter(row_number() == 1) %>% 
  ungroup() %>% 
  dplyr::arrange(adj.p.value) %>% 
  dplyr::select(Genetic_alt) %>% 
  unlist()

############################# Define FDR cutoff #################################
fdr = 0.1
  
############################################ Plot ########################################
p_values %<>% 
  mutate(Drug=factor(Drug, levels = Drug_order)) %>% 
  mutate(Genetic_alt=factor(Genetic_alt, levels = Genetic_alt_order))
  
  ggplot(dplyr::filter(p_values, adj.p.value>fdr), 
         aes(x = Drug, y = -log10(adj.p.value))) +
  geom_point(color = "lightgrey", size = 3) +
  geom_beeswarm(data = dplyr::filter(p_values, adj.p.value <= fdr), 
                aes( color=Genetic_alt), size = 3, cex = 1.7) +
  ##FDR line  
  geom_hline(yintercept = -log10(fdr),linetype = "dashed", size=0.3)+
  
  ##Main Theme
  t1 +
  
  theme(axis.text.x = element_text(angle = 45,
                                   hjust = 1, 
                                   vjust = 1),
        axis.title.x=element_blank()) +
  ##Legend Theme
  theme(legend.position=c(0.75,0.75), 
        legend.title = element_text(face='bold', hjust = 0, size=13), 
        legend.key = element_blank(),  
        legend.text = element_text(size=11)) +
  guides(colour = guide_legend(nrow=7, title = "Mutations"), 
         shape = guide_legend(ncol = 1)) +
  scale_color_manual(name = "Mutations", values = c(colors, "#373A36", "#A8A99E", "#54585A", colors), labels=c("IGHV.status"="IGHV status", 
                                                                   "del9p"="del(9p)", 
                                                                   "trisomy12"="trisomy 12", 
                                                                   "gain17q"="gain(17q)", 
                                                                   "gain2p"="gain(2p)",
                                                                   "gain19p"="gain(19p)",
                                                                   "gain19q"="gain(19q)",
                                                                   "del7q"="del(7q)",
                                                                   "del9q"="del(9q)", 
                                                                   "del4p"= "del(4p)", 
                                                                   "SPEN"="SPEN", 
                                                                   "del11q"="del(11q)", 
                                                                   "del1q" = "del(1q)")) +
  scale_y_continuous(expression("BH-adjusted  "* italic(p)*"-value"), 
                     breaks=seq(0, 10, 5),
                     labels=math_format(expr=10^.x)(-seq(0,10,5))) 

```


In addition, it also highlighted the broad impact of trisomy 12 status on drug response _ex vivo_ (Figure \@ref(fig:drugGeneIntEx)). While trisomy 12 is known to accelerate disease progression, tumours with trisomy 12 are more treatable due to higher sensitivity to chemotherapeutics and ibrutinib. In my results, trisomy 12 increased sensitivity to all drugs except ------fludarabine, BAY-11-7085 and Everolimus. These included increased sensitivity to p38 MAPK inhibition by ralimetinib, agreeing with previous work demonstrating higher trisomy 12 sensitivity with a range of MEK and ERK inhibitors, suggesting an essential role for MEK/ERK signalling in trisomy 12 CLL [@JCIpaper]. Trisomy 12 samples also showed higher sensitivity to Nutlin-3a, an MDM2 inhibitor. Interesting, the MDM2 gene is located on chromosome 12, and is more abundant in trisomy 12 samples [ADD SUPP FIG], which may explain this.  I also noted increased trisomy 12 sensitivity to IBET-762, a BET inhibitor that has been shown to suppress transcriptional responses to cytokine signalling via JAK-STAT [@Chan2015]. If trisomy 12 samples are more dependent on cytokine signals, as the results in section \@ref(trisomy12-modulator) suggest, this makes sense. 

(ref:drugGeneIntEx) Beeswarm-boxplots showing control-normalised log transformed viability values after treatment with (A) Ralimetinib, (B) Nutlin-3a and (C) IBET-762, stratified by trisomy 12.

```{r drugGeneIntEx, fig.cap='(ref:drugGeneIntEx)', message = FALSE,  echo = FALSE, fig.height=5, fig.width = 10, fig.align="center", out.width = '60%', dev = 'cairo_pdf', fig.pos="H"}

df_patmeta <- left_join(df_univariate, patMeta, by = "PatientID")

ibet_tri12 <- 
df_patmeta %>% 
  dplyr::filter(Drug=="I-BET 762", Drug_Concentration == "High", Cytokine == "No Cytokine") %>%
      filter(!is.na(trisomy12)) %>%
         
  ggplot(aes(x=trisomy12,y=Log,color=(trisomy12)))+
  geom_hline(yintercept = 0)+
  geom_boxplot()+
  geom_beeswarm(cex=1.5) +
  guides(color="none", shape="none")+
  stat_compare_means(method = "t.test",
                     label.x.npc = "center", 
                     size=3)+
  xlab("trisomy 12") +
  ylab("Log(Viability)") +
  ggtitle("IBET-762") +
  scale_color_manual(values=c(colors[1], colors[2])) + 

  t2




nutlin_tri12 <- 
df_patmeta %>% 
  dplyr::filter(Drug=="Nutlin-3a", Drug_Concentration == "High", Cytokine == "No Cytokine") %>%
      filter(!is.na(trisomy12)) %>%
         
  ggplot(aes(x=trisomy12,y=Log,color=(trisomy12)))+
  geom_hline(yintercept = 0)+
  geom_boxplot()+
  geom_beeswarm(cex=1.5) +
  guides(color="none", shape="none")+
  stat_compare_means(method = "t.test",
                     label.x.npc = "center", 
                     size=3)+
  xlab("trisomy 12") +
  ylab("Log(Viability)") +
  ggtitle("Nutlin-3a") +
  scale_color_manual(values=c(colors[1], colors[2])) + 

  t2

ralimetinib_tri12 <- 
df_patmeta %>% 
  dplyr::filter(Drug=="Ralimetinib", Drug_Concentration == "High", Cytokine == "No Cytokine") %>%
      filter(!is.na(trisomy12)) %>%
         
  ggplot(aes(x=trisomy12,y=Log,color=(trisomy12)))+
  geom_hline(yintercept = 0)+
  geom_boxplot()+
  geom_beeswarm(cex=1.5) +
  guides(color="none", shape="none")+
  stat_compare_means(method = "t.test",
                     label.x.npc = "center", 
                     size=3)+
  xlab("trisomy 12") +
  ylab("Log(Viability)") +
  ggtitle("Ralimetinib") +
  scale_color_manual(values=c(colors[1], colors[2])) + 

  t2

ralimetinib_tri12 + nutlin_tri12 + ibet_tri12
```
## The modulatory effect of mutations on drug - stimuli interactions {#drug-stimulus-gene-interactions} 
### Patient-specific linear modelling identifies drug - stimulus - gene interactions 
Sections \@ref(mapping-interactions) & \@ref(mapping-genetic-modulators), explore the effects of microenvironmental signals and molecular features on drug response independently. I next aimed to investigate their collective effect on _in vitro_ drug efficacy in CLL, by quantifying the extent to which genetic driver mutations modulated the interactions between stimuli and drugs [@Giles2021].

I began by conceptualising methods to quantify the collective effect of the drugs, stimuli and genetic features on CLL viability and biology.  I decided to adapt the linear model in Equation \@ref(eq:drugCytInt), by fitting this model in a patient sample - specific manner [@Giles2021]:  

Equation \@ref(eq:drugCytGeneInt) quantifies how the viability with any combination can be predicted:
    \begin{equation}
            log(V) = \beta_{drug}X_{drug} + \beta_{stimulus}X_{stimulus} + \beta_{patient}X_{patient} + \beta_{drug-stimulus}X_{drug}X_{stimulus} + \beta_{drug-patient}X_{drug}X_{patient} + \beta_{stimulus-patient}X_{stimulus}X_{patient} +\beta_{int}X_{drug}X_{stimulus}X_{patient} + \epsilon
                                       (\#eq:drugCytGeneInt)
    \end{equation}
    
_where $V$ is the predicted viability of a patient sample with a given treatment,_ $\beta_{drug}$,  $\beta_{stimulus}$, $\beta_{int}$, $\beta_{drug-stimulus}$, $\beta_{drug-patient}$, $\beta_{drug-stimulus}$ _and_ $\beta_{int}X_{drug}$ _are regression coefficients for the drug, stimulus, patient sample and combinatorial terms and_ $X_{drug}$, $X_{stimulus}$ _and_ $X_{patient}$ _are indicator variables (0 or 1) for the presence or absence of a drug/stimulus/patient sample._ $\epsilon$ _is the vector of model residuals. See also Methods section \@ref(drug-stimulus-gene-interactions-method). Equation from [@Giles2021]._

This generates a higher order interaction term $\beta_{int}X_{drug}X_{stimulus}X_{patient}$. This term represents a _patient sample-specific_ $\beta{int}$ for each drug - stimulus combination, quantifying the size of an interaction between a drug and stimulus in each patient genetic background. 

With these patient sample-specific $\beta{int}$ terms, it was possible to search for associations between the size of $\beta_{int}$) and genetic features. The aim was to screen for molecular features that increased or decreased the size of a drug - stimulus interactions, using multivariate regression with L1 (lasso) regularisation.

As input to the model, the response matrix was composed of the sample - specific $\beta_{int}$ values for each drug-stimulus combination. To generate the feature matrix (137 samples versus 40 features), I excluded genetic features for which >20\% of the values were missing, and  patient samples with incomplete annotation. As predictors, I included genetic mutations and CNVs (p=39) and IGHV status (coded as 0-1). I ran lasso regression, as implemented in the `R` package `glmnet`[@R-glmnet], using three-fold cross-validation with misclassification error as loss. The resulting predictors are the mean of those coefficients that were selected in at least 90% of 30 bootstrapped repeats. 

This analysis revealed that 60/204 drug - stimulus interactions were modulated by at least one genetic feature (Figure \@ref(fig:drugCytGeneIntMap),  Appendix Figure \@ref(fig:drugcytGeneIntAll)). A positive coefficient here indicates that the presence of the genetic feature is associated with more positive $\beta{int}$, in other words, the viability with the drug and stimulus is higher than expected in the presence of the genetic feature. 

*Have to make sure this figure takes a whole page but doesn't float to the bottom*

(ref:drugCytGeneIntMap) Heatmap summarising genetic predictors of drug - stimulus interactions. Each row depicts a single drug - stimulus combination, and each tile indicates that $\beta{int}$ for given drug and stimulus combination is modulated by corresponding genetic feature. Thus each row represents the output of a single model fit (as in Figure \@ref(fig:FluCpGExA)). Colour of tile indicates size and sign of coefficient assigned to genetic feature, where a positive coefficient corresponds to a more positive $\beta{int}$ if the feature is present. Only top 8 most commonly selected genetic features are shown. Drug - stimulus combinations with no genetic predictors of $\beta{int}$ amongst top 8 shown are omitted for clarity. See Methods section \@ref(drug-stimulus-gene-interactions-method). _Figure from [@Giles2021]._

```{r drugCytGeneIntMap, fig.cap='(ref:drugCytGeneIntMap)', message = FALSE,  echo = FALSE, fig.height=18, fig.width = 9, fig.align="center", out.width = '50%', dev = 'cairo_pdf'}

#cluster coeff.mat
fit <-
  coeff.mat %>% 
  dist() %>% 
  hclust()
  
order_comb <- rownames(coeff.mat)[fit$order]

#Put matrix into long format for ggplot
coeff.long.mat <- coeff.mat %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column("int") %>% 
  tidyr::gather(key = "gene", value = "coeff", -int)

#get top 8 coefficients 
order_alt <- 
  coeff.long.mat %>% 
  dplyr::filter(coeff!=0) %>% 
  dplyr::group_by(gene) %>% 
  dplyr::count(sort=T) %>% 
  dplyr::select(gene) %>% 
  unlist() %>%
  .[1:8] 

coeff.long.mat.ordered <-  
  coeff.long.mat %>%
  dplyr::filter(gene %in% order_alt, coeff!=0) %>% 
  dplyr::mutate(gene = factor(gene, levels = order_alt), int = factor(int, levels = rev(order_comb)))  %>% 
  mutate(direction = ifelse(coeff>0, "Positive", "Negative")) %>% 
  separate(int, sep=" \\+ ", into=c("Drug", "Stimulus"), remove = FALSE)

Order_Stim <- 
  coeff.long.mat.ordered %>% 
  dplyr::filter(coeff!=0) %>% 
  dplyr::group_by(Stimulus) %>% 
  dplyr::count(sort=T) %>% 
  dplyr::select(Stimulus) %>% 
  unlist()

Order_Drug <- 
  coeff.long.mat.ordered %>% 
  dplyr::filter(coeff!=0) %>% 
  dplyr::group_by(Drug) %>% 
  dplyr::count(sort=T) %>% 
  dplyr::select(Drug) %>% 
  unlist() %>% 
  rev()

coeff.long.mat.ordered <-
  coeff.long.mat.ordered %>%
  mutate(Stimulus=factor(Stimulus, levels = Order_Stim), 
         Drug=factor(Drug, levels = Order_Drug))
  

cyt_labels = c("TGF-b1"="TGF-\u03B21", "sCD40L+IL-4"="sCD40L + IL4", "IL-1b"="IL1\u03B2", "IL-4"="IL4", "IL-6"="IL6","IL-15"="IL15","IL-10"="IL10", "IL-21"="IL21","IL-2"="IL2", "Interferon gamma"= "Interferon \u03B3", "SDF-1a"="SDF-1\u03B1", "CpG ODN"="CpG ODN", "Resiquimod"="Resiquimod", "sCD40L"="sCD40L", "HS-5 CM"="HS-5 CM", "soluble anti-IgM"="soluble anti-IgM", "BAFF"="BAFF")

gene_labels = c("IGHV.status" = "IGHV status", "del11q" = "del(11q)", "del13q" = "del(13q)", "del17p" = "del(17p)", "trisomy12" = "trisomy 12", "TP53" = "TP53", "SF3B1" = "SF3B1", "ATM" = "ATM")

coeff.long.mat.ordered %>% 
  ggplot(aes(y=Drug, x=gene)) +
  geom_tile(aes(fill=coeff),color = "white") +
  scale_fill_gradientn(colors=c(rep(palblues[1:4],each=2),
                                "white", 
                                rep(palreds[5:8], each=2)),  
                       limits=c(-0.8,.8)) +
  t1 +
  theme(axis.text.x = element_text(angle = 35, hjust = 1, vjust = 1),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks = element_blank(),
        axis.ticks.x =element_blank(),
        panel.background = element_rect(color = "black", fill=NA), 
        axis.line.x = element_blank(),
        axis.line.y = element_blank(),
        legend.key.height=unit(0.5, "cm"),
        legend.key.width=unit(2, "cm"),
        legend.title = element_text(face='bold', hjust = 0, size=fontsize+2),
        legend.position = "top",
        legend.key = element_blank(),
        legend.text = element_text(size=fontsize),
        legend.background = element_rect(color = NA),
        strip.text.y.left = element_text(size=18, angle = 0, face="bold"),
        strip.background = element_blank())+
  
  
  labs(fill = expression(beta["int"]))+
  scale_y_discrete(position = "right")+
  scale_x_discrete(labels = gene_labels) +
  facet_grid(Stimulus~.,  scales = "free_y",  space="free_y", switch = "both", labeller =labeller( Stimulus=cyt_labels))
  

```

Applying this broad scale screening approach established some wider trends. Firstly, Trisomy 12 and IGHV status impacted the largest number of drug - stimulus interactions,indicating that their impact on stimulus and drug response individually also extends to drug-stimulus interactions. Secondly, out of all the stimuli, IL4 and sCD40L + IL4-based interactions were modulated by the largest number of features. IL4 alone increased viability uniformly accross genetic backgrounds; its interaction with drugs may be more influenced by genetic features. Notably, the presence of  trisomy 12  acted to increase IL4-induced drug resistance, in almost all drugs for which $\beta{int}$>0.05.  Interestingly, the presence of TP53 and del(17q) reduced the degree to which IL4 induced resistance to the chemotherapeutics Nutlin-3a and Fludarabine, most likely because both these mutations already induce resistance alone. 

### Patient - specific drug - stimulus interactions may have clinical significance
This broad scale screening approach was not perfect in itself. Modelling can only go so far, and it is likely that some cases identified here are artefacts of the modelling approach. Therefore, I decided to investigate each individual case one - by - one to establish which drug - stimulus - gene interactions are the most biologically interesting. This analysis highlighted several drug - stimulus - gene interactions that may have clinical importance and warrant further investigation [@Giles2021].

The first of these concerned the interaction between fludarabine and CpG ODN. The value of $\beta_{int}$ for fludarabine and CpG ODN was modulated by six genetic factors; most strongly by IGHV status, del(11q) and trisomy 12 (Figure \@ref(fig:FluCpGExA), [@Giles2021]).

(ref:FluCpGExA) Genetic features that modulate the size of $\beta{int}$ between fludarabine and CpG ODN. Scatter plot, heatmap and bar plots indicate inputs and outputs of equation \@ref(eq:drugCytGeneInt). Bar plots indicate size and sign of coefficients assigned to genetic features named on right. Scatter plot depicts $\beta{int}$ values for each patient sample i.e.response matrix.  Heatmap tiles indicate mutation status for the selected genetic features (i.e. feature matrix) corresponding to same sample in scatter plot below, to show how size of $\beta{int}$ varies with each feature. See Methods section \@ref(drug-stimulus-gene-interactions-method). _Figure and caption from @Giles2021._


```{r FluCpGExA, fig.cap='(ref:FluCpGExA)', message = FALSE,  echo = FALSE, fig.height=6, fig.width = 10, fig.align="center", out.width = '60%', dev = 'cairo_pdf'}


# Plot predictor profile for Ibrutinib + IL4

grid.arrange(grobs = Lasso_Plots_Fig6["Fludarabine + CpG ODN"], ncol = 1)

```

To look in more detail at what this means, I visualised sample responses to CpG ODN and fludarabine, stratified by IGHV status and trisomy 12 (Figure \@ref(fig:FluCpGExB), [@Giles2021]). In IGHV-M non-trisomy 12 samples, TLR stimulation increased fludarabine efficacy. In samples that were either IGHV-U or trisomy 12+, TLR stimulation induced resistance to fludarabine. TLR stimulation _ex vivo_ can increase or decrease viability, depending on the genetic background and this also impacts on drug response. It would be useful to establish whether these dynamics are replicated within the _in vivo_ niche. 

(ref:FluCpGExB)  Beeswarm-boxplots of log(viability) values for fludarabine and CpG ODN single and combinatorial treatments, faceted by IGHV status and trisomy 12 status. P-values from paired Students t-tests. _Figure and caption from @Giles2021._

```{r FluCpGExB, fig.cap='(ref:FluCpGExB)', message = FALSE,  echo = FALSE, fig.height=6, fig.width = 6, fig.align="center", out.width = '60%', dev = 'cairo_pdf'}


#set facet labels 
tri12.labs <- c("0" = "Non-\ntrisomy 12", "1" = "trisomy 12")
ighv.labs <- c("U" = "IGHV-U", "M" =  "IGHV-M")

#join viability and genetic data tables
  left_join(df_complete, patMeta, by = "PatientID") %>% 
  #filter for drug:stimulus combinations of interest, make sure no NAs in genetic data
  dplyr::filter(Drug_Concentration %in% c("High", "None"),
    DCK%in%c("DMSO:CpG ODN","Fludarabine:CpG ODN","Fludarabine:No Cytokine"),
                !is.na(trisomy12),
                !is.na(IGHV.status)) %>%
    mutate(DCK=factor(DCK, levels=c("DMSO:CpG ODN","Fludarabine:No Cytokine","Fludarabine:CpG ODN"))) %>% 
  
  #plot treatment combination against viability 
  ggplot(aes(x = DCK,y = Log,color= DCK))+
  geom_hline(yintercept = 0)+
  geom_boxplot()+
  geom_beeswarm(cex=1.5) +
  guides(color="none", shape="none")+
  #add p values
  stat_compare_means(method = "t.test",
                     label.y.npc = 0.8, 
                     paired = TRUE, 
                     comparisons = list(c(1,3), c(2,3)),
                     step.increase=0.2, 
                     size=3) +
  xlab("") +
  ylab("Log(Viability)") +
  #facet by trisomy 12 and IGHV status
  facet_grid(vars(trisomy12), 
             vars(IGHV.status),
             labeller = labeller(trisomy12 = tri12.labs, IGHV.status = ighv.labs))+
  
  scale_x_discrete(labels=c("DMSO:CpG ODN"="CpG ODN",
                            "Fludarabine:No Cytokine"="Fludarabine",
                            "Fludarabine:CpG ODN"="Fludarabine \n+ CpG ODN "))+  
    
  scale_color_manual(values=c(colors[4], colors[5],colors[3])) + 
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))+
  t2 +
  theme(strip.background =element_rect(fill=NA),
        strip.text = element_text(size=fontsize+4, face="bold"),
        strip.text.y = element_text(angle = 0),
        axis.text.x = element_text(size=fontsize+4, angle = 35, hjust = 1, vjust = 1))
 


```

TLR-induced drug resistance in certain genetic backgrounds extended to other chemotherapeutic drugs. Nutlin-3a is an MDM2 inhibitor, which in our dataset shows higher efficacy in certain backgrounds including IGHV-U, trisomy12, and del(11q) (Figure \@ref(fig:drugGeneIntMap), Figure \@ref(fig:drugGeneIntEx), Appendix Figure \@ref(fig:NutlinPredictors)). Despite this, TLR stimulation reduces Nutlin-3a toxicity in these same genetic backgrounds, such that treatment with Nutlin-3a and CpG ODN is more toxic to IGHV-M, non-trisomy 12 and non-del(11q) samples i.e. the opposite(Figure \@ref(fig:nutlinCpgEx), Figure \@ref(fig:TLRCpGIGHVdel11q)).  This observation underlines the need to observe drug activity _in vitro_ in the context of microenvironmental signals, in CLL and in other cancers. 

(ref:nutlinCpgEx) Predictor profile depicting genetic features that modulate the size of $\beta{int}$ between Nutlin-3a (MDM2) and CpG ODN.  Scatter plot (bottom) and heatmap (right) and bar plots (left) indicate inputs and outputs of fitting model in Equation \@ref(eq:drugCytGeneInt). Bar plots indicate size and sign of coefficients assigned to genetic features named on the right. Scatter plot depicts ranked patient-specific $\beta{int}$ values i.e. the response matrix.  Heatmap above indicates patient mutation status for the selected genetic features i.e. feature matrix. Heatmap tiles correspond to points in scatter plot (i.e. patient data is aligned), to indicate how the size of $\beta{int}$ varies with selected genetic feature. See Methods section \@ref(drug-stimulus-gene-interactions-method). _Figure and caption from @Giles2021._


```{r nutlinCpgEx, fig.cap='(ref:nutlinCpgEx)', message = FALSE,  echo = FALSE, fig.height=6, fig.width = 10, fig.align="center", out.width = '60%', dev = 'cairo_pdf'}

# Plot predictor profile for Nutlin + CpG ODN

grid.arrange(grobs = Lasso_Plots_Fig6["Nutlin-3a + CpG ODN"], ncol = 1)

```


(ref:TLRCpGIGHVdel11q) Beeswarm-boxplot showing control-normalised log transformed viability values, after treatment with CpG ODN + Nutlin-3a, stratified by del(11q) and IGHV status. p-values from Students t-tests. _Figure from [@Giles2021]_

```{r TLRCpGIGHVdel11q, fig.cap='(ref:TLRCpGIGHVdel11q)', message = FALSE,  echo = FALSE, fig.height = 4, fig.width = 6,  fig.align="center", out.width = '80%', dev = 'cairo_pdf'}

 left_join(df_complete, patMeta, by = "PatientID") %>%
  dplyr::filter(Cytokine=="CpG ODN",
                Drug== "Nutlin-3a",
                Drug_Concentration== "High", 
                !is.na(del11q),
                !is.na(IGHV.status)) %>%
         
  ggplot(aes(x=interaction(del11q, IGHV.status),
                    y=Log,
                    color=(del11q)))+
  geom_hline(yintercept = 0)+
  geom_boxplot()+
  geom_beeswarm(cex=1.5) +
  guides(color="none", shape="none")+
  scale_x_discrete(labels=c("0.M"="IGHV-M\n WT",
                            "0.U"="IGHV-U\n WT",
                            "1.M"="IGHV-M\n del(11q)",
                            "1.U"="IGHV-U\n del(11q)"))+
  stat_compare_means(method = "t.test",
                     label.x.npc = "center", 
                     comparisons = list( c(1,2), c(3,4), c(1,3)),
                     size=3)+
  xlab("") +
  ylab("Log(Viability)") +
  ggtitle("Nutlin-3a + CpG ODN") +
  scale_color_manual(values=c(colors[1], colors[2])) + 

  t1+
  theme(axis.text.x = element_text(angle = 45, vjust =1))


```

The modelling approach also identified interactions, where the differential size of $\beta{int}$ was driven by the single treatment effect, rather than the combination. For example, if a drug is more efficacious in a certain genetic background, but a stimulus nonetheless eradicates drug toxicity, the size of the interaction will be larger in this genetic background. This is because the increase in expected viability in these samples (quantified by $\beta{int}$) would be even higher than in other genetic backgrounds. This is the case  with Ibrutinib + IL4 for instance [@Giles2021].  Ibrutinib showed higher efficacy in trisomy 12 and IGHV-U samples and IL4 induced complete resistance to ibrutinib independently of genetic background. Thus in trisomy 12 and IGHV-U samples treated with ibrutinib, the increase in viability in the context of IL4 stimulation is larger than in non-trisomy 12 and IGHV-M samples, and thus these features are assigned positive coefficients. This highlights the breadth of IL4-induced resistance to ibrutinib in CLL across genetic backgrounds, even where genetic features would indicate that ibrutinib efficacy should be high, and underlines the importance of understanding the activity of this resistance pathway _in vivo_ (Figure \@ref(fig:ibrIl4ExA), Figure \@ref(fig:ibrIl4ExB), [@Giles2021]).

(ref:ibrIl4ExA) Predictor profile depicting genetic features that modulate the size of $\beta{int}$ between ibrutinib (BTK) and IL4. Plot generated as in Figure \@ref(fig:FluCpGExA). See also Methods section \@ref(drug-stimulus-gene-interactions-method). _Figure from @Giles2021._

(ref:ibrIl4ExB) Beeswarm boxplots of log(viability) values for ibrutinib and IL4 single and combinatorial treatments, faceted by IGHV status and trisomy 12 status. P-values from paired Students t-tests. _Figure and caption from @Giles2021._

```{r ibrIl4ExA, fig.cap='(ref:ibrIl4ExA)', message = FALSE,  echo = FALSE, fig.height=6, fig.width = 10, fig.align="center", out.width = '60%', dev = 'cairo_pdf'}

# Plot predictor profile for Ibrutinib + IL4

grid.arrange(grobs = Lasso_Plots_Fig6["Ibrutinib + IL-4"], ncol = 1)
```


```{r ibrIl4ExB, fig.cap='(ref:ibrIl4ExB)', message = FALSE,  echo = FALSE, fig.height=6, fig.width = 6, fig.align="center", out.width = '60%', dev = 'cairo_pdf'}

  left_join(df_complete, patMeta, by = "PatientID")%>%
  #filter for drug:stimulus combinations of interest, make sure no NAs in genetic data
  dplyr::filter(DCK%in%c("DMSO:IL-4","Ibrutinib:IL-4","Ibrutinib:No Cytokine"),
                Drug_Concentration %in% c("High", "None"),
                !is.na(trisomy12),
                !is.na(IGHV.status)) %>%
     mutate(DCK = factor(DCK, levels=c("DMSO:IL-4",
                                      "Ibrutinib:No Cytokine",
                                      "Ibrutinib:IL-4"))) %>%
  
  #plot treatment combination against viability 
  ggplot(aes(x = DCK,y = Log,color= DCK))+
  geom_hline(yintercept = 0)+
  geom_boxplot()+
  geom_beeswarm(cex=1.5) +
  guides(color="none", shape="none")+
  #add p values
  stat_compare_means(method = "t.test",
                     label.y.npc = 0.8, 
                     paired = TRUE, 
                     comparisons = list(c(1,3), c(2,3)),
                     step.increase=0.2, 
                     size=3) +
  xlab("") +
  ylab("Log(Viability)") +
  #facet by trisomy 12 and IGHV status
  facet_grid(vars(trisomy12), 
             vars(IGHV.status),
             labeller = labeller(trisomy12 = tri12.labs, IGHV.status = ighv.labs))+
  
  scale_x_discrete(labels=c("DMSO:IL-4"="IL4",
                            "Ibrutinib:No Cytokine"="Ibrutinib",
                            "Ibrutinib:IL-4"="Ibrutinib \n+ IL4"))+  
    
  scale_color_manual(values=c(colors[4], colors[5], colors[3])) + 
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))+
  t2 +
  theme(strip.background =element_rect(fill=NA),
        strip.text = element_text(size=fontsize+4, face="bold"),
        strip.text.y = element_text(angle = 0),
        axis.text.x = element_text(size=fontsize+4, angle = 35, hjust = 1, vjust = 1))
 


```

this is a resource, all models , and drug - stimulu resposnes stratifed by genetic features can be explored: 

## Summary 
This work represents the first large - scale attempt to integrate the effects of mutations and the microenvironment in drug response in lymphoma, and to the best of my knowledge, in other cancers. The results provide a proof-of-principle of using combinatorial screening, molecular profiling and statistical modelling can generate more complex biological insights to guide further clinical studies. 

More specifically, these results identify a number of key drug-resistance pathways, including the impact of IFN$\gamma$ on ibrutinib. I highlight the importance of the NOS pathway in IL4 and IFN$\gamma$ inhibitor of apoptosis. I also show how some of these resistance pathways are context-dependent. In particular, the ability of TLR stimulation to induce resistance to chemotherapeutics is modulated by many genetic feautres. THis warrants further follow-up. 

Overall these findings illustrate how the presence of known genetic alterations determines how drugs and external stimuli interact with each other.

## Discussion 
IL4and ibrutinib- what's knwon 
implications of ifn gamma and also p38 and ralitmetinib and ifn gamma 

strategies for targeting these - pyridone 6, but not licensned? BET inihibition - paper showing that bet specfically  inhibits ifn and il4 targets - so maybe somehting more braod specturm is useful 

implciatiosnfor studying drug action ex vivo 
In addition, we aimed to demonstrate that combinatorial screening in conjunction with linear modelling is a powerful tool to study the impact of the microenvironment on drug response in cancer, which is often under-appreciated relative to the impact of molecular features. 


In addition, we aimed to demonstrate that combinatorial screening in combination with linear modelling is a powerful approach to study the impact of the microenvironment on drug response in cancer, which is often under-appreciated relative to the impact of molecular features. 


How each category  might be useful 

the importanceof the resource here 

OUr approach distinguishedbetween baseline and interaction s- this is a powerful proof of rpinciple
cancer biology is integratively influenced by cell-intrinsic and cell-extrinsic features <- what attempts hhave their been, persuade whythis needs to continue, how canthis behepful in the clinic 


MUST READ: BET inhibitors and supression of cytokine signalling : BET Bromodomain Inhibition Suppresses Transcriptional Responses to Cytokine-Jak-STAT Signaling in a Gene-specific Manner in Human Monocytes

Nutline and IBET 
nutlin-3ain trisomy 12 - MDM2 is higher 

also IBET -
these also correlate - its possible that they work via a simliar mechanism -ibet inhibits brd4 whcih is inhibtiing p53 targets, so p53 is more active in thesecels 

https://www.nature.com/articles/s41467-020-20378-8

NOS pathway 

TLR - wahts going on in vivo, is it reallyhaving thhis bipolar effect in vivo
Flduarabine response - does this help to explain - more work needed 

In Vitro Sensitivity of CLL Cells to Fludarabine May Be Modulated by the Stimulation of Toll-like Receptors
Eleonora Fonte

IL4-induced resistance- this is clearly important accross a rnage of drufs adn genetic background, more needed in vivo, possible combintatorial therapies


in vivo IL4 evidence: 
from T-cells in chronic lymphocytic leukemia: Guardians or drivers of disease?
Philipp M. Roessner & Martina Seiffert 

Leukemia volume 34, pages 20122024 (2020)
In 2005, Grgn and colleagues investigated T-cell function in CLL. Gene expression profiling (GEP) of sorted T-cells from peripheral blood mononuclear cells (PBMCs) of CLL patients suggested that CD4+ T-cells are skewed toward a Th2 phenotype due to decreased expression of c-Jun NH2-terminal kinase (JNK) and p38 mitogen-activated protein kinase (MAPK) pathway-related genes [12]. This is supported by an increased number of interleukin (IL)-4 secreting CD4+ T-cells in PBMCs of CLL patients in comparison to cells of healthy controls (HC). Studies using the immunocompetent E-TCL1 mouse model of CLL and after adoptive transfer of TCL1 leukemia (TCL1 AT) into BL/6 wild-type (WT) mice revealed more IL-4 secretion of CD4+ T-cells in comparison to control mice [13, 14]. Similarly, conditioned medium of CLL cells was able to induce secretion of IL-4 in allogeneic mixed lymphocyte reactions (MLR) of healthy individuals, suggesting that soluble factors secreted by CLL cells trigger Th2 T-cell polarization in vitro [15].

in other cancers we shoudl integrate drugs - stimul and genetics 

flurdarabine and Cpg, different outcomes in IGHV M and U patients 

the imrpotatnce of IL4 in all patietn backgrounds 


TLR Signaling Is Activated in Lymph NodeResident CLL Cells and Is Only Partially Inhibited by Ibrutinib
Eman L. Dadashian
https://cancerres.aacrjournals.org/content/79/2/360.short

There are models to investigate ibrutinib resistance in vivo, but they are lacking in for ME: 
 Development and characterization of prototypes for in vitro and in vivo mouse models of ibrutinib-resistant CLL
Burcu Aslan


also check out discussion from :https://www.sciencedirect.com/science/article/pii/S0006497119402206


## Contributions Statement
The author of this thesis performed the statistical inference and regression modelling outlined in this chapter. In essence, she conceptualised and performed the linear modelling and multivariate modelling in Equations .... In addition, she  performed univariate analysis of genetic determinants of drug response and data visualisation and investigation of hits. 

Peter-Martin Bruch generated the viability screening data used in the model. The genetic annotation data was taken from [@JCIpaper]. The categorisation of drug - stimulus interactions, and investigation of examples of drug - stimulus interactions was performed in collaboration with Peter-Martin Bruch for the paper Bruch & Giles et al 2021. The author also received support from Frederich Zwiebell and Junyan Lu, who provided advice with regards to modelling approaches. 


-check pdf and tidy 
referencing
linear model 
figures tidy 
extra text 

-check of text 
= 6 done! 

