# Results {#chapter7}
### IL4 induces resistance to BCR inhibitors and chemotherapeutics {#IL4resistance}
(Figure \@ref(fig:IL4Res)). 

(ref:IL4Res) Examples of IL4-induced resistance. Plots show log transformed viability values with each treatment, for all samples. Matching samples are linked across treatments. Black and blue horizontal lines indicate the predicted viability from the linear model for each single and combinatorial treatment. In combinatorial treatment, both the expected viability based on the additive effect of drug and stimulus (blue), and the viability with interaction (black) are shown, to indicate the impact of the interaction. Plots show IL4 + (A) Idelalisib (Pi3K), (B) PRT062607 (SYK) (C) Everolimus (mTOR), and (D) Selumetinib  (p38 MAPK). See Methods section \@ref(drug-stimulus-linear-model-method). _Figure from Bruch & Giles et al. 2021._

```{r IL4Res, fig.cap='(ref:IL4Res)', message = FALSE,  echo = FALSE, fig.height=6.6, fig.width = 12, fig.align="center", out.width = '80%', dev = 'cairo_pdf'}


(wrap_plots(plotList$`treatment_drug:Idelalisib:treatment_cytokine:IL-4`) + wrap_plots(plotList$`treatment_drug:PRT062607:treatment_cytokine:IL-4`)) /
  
(wrap_plots(plotList$`treatment_drug:Everolimus:treatment_cytokine:IL-4`) +
  
wrap_plots(plotList$`treatment_drug:Selumetinib:treatment_cytokine:IL-4`)) + plot_annotation(tag_levels = c("A"))

```

* Organize material and present results.

* Use tables, figures (but prefer visual presentation):

  + Tables and figures should supplement (and not duplicate) the text.

  + Tables and figures should be provided with legends.

  + _Figure \@ref(fig:graph) shows how to include and reference graphics.
    The graphic must be labelled before. Files must be in __.eps__ format. You
    can do this really easily in R Markdown with `knitr::include_graphics()`_!

  + Figures can be referenced with `\@ref(fig:<name>)`, where `<name>` is the
    name of the code chunk.

```{r graph4, echo=FALSE, fig.cap="Estimated residuals from model XXX4. ...", out.width = "50%"}
knitr::include_graphics("figures/graph.eps")
```
  + Tables and graphics may appear in the text or in the appendix, especially if
  there are many simulation results tabulated, but is also depends on the study
  and number of tables resp. figures. The key graphs and tables must appear in
  the text!

* R Markdown can also supports math equations just like _LaTeX_!

  + _Equation \@ref(eq:SpecDens) represents the ACs of a stationary
    stochastic process:_
    \begin{equation}
            f_y(\lambda) = (2\pi)^{-1} \sum_{j=-\infty}^{\infty}
                           \gamma_j e^{-i\lambda j}
                         =(2\pi)^{-1}\left(\gamma_0 + 2 \sum_{j=1}^{\infty}
        \gamma_j \cos(\lambda j)\right)
                                       (\#eq:SpecDens)
    \end{equation}
    _where $i=\sqrt{-1}$ is the imaginary unit, $\lambda \in [-\pi, \pi]$ is the
    frequency and the $\gamma_j$ are the autocovariances of $y_t$._

  + Equations can be referenced with `\@ref(eq:<name>)`, where name is defined
    by adding `(\#eq:<name>)` in the line immediately before `\end{equation}`.

## Review of Results

*  Do the results support or do they contradict economic theory ?

*  What does the reader learn from the results?

*  Try to give an intuition for your results.

*  Provide robustness checks.

*  Compare to previous research.
