# Results {#chapter7}
### IL4 induces resistance to BCR inhibitors and chemotherapeutics {#IL4resistance}

```{r}
library(patchwork)

```

(Figure \@ref(fig:IL4Res)). 

(ref:IL4Res) Examples of IL4-induced resistance. Plots show log transformed viability values with each treatment, for all samples. Matching samples are linked across treatments. Black and blue horizontal lines indicate the predicted viability from the linear model for each single and combinatorial treatment. In combinatorial treatment, both the expected viability based on the additive effect of drug and stimulus (blue), and the viability with interaction (black) are shown, to indicate the impact of the interaction. Plots show IL4 + (A) Idelalisib (Pi3K), (B) PRT062607 (SYK) (C) Everolimus (mTOR), and (D) Selumetinib  (p38 MAPK). See Methods section \@ref(drug-stimulus-linear-model-method). _Figure from Bruch & Giles et al. 2021._

```{r IL4Res, fig.cap='(ref:IL4Res)', message = FALSE,  echo = FALSE, fig.height=6.6, fig.width = 12, fig.align="center", out.width = '80%', dev = 'cairo_pdf'}


(wrap_plots(plotList$`treatment_drug:Idelalisib:treatment_cytokine:IL-4`) + wrap_plots(plotList$`treatment_drug:PRT062607:treatment_cytokine:IL-4`)) /
  
(wrap_plots(plotList$`treatment_drug:Everolimus:treatment_cytokine:IL-4`) +
  
wrap_plots(plotList$`treatment_drug:Selumetinib:treatment_cytokine:IL-4`)) + plot_annotation(tag_levels = c("A"))

```

### IL4 levels in vivo correlate with outcome {#il4-ihc}
(Figure \@ref(fig:IL4ihc)). 

(ref:IL4ihc) See Methods section  \@ref(il4-ihc-exp) and \@ref(il4-ihc-method).  _Figure from Bruch & Giles et al. 2021._

```{r IL4ihc, fig.cap='(ref:IL4ihc)', message = FALSE,  echo = FALSE, fig.height=6.6, fig.width = 12, fig.align="center", out.width = '80%', dev = 'cairo_pdf'}



```

In addition to the screen, we acquired the following validatory data (Figure \@ref(fig:ibribetil4)). 

with surv cures, make sure update the label to TTT (beaucs its the same as TFS)

(ref:ibribetil4) Overview of the approach for the validation experiments. Four CLL PBMC samples were incubated with ibrutinib, IL4 and IBET-762, both individually and in all combinations. Samples were taken for ATAC sequencing and RNA sequencing after 6 hours and for mass spectrometry after 24 hours. Sequencing data was processed for downstream analysis, outline in Chapter \@ref(chapter7). 


```{r ibribetil4, out.width = "100%", fig.cap='(ref:ibribetil4)', echo = FALSE}

knitr::include_graphics("figures/ibribetil4.eps")

```

