# Appendix {-}
```{r loadDataAppendix}
#drug-cyt-gene interactions
load("data/coeff.long.mat_Fig6.RData") # from data for figure 6 in supplement
```

## Figures {-}



(ref:diffTFsmallvolPlot) Analysis of ATACseq dataset of two trisomy 12 and two non-trisomy 12 untreated CLL PMBC samples. The volcano plot depicts change in TF activity (x axis) versus BH-adjusted p-values (y axis) for  636  TFs, comparing trisomy 12  and non-trisomy 12 samples. The [@diffTF][REFERENCE] package was ran in analytical mode to calculate TF activity, measured as  weighted mean difference. TFs are labeled if adjusted p-value < 0.01 and absolute weighted mean difference > 0.15.


```{r diffTFsmallvolPlot, fig.cap='(ref:diffTFsmallvolPlot)', message = FALSE,  echo = FALSE, fig.height=5, fig.width = 5, fig.align="center", out.width = '60%', dev = 'cairo_pdf'}
knitr::include_graphics("figures/graph.eps")
```

(ref:drugcytGeneIntAll) 


```{r drugcytGeneIntAll, fig.cap='(ref:drugcytGeneIntAll)', message = FALSE,  echo = FALSE, fig.height=5, fig.width = 5, fig.align="center", out.width = '60%', dev = 'cairo_pdf'}
colnames(Cytokine_labels) <- c("Stimulus", "label")
coeff.long.mat_Fig6 %>%
  tidyr::separate(int, sep=" \\+ ", into=c("Drug", "Stimulus"), remove =FALSE)%>% 
  left_join(Cytokine_labels, by = "Stimulus") %>%
  dplyr::filter(coeff!=0) %>%
  ggplot(aes(y=Drug, x=gene))+
  geom_tile(aes(fill=coeff),color = "white")+
  scale_fill_gradientn(colors=c(rep(palblues[1:4],each=2),"white", rep(palreds[5:8], each=2)),  limits=c(-0.8,.8))+
  t1+
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks = element_blank(),
        axis.ticks.x = element_blank(),
        panel.background = element_rect(color = "black", fill=NA),
        axis.line.x = element_blank(),
        axis.line.y = element_blank(),
        legend.key.height = unit(0.5, "cm"),
        legend.key.width = unit(2, "cm"),
        legend.title = element_text(face='bold', hjust = 0, size=fontsize+2),
        legend.position = "bottom",
        legend.key = element_blank(),
        legend.text = element_text(size=fontsize),
        legend.background = element_rect(color = NA),
        strip.text.y.left = element_text(angle = 0, face="bold", size = 22),
        strip.background = element_blank())+
  labs(fill = expression(paste(beta, "-int")))+
  scale_y_discrete(position = "right",limits=rev)+
  scale_x_discrete(labels=c("KRAS,\nNRAS,\nBRAF"="RAS/RAF"))+
   facet_grid(label~., scales = "free_y",  space="free_y", switch = "both")
```

## Tables {-}

(ref:receptorPairs) Table depicting stimulus and receptor pairs used in stimulus response - stimulus receptor expression analysis. 

```{r receptorPairs, tab.cap='(ref:receptorPairs)', echo = FALSE}

cyt_and_receptors %>%  
    kable() %>%
    kable_styling(latex_options = "striped", 
                stripe_color = colors[7],
                html_font = "Arial Narrow") 

```

\begin{table}[ht]
    \begin{center}
        {\footnotesize
        \begin{tabular}{l|cccccccccc}
        \hline \hline
                        & 3m    & 6m    & 1yr   & 2yr   & 3yr   & 5yr   & 7yr   & 10yr  & 12yr  & 15yr   \\
            \hline
                Mean   & 3.138 & 3.191 & 3.307 & 3.544 & 3.756 & 4.093 & 4.354 & 4.621 & 4.741 & 4.878  \\
                Median & 3.013 & 3.109 & 3.228 & 3.490 & 3.680 & 3.906 & 4.117 & 4.420 & 4.575 & 4.759  \\
                Min    & 1.984 & 1.950 & 1.956 & 2.010 & 2.240 & 2.615 & 2.850 & 3.120 & 3.250 & 3.395  \\
                Max    & 5.211 & 5.274 & 5.415 & 5.583 & 5.698 & 5.805 & 5.900 & 6.031 & 6.150 & 6.295  \\
                StD    & 0.915 & 0.919 & 0.935 & 0.910 & 0.876 & 0.825 & 0.803 & 0.776 & 0.768 & 0.762  \\
            \hline \hline
        \end{tabular}}
    \end{center}
    \caption{Detailed descriptive statistics of location and dispersion for
    2100 observed swap rates for the period from
    February 15, 1999 to March 2, 2007. Swap rates measured as 3.12 (instead of 0.0312).}
    \label{tab:apptable}
\end{table}

\newpage
